Tracking maternal proteins uncovers a central role for the residual body in organelle recycling during Toxoplasma gondii replication

Version 1

Abstract

Toxoplasma gondii replicates through endodyogeny, an unconventional form of internal budding in which two daughter cells are assembled within a single mother cell. During this process, daughter cells must acquire a full complement of organelles, which may be inherited from the mother, formed de novo, or assembled through a combination of both mechanisms. To date the fate of maternal components during replication remains poorly understood. We previously showed that F-actin–driven dynamics generate the intravacuolar network, which defines the residual body (RB) and facilitates recycling of microneme proteins. However, the inheritance and recycling of other organelles have not been systematically analysed.

To address this, we employed a dual HaloTag-based pulse-chase fluorescence labelling strategy to distinguish between de novo–synthesized and recycled proteins in replicating tachyzoites. This approach reveals three distinct organelle inheritance patterns: (1) direct transmission of intact maternal organelles (e.g., rhoptries, micronemes), (2) expansion and division of pre-existing maternal organelles with incorporation of newly synthesized components (e.g., Golgi apparatus, apicoplast), and (3) degradation of maternal structures without recycling (e.g., inner membrane complex). Furthermore, we identify Myosin F (MyoF) as the key motor protein that mediates the selective recycling of maternal organelles via the RB. These findings redefine the RB as an active trafficking hub and reveal a selective, regulated system of organelle inheritance and recycling that is critical for intracellular organization and parasite development.

Introduction

Toxoplasma gondii tachyzoites replicate through a specialized form of internal budding known as endodyogeny, wherein two daughter cells (DCs) are assembled within the cytoplasm of the mother cell (Francia and Striepen, 2014; Hu et al., 2002). This unique replication mechanism demands precise orchestration of organelle inheritance and spatial organization within the parasitophorous vacuole (PV). Apicomplexan parasites, including T. gondii, exhibit a highly polarized cellular architecture defined by specialized secretory organelles—micronemes, rhoptries, and dense granules—essential for host cell invasion and intracellular survival. These are accompanied by canonical organelles such as the Golgi apparatus, endoplasmic reticulum (ER), a single mitochondrion, and the apicoplast, a non-photosynthetic plastid. Notably, unlike most eukaryotes, T. gondii is enveloped by a pellicle consisting of the plasma membrane and an underlying inner membrane complex (IMC) (Ouologuem and Roos, 2014), a structure composed of flattened vesicles tightly associated with the subpellicular cytoskeleton (Harding and Meissner, 2014).

As DCs emerge during endodyogeny, they progressively encapsulate nearly all maternal cytoplasmic contents. While the sequence of organelle acquisition has been extensively characterized (Nishi et al., 2008), the extent to which maternal organelles and proteins are recycled remains unclear. We previously showed that maternal micronemal proteins, such as MIC2, are almost entirely recycled into daughter cells via an F-actin–dependent mechanism and are trafficked through the intravacuolar network (IVN), which structurally organizes the residual body (RB) (Periz et al., 2019). Historically viewed as a passive remnant of maternal cytoplasm, the RB has emerged as a key structure in coordinating replication, facilitating organelle recycling, and maintaining cytoplasmic continuity across individual parasites (Periz et al., 2017b). The RB forms at the posterior end of dividing parasites, physically linking them through the actin-rich IVN and enabling cytoplasmic exchange and synchronized development within the PV (Periz et al., 2017b; Tosetti et al., 2019).

Recent studies have shed some light on the molecular architecture and regulation of the IVN. We demonstrated that F-actin filaments organize the IVN, with disruption of actin (ACT1) resulting in parasite disorganization and impaired microneme recycling (Periz et al., 2019; Periz et al., 2017b). The RB and IVN also require the function of unconventional myosins MyoI and MyoJ, which are implicated in cell-cell communication (Frenal et al., 2017). Additionally, the actin-nucleating factors Formin-2 and Formin-3 are required for IVN formation (Stortz et al., 2019; Tosetti et al., 2019). However, the mechanisms driving the physical transport of maternal organelles remain poorly defined. F-actin could mediate this process either by serving as a track for myosin-dependent vesicular transport, as described for myosin VI in other eukaryotes (Frank et al., 2004), or through the intrinsic mobility and dynamic association of actin bundles themselves, which could transiently interact to promote vesicle exchange (Das et al., 2021; Khaitlina, 2014; Moore et al., 2021).

Although the biogenesis of daughter cell structures has been well described, the fate of maternal organelles during replication remains incompletely understood. Evidence suggests that maternal components such as MIC2 and the IMC protein GAP40 can be recycled via actin-dependent mechanisms (Ouologuem and Roos, 2014; Periz et al., 2019). However, a systematic investigation of organelle recycling during endodyogeny has yet to be performed.

In this study, we employ a dual-labeling strategy using the HaloTag system (Urh and Rosenberg, 2012) to investigate the fate of maternal organelles during T. gondii replication. Through fluorescence intensity-based analysis, we identify three distinct fates for maternal proteins: full inheritance of intact organelles, incorporation of newly synthesized components into expanded maternal structures, and degradation without recycling. Furthermore, we demonstrate that the residual body functions as a key recycling hub, with this process dependent on the activity of the class XXII myosin F (MyoF). These findings provide new mechanistic insights into organelle inheritance and reveal how intracellular organization is maintained during parasite replication.

Results

HaloTag-Based pulse chase labelling allows discrimination of recycled and de novo material

To evaluate HaloTag as a tool for distinguishing maternal from newly synthesized proteins during endodyogeny, two marker proteins with known fates were used: MIC2, which undergoes extensive recycling (Periz et al., 2019), and IMC1, primarily synthesized de novo (Ouologuem and Roos, 2014). Using CRISPR/Cas9, endogenous loci were tagged to generate MIC2-Halo and IMC1-Halo strains (Singer et al., 2023).

Both strains were sequentially labeled: maternal proteins were marked with Halo JF646 (M-MIC2, M-IMC1), followed 24 hours later by Halo JF549 to label newly synthesized proteins (n-MIC2, n-IMC1) (Figure 1A). Fluorescence intensity (FI) analysis across replication stages (Figure 1B–D) revealed a progressive decline of M-MIC2 and an increase in n-MIC2, confirming protein recycling. In contrast, M-IMC1 FI dropped from 100% to 14% between stages 1 and 2, with a corresponding rise in n-IMC1 to ∼80%, indicating predominant de novo synthesis. These distinct FI dynamics demonstrate HaloTag’s utility for resolving protein origin and support the de novo assembly of the IMC.

Discrimination between maternal and de novo material.

A) Dual staining scheme. Maternal material is labelled initially with Janila Fluor®-646, and the excess washed out. After replication of parasites de novo synthesised material is labelled with Janila Fluor®-549, allowing efficient discrimination of the two populations. B) Fluorescence intensity quantification of maternal (M) and de novo sythenetised (n) MIC2 and IMC1 during replication from stage 1 to 8. Magenta: M-MIC2, Rosa: M-IMC1, Green: n-MIC2, Dark Green: n-IMC1. C) Representative picture of MIC2-Halo parasites during replication (stage 1 to 8) stained for both maternal (M) and de-novo (n) MIC2. Magenta: M-MIC2, Green: n-MIC2. Maternal MIC2 is efficiently recycled into the daughters, while de novo MIC2 is formed after each replication cycle D) Representative picture of IMC1-Halo parasites during replication (stage 1 to 8) stained for both maternal (M) and de-novo (n) IMC1. Magenta: M-IMC1, Green: n-IMC1. In contrast to micronemes, maternal IMC is degraded, and only residual amount is detected after the first replication cycle. Three biological replicates were used for all analyses; error bars are standard deviations, and the centre measurement of the graph bars is the mean. All scale bars = 1 µm.

HaloTag profiling reveals three distinct patterns of protein inheritance

To further investigate protein inheritance during endodyogeny, endogenous HaloTag fusions were generated for a set of organelle-associated proteins: RON2 and ROP1 (rhoptries) (Besteiro et al., 2009; Striepen et al., 2001), GRA1 (dense granules) (Carruthers and Sibley, 1997), TIC20 (apicoplast) (Carruthers and Sibley, 1997), SortLR (Golgi) (Sloves et al., 2012), ANKER1 (ER) (Barylyuk et al., 2020), GAPM1a (IMC) (Harding et al., 2019), and MyoA (glideosome) (Meissner et al., 2002) (Figure 2A, Figure S1).

Three recycling fates for the maternal organelles.

A) Overview of T. gondii organelles and molecular markers used for visualisation with endogenous Halo-Tags. B) Fluorescence intensity quantification of maternal (M) molecular marker listed in A. The fluorescence intensity variation during replication, allows to group them into three groups: 1) Efficient, almost quantitative recycling. 2) even distribution of maternal material and 3) almost exclusive de novo synthesis. Dark blue: RON2, Blue: ROP1, Pale blue: MIC2, Dark green: SortLR, Green: MyoA, Pale green: Tic20, Greenish white: ANKER1, Gold: GAPM1a, Yellow: IMC1. C) Representative picture of parasites expressing indicated Halo tagged proteins (stage 1 to 8) stained for maternal (M) proteins. Three biological replicates were used for all analyses; the graph indicate the mean value of the triplicate. All scale bars = 1 µm.

Following the same sequential labeling strategy used for MIC2 and IMC1, fluorescence intensity (FI) of maternal proteins was quantified over successive replication cycles (Figure 2B, C). Analysis revealed three distinct inheritance profiles:

Group 1: RON2 and ROP1 exhibited stable maternal FI, with minimal decline of FI within a single rhoptry, indicating recycling of whole organelles. In good agreement with this hypothesis, detailed analysis of M-RON2 inheritance demonstrated that the maternal rhoptries are separated during the distribution and that some daughter cells did not obtain maternal organelles after successive round of replication (Figure 2, Figure S2).

Group 2: Proteins from diverse organelles (SortLR, ANKER1, TIC20, MyoA) showed a stepwise ∼50% reduction in maternal FI per division, suggesting that recycled, maternal material and de novo material end up in the same organelle

Group 3: IMC proteins (IMC1 and GAPM1a) displayed a sharp FI loss between the first and second replication cycles, consistent with predominant de novo synthesis of the IMC and degradation of maternal material.

Interestingly, MIC2 exhibited intermediate behavior, prompting further analysis to refine its classification. These patterns highlight distinct organelle-specific protein inheritance mechanisms in T. gondii.

Micronemes and rhoptries are recycled as intact organelles during Toxoplasma endodyogeny

Micronemes are known to comprise distinct subpopulations (Kremer et al., 2013), and prior work has shown that recycled and de novo MIC2 localize to separate microneme subsets (Periz et al., 2019). Based on these observations, we hypothesized that both micronemes and rhoptries are recycled as intact organelles, with new organelles formed independently via de novo synthesis.

To test this, we conducted detailed analyses of MIC2 and rhoptry inheritance. As expected, recycled (M-MIC2) and newly synthesized MIC2 (n-MIC2) were largely segregated into distinct micronemes with minimal colocalization (Figure S3)(Periz et al., 2019). Focusing on sparsely distributed micronemes outside the apical region (Figure 3A, boxed area), we measured fluorescence intensity (FI) of individual M-MIC2-positive micronemes at replication stages 2, 4, and 8 (Figure 3A, zoomed region; 3C). The relatively stable FI of these isolated structures supports whole organelle recycling, consistent with the FI profiles of rhoptries (Figure 2B, C).

Micronemes and rhoptries are recycled as whole organelles

A) Representative picture of MIC2-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) MIC2. Magenta: M-MIC2, Green: n-MIC2. Zoom windows allow to visualise isolated apical M-MIC2 signal for which the fluorescence intensity is quantified in C. B) Quantification of the inheritance of M-MIC2 by the daughter cells during replication (stage 1 to 8). All daughter cells inherited M-MIC2. C) Fluorescence intensity quantification of isolated M-MIC2 signal as illustrated in A. D) Representative picture of RON2-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) MIC2. Magenta: M-RON2, Green: n-RON2. Zoom windows allow to visualise isolated apical M-RON2 signal for which the fluorescence intensity is quantified in F. E) Quantification of the inheritance of M-RON2 by the daughter cells during replication (stage 1 to 8). Not all daughter cells inherit M-RON2 (asterisks), best seen in later replication stages. F) Fluorescence intensity quantification of isolated M-RON2 signal as illustrated in D. Three biological replicates were used for all analyses; error bars are standard deviations, and the centre measurement of the graph bars is the mean. All scale bars = 1 µm.

Rhoptry analysis further corroborated this model. As with MIC2, M-RON2 and n-RON2 localized to distinct rhoptry populations (Figure S2), with divergence becoming more apparent at stage 8 (Figure 3D, Figure S2D). M-RON2 FI remained relatively constant through replication (Figure 3 F), while n-RON2 gradually increased from stage 2 onward (Figure S2C). Due to their low copy number (8–12 per tachyzoite), maternal rhoptries were separated and often absent in some daughter cells after successive divisions, consistent with stochastic whole-organelle inheritance illustrated by decrease of the M-RON2 fluorescence signal average area and the constancy of the total surface area (Figure S2A, B). As consequence a progressive decline in M-RON2-positive parasites per vacuole can be observed (Figure 3E, D; Figure S2D). In contrast, micronemes showed a more controlled segregation: even in large parasitophorous vacuoles, all daughter parasites retained a mix of recycled and newly synthesized micronemes (Figure S4), indicating regulated partitioning.

Together, these results demonstrate that both micronemes and rhoptries are recycled as intact organelles. De novo organellogenesis acts in parallel to maintain a full complement in each daughter cell, ensuring faithful inheritance during endodyogeny.

Golgi inheritance involves coordinated expansion and partitioning of maternal and de novo proteins

Next, we analyzed SortLR, a Golgi marker, using dual HaloTag labeling. Maternal SortLR (M-SortLR) fluorescence declined by ∼50% per replication round, while newly synthesized SortLR (n-SortLR) increased proportionally, without abrupt loss (Figure 4A–B). This gradual dilution contrasts sharply with the rapid decay seen in Group 3 proteins.

Expansion of the mother organelle for equivalent sharing to the daughter.

A) Representative picture of SortLR-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) SortLR. Magenta: M-SortLR, Green: n-SortLR. B) Fluorescence intensity quantification of M-SortLR and n-SortLR illustrated in A. Magenta: M-SortLR, Green: n-SortLR. C) Quantification of the inheritance of M-SortLR by the daughter cells during replication (stage 1 to 8). All daughter cells inherited M-SortLR. D) Quantification of the signal area of M-SortLR during replication. The total surface of M-SortLR increase at each replication by about a factor 2. Three biological replicates were used for all analyses; error bars are standard deviations, and the centre measurement of the graph bars is the mean. All scale bars = 1 µm.

M-SortLR and n-SortLR exhibited strong colocalization across replication stages, and all daughter cells retained M-SortLR (Figure 4C). The total Golgi area expanded from ∼2 µm² to ∼10 µm² by stage 8 (Figure 4D), consistent with coordinated Golgi elongation and medial fission, as described previously (Pelletier et al., 2002). These results demonstrate that the Golgi undergoes autonomous duplication, incorporating both maternal and de novo proteins during its growth and partitioning.

Thus, Group 2 organelles follow a distinct inheritance mode, involving regulated expansion, integration of new material, and equal distribution to progeny—distinct from the whole-organelle recycling of rhoptries/micronemes and the complete turnover seen in IMC proteins.

Maternal IMC proteins are degraded in the RB during replication

Group 3 proteins, typified by IMC components like GAPM1a and IMC1, exhibit a sharply distinct inheritance profile, characterized by rapid loss of maternal fluorescence during replication. Using HaloTag-based dual labeling, we observed a ∼90% reduction in maternal GAPM1a signal by stage 2, with minimal retention through later stages (Figure 5A–C), indicative of active degradation rather than recycling. M-GAPM1a and n-GAPM1a did not colocalize, and residual maternal signal localized transiently at the posterior pole before disappearing entirely, suggesting disposal via the residual body (RB) (Figure 5D).

Degradation of the inner membrane complex.

A) Representative picture of GAPM1a-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) GAPM1a. B) Fluorescence intensity quantification of M-GAPM1a and n-GAPM1a illustrated in A. C) Quantification of the inheritance of M-GAPM1a by the daughter cells during replication (stage 1 to 8). After replication, M-GAPM1a is not visible in daughter cells. D) GAPM1a is degraded in the RB. Representative images of GAPM1a-Halo parasites at different stages of daughter cell development. M-GAPM1a collapses toward the forming residual body, where the signal disappears after completion of replication, indicating it’s degradation. E) Time series of IMC1-Halo parasites during the first replication. Parasite were stained for the maternal (M) IMC1 prior invasion. Three regions of the parasites were analysed: the apical (green arrow), the cytoplasmic (blue arrow) and the basal (magenta arrow). F) Fluorescence intensity analysis of the three regions defined in E. The curve and analysed region arrow share the same colour code. Green: Apical, Blue: Cytoplasm, Magenta: Basal. After the accumulation of the mother IMC at the basal pole, the FI of M-IMC1 decrease without redistributing to any other region. Three biological replicates were used for all analyses; error bars are standard deviations, and the centre measurement of the graph bars is the mean. All scale bars = 1 µm.

Live-cell imaging of IMC1 further supported this model: maternal IMC1 (M-IMC1) condensed at the posterior pole during daughter budding, followed by a marked fluorescence drop, without redistribution to other regions (Figure 5E, F and Figure S5, Video S1). This signal loss was replication-specific, as non-dividing parasites retained stable M-IMC1 levels (Figure S5D). Quantification across regions revealed a transient increase at the posterior, followed by complete disappearance, consistent with degradation (Figure 5F, Figure S5C).

Interestingly, daughter IMCs (DCs) exhibited brighter fluorescence than the maternal IMC (Figure S5E, F), especially during early stages of elongation (Figure S5 G, H). FI peaked when DCs reached ∼3 μm in length, suggesting the IMC is fully assembled before emergence and subsequently unfolds without requiring additional material. This preformation explains the absence of maternal IMC recycling. These findings align with earlier observations that IMC components are synthesized de novo during elongation (Ouologuem and Roos, 2014).

Distinct recycling pathways govern whole-organelle inheritance of rhoptries and micronemes

Our data suggest that the residual body (RB) plays a selective role in organelle recycling: while microneme and rhoptry proteins are routed through the RB, IMC proteins are primarily degraded within this compartment. In contrast, recycling of maternal Golgi, ER, and apicoplast proteins appears to occur directly, bypassing the RB.

To investigate this further, we performed live-cell imaging of rhoptry and microneme inheritance (Figure 6A, B, Video S2, S3). MIC2-Halo parasites co-expressing IMC1-YFP, showed that maternal MIC2 (M-MIC2) transits through the RB (Figure 6A, white arrows). Notably, after endodyogeny was completed, M-MIC2 was redistributed from the RB to the apical tip of the daughter cells (Figure 6A, 11:30, 16:00 h), confirming that the RB serves as a temporary reservoir during microneme recycling (Periz et al., 2019).

Micronemes and rhoptries are recycled similarly, but differ in timing and location.

A) Time-lapse of MIC2-Halo over two replication cycles. Maternal MIC2 (M-MIC2) is stained, and daughter cell formation is visualized via IMC1-YFP. M-MIC2 is transported via the residual body (RB) (white arrows; see Video S2). B) Time-lapse of RON2-Halo parasites with IMC-YFP. Maternal RON2 (M-RON2) is transported before mother cell collapse and RB formation (white arrows). C) Representative images showing maternal organelles associating with F-actin and the RB. Insets highlight colocalization of maternal proteins and F-actin. D) Quantification of M-MIC2 and M-RON2 colocalization with F-actin. Data from three biological replicates; bars represent means ± SD. Scale bars = 1 µm. Myosin-F drives F-actin mediated recycling of maternal organelles via the residual body.

For rhoptries, we examined the inheritance of maternal RON2 (M-RON2) in RON2-Halo parasites co-expressing IMC1-YFP. In approximately 90% of parasites undergoing replication, M-RON2 was integrated into daughter rhoptries prior to mother cell collapse and formation of the RB (Figure 6B, 4:15–4:30 h and 10:30–10:45 h). Like M-MIC2, M-RON2 was occasionally detected in the RB, though less prominently, suggesting more rapid or tightly regulated recycling.

To test whether rhoptry recycling also depends on F-actin, as previously reported for micronemes (Periz et al., 2019), we expressed the F-actin marker Cb-Emerald (Periz et al., 2017b) in the RON2-Halo line (Figure 6C, D). While m-MIC2 displayed clear F-actin– associated movement along filaments in the residual body (Figure 6C), colocalization of M-RON2 with Cb-Emerald in the residual body was infrequent. On the other hand, both secretory organelles were found to be associated with cytoplasmic F-actin (Figure 6D).

In summary, both rhoptries and micronemes are inherited as intact organelles during T. gondii replication. However, the underlying timing differs: microneme recycling is delayed until after daughter emergence, whereas rhoptry inheritance is more direct and occurs prior to mother cell collapse. These findings highlight the organelle-specific complexity of recycling pathways and support whole-organelle recycling of micronemes and rhoptries.

Given the F-actin dependency of this recycling pathway (Periz et al., 2019), we hypothesized that a conserved apicomplexan myosin mediates this process. MyoF, a class XXII myosin is conserved across Apicomplexa (Jacot et al., 2013) and has been implicated in diverse functions—from apicoplast inheritance to F-actin–dependent vesicular transport (Carmeille et al., 2021; Heaslip et al., 2016; Kellermeier and Heaslip, 2024).

We endogenously Halo-tagged MIC2, RON2, and SortLR in the auxin-inducible degron line MyoF-mAID (Carmeille et al., 2021) and performed dual HaloTag labelling to differentiate maternal (M) from de novo (n) protein pools.

Live imaging revealed that MyoF depletion resulted in the accumulation of maternal but not de novo MIC2 and RON2 in the RB during early replication stages (Figure 7A, B). At later stages, recycled de novo organelles from earlier cycles (now maternal) also began to accumulate. Immunofluorescence with α-AMA1, α-MIC4, and α-MIC8 confirmed that this phenotype broadly affects the entire repertoire of micronemal proteins (Figure S6, S7). Quantification showed that approximately 90% of vacuoles exhibited RB accumulation of maternal material upon auxin treatment (Figure 7D), indicating that MyoF specifically mediates the recycling of maternal micronemes and rhoptries. Interestingly, although our time lapse analysis indicated that rhoptries only rarely traffic via the RB, upon depletion of MyoF, they are accumulating in the RB, supporting the hypothesis that micronemes and rhoptries are transported in the same manner during recycling.

MyoF is essential for recycling apical secretory organelles.

A) Effect of MyoF knockdown (KD) on maternal microneme inheritance. MyoF-mAID MIC2-Halo parasites were labeled and grown ± auxin for 24h. Magenta: maternal MIC2 (M-MIC2), Green: newly synthesized MIC2 (n-MIC2). Without MyoF, M-MIC2 accumulates in the residual body (RB) instead of being passed to daughter cells (Stage 1–2), with increasing accumulation over replication cycles (Stage 2–8). B) Effect of MyoF KD on maternal rhoptry inheritance. MyoF-mAID RON2-Halo parasites show M-RON2 retention in the RB, mirroring the pattern seen with MIC2. C) Effect of MyoF KD on maternal Golgi inheritance. MyoF-mAID SortLR-Halo parasites show no significant RB accumulation of M-SortLR, unlike MIC2 and RON2. D) Quantification of vacuoles showing RB accumulation. White: control; Gray: MyoF-KD. Three biological replicates were used; bars show means ± SD. All p-values ≤ 0.001 (***), using one-tailed unpaired Student’s t-test. Scale bars = 1 µm

In contrast, Golgi inheritance (marked by SortLR) remained unaffected by MyoF depletion (Figure 7C). Neither accumulation in the RB nor separation of maternal and de novo SortLR signals was observed, consistent with previous reports that Golgi duplicates and partitions independently (Pelletier et al., 2002).

To confirm that this phenotype results directly from MyoF depletion and not broader disruption of endomembrane architecture, we performed a rescue experiment. MIC2-Halo MyoF-mAID parasites were cultured in auxin for 24 hours to induce M-MIC2 retention in the RB, followed by auxin wash-off and continued replication (Figure 8A). Remarkably, M-MIC2 was redistributed to daughter cells in the following cycles, occasionally in an asymmetric fashion (Figure 8B), possibly reflecting unequal MyoF re-expression across daughters.

The residual body (RB) functions as a recycling center, not a dead end.

To assess the fate of material accumulated in the RB, auxin chase experiments were performed: 24h with auxin followed by 24h without. A) Representative images of MyoF-mAID MIC2-Halo parasites after 48h in control (no auxin), continuous auxin (Aux), or auxin washout (Aux washed). Magenta: M-MIC2, Green: n-MIC2. Continuous auxin led to M-MIC2 accumulation in the RB, while auxin washout enabled redistribution. B) Close-up of heterogeneous M-MIC2 redistribution post-auxin washout. C) Quantification of vacuoles showing normal, accumulated, or redistributed M-MIC2. White: control; Gray: MyoF-KD; Blue: MyoF-KD auxin chase. D) Time-lapse without auxin: M-MIC2 briefly passes through the RB (white arrows). E) Time-lapse with auxin: sustained M-MIC2 accumulation in the RB throughout replication. F) Time-lapse after auxin washout: initial M-MIC2 accumulation in the RB resolves by the second replication cycle, with redistribution observed. Three biological replicates were used; bars show means ± SD. Scale bars = 1 µm.

Live-cell imaging confirmed these dynamics (Video S4, S5). In untreated parasites, M-MIC2 entered the RB ∼7 h post-replication and was later redistributed to daughter cells (Figure 8D, Video S4). Under auxin treatment, M-MIC2 accumulated post-budding and failed to redistribute (Figure 8E; Video S5). Wash-off rescued redistribution during the second cycle, though visualization required intensity adjustments due to signal differences between accumulated and redistributed pools (Figure 8F; Video S6).

Together, these findings establish that MyoF is essential for recycling maternal micronemes and rhoptries via the RB, redefining the RB as a dynamic trafficking hub, not merely a degradative compartment. MyoF acts as a mechanistic bridge between F-actin dynamics and organelle inheritance, orchestrating precise redistribution of maternal cargo.

Discussion

Use of dual labelling to differentiate maternal and de novo material

Understanding the complexity of organelle inheritance in Toxoplasma gondii requires tools capable of resolving both the spatial and temporal dynamics of protein trafficking. Conventional single-labelling approaches have been limited to static snapshots, obscuring key processes such as organelle biogenesis, recycling, and selective degradation. To overcome these challenges, we employed a dual HaloTag pulse-chase labelling strategy to distinguish maternally inherited from de novo–synthesized proteins. This approach, previously underutilized in apicomplexan parasites (Koreny et al., 2023; Periz et al., 2019), allowed us to resolve protein fates across replication cycles and map distinct inheritance routes for multiple organelles.

The residual body as a regulatory hub

Our findings reaffirm the emerging view of the RB as an active, multifunctional structure rather than a passive cytoplasmic remnant (Frenal et al., 2017; Periz et al., 2017a). First described over 50 years ago (Sheffield and Melton, 1968), the RB was long believed to serve as a repository for discarded material, often called a “waste bin” (Nishi et al., 2008). However, we show that the RB temporarily stores maternal secretory organelles, such as micronemes and rhoptries, which are later redistributed to daughter cells in a MyoF-dependent manner.

Live-cell imaging revealed that organelles in the RB are not static but can re-enter functional pathways (Figures 8–10), establishing the RB as a trafficking checkpoint (Periz et al., 2019). This active role is further supported by the dependence of RB-mediated recycling on F-actin and the class XXII myosin MyoF, which we show to be essential for retrieval of maternal MIC2 and RON2 but dispensable for Golgi inheritance although we noticed a fragmentation of the Golgi, which has been described to depend on MyoF (Carmeille et al., 2021).

The RB is also critical for cytoplasmic continuity across parasites, synchronizing replication and facilitating inter-parasite protein exchange (Frenal et al., 2017; Muniz-Hernandez et al., 2011; Periz et al., 2017a). Its formation is dependent on actin-nucleating factors and myosins, indicating a complex and regulated origin rather than a passive leftover of cytokinesis (Stortz et al., 2019; Tosetti et al., 2019). Given these parallels, the RB may be functionally analogous to the mammalian midbody remnant, which regulates post-mitotic signaling and intracellular trafficking (Kuriyama et al., 2025; Peterman and Prekeris, 2019).

Fate determination: inheritance or degradation?

An important question raised by our study is how the parasite determines whether maternal proteins are recycled or degraded. Our results indicate that this decision is selective rather than stochastic (Figure 9). For instance, maternal micronemes persist for at least six replication cycles in the absence of MyoF, without signs of degradation. In contrast, IMC proteins like GAPM1a and IMC1 are lost within a single cycle and show no evidence of recycling. Degradation appears to occur within the RB itself, raising questions about the underlying mechanism. Although T. gondii encodes homologs of lysosomal and autophagic degradation machinery (Besteiro et al., 2011; Smith et al., 2021; Thaprawat et al., 2025), none have been localized to the RB or directly linked to IMC degradation. However, ubiquitin ligases and proteasomal components have been identified in the RB proteome (O’Shaughnessy et al., 2023; Que et al., 2002). IMC proteins are also part of the ubiquitinated proteome (Silmon de Monerri et al., 2015), and recent work showed that deletion of the kinase ERK7 prevents sequestration of the E3 ligase CSAR1, resulting in aberrant degradation of conoids (O’Shaughnessy et al., 2023). These findings support a model in which protein fate within the RB is determined through selective ubiquitination and regulated proteolysis.

Schematic summary of the three fates of the maternal organelles.

Group 1: Efficient, intact organelle recycling. De novo organelles are generated independently of the maternal organelle but in a similar location. Both de novo and maternal are distributed to the daughter cells. The Fluorescence intensity of the maternal organelle is relatively stable. Group 2: Even distribution of maternal material through organelle expansion with the insertion of the de novo material in the mother organelle. The Fluorescence intensity of the maternal organelle is divided by two at each replication step but their signal surface is increased by two. Group 3: Almost exclusive de novo synthesis. The de novo organelle is generated independently of the maternal organelle and in a different location. The maternal organelle is not clearly observable after a cycle of replication. The fluorescence intensity of the maternal organelle drastically drops after the first replication. Magenta: maternal organelle, Green: de novo organelle, Yellow: Colocalization between maternal and de novo.

A call to reassess recycling factors in Toxoplasma gondii

The recognition of the residual body (RB) as a key site for organelle recycling has fundamentally revised our understanding of intracellular trafficking in T. gondii (Periz et al., 2019; Tosetti et al., 2019). Earlier models emphasized de novo synthesis of organelles and proposed a repurposing of classical endocytic machinery for secretory functions (Tomavo, 2014). However, our HaloTag-based pulse-chase studies show that secretory organelles such as micronemes and rhoptries are extensively recycled via the RB, in a process dependent on F-actin and the unconventional myosin MyoF.

This shift in our understanding calls for a systematic re-evaluation of trafficking factors traditionally implicated in protein targeting, vesicle transport, and organelle biogenesis. Many of these particularly Rab-GTPases, SNAREs, and dynamins were previously studied without consideration of recycling pathways. Notably, accumulation of microneme material in the RB has been observed upon perturbation of several trafficking regulators, including Rab5A (Kremer et al., 2013), ArlX3 (Klinger et al., 2024), SORTLR (Sloves et al., 2012), VPS8 (Morlon-Guyot et al., 2018), and VPS11 (Morlon-Guyot et al., 2015). Yet, whether these phenotypes reflect a block in secretion, synthesis, or recycling remains unclear.

To resolve this, a targeted reanalysis of trafficking factors using our dual-labelling HaloTag assay combined with the splitCas9 system (Li et al., 2022) would allow high-throughput functional screening to specifically distinguish defects in recycling from those in de novo biogenesis, offering a path to redefine the roles of classical and lineage-specific trafficking regulators in T. gondii.

Material and Methods

Parasite culture and genetic manipulation

Growth and generation of transgenic T. gondii

T. gondii tachyzoites from the RH strain and derived lines, including RH Δku80/TATi and RH Δku80/Tir1, were maintained at 37 °C with 5% CO₂ in human foreskin fibroblasts (HFFs; ATCC, SCRC 1041) cultured in Dulbecco’s Modified Eagle Medium (DMEM; Sigma, D6546) supplemented with 10% fetal bovine serum (FBS; BioSell, FBS.US.0500), 4 mM L-glutamate (Sigma, G7513), and 20 μg/mL gentamycin (Sigma, G1397) as previously described (Gras et al., 2019).

Generation of transgenic parasites

New strains were generated using CRISPR/Cas9 as previously described (Li et al., 2022). Guide RNAs (gRNAs) targeting the regions of interest were designed using EuPaGDT (Alvarez-Jarreta et al., 2024). All gRNA and primer sequences are listed in Supplementary Table 1. Briefly, gRNA oligos were annealed, ligated into the Cas9-YFP vector, and verified by sequencing (Eurofins Genomics). Repair templates were generated by PCR amplification of the Halo or YFP tag flanked by 50 bp homology arms to the target gene, using Q5 High-Fidelity DNA Polymerase (New England BioLabs). PCR products were purified with a PCR purification kit (Blirt, EM26.1). Tachyzoites were mixed with the repair template and 10 µg of the corresponding Cas9 vector, and transfected using the Amaxa 4D-Nucleofector system (Lonza, AAF-1003X). Transfected parasites were allowed to invade fresh HFFs and replicate for 48 h. Following manual egress and filtration through a 3 μm filter, Cas9-YFP-expressing parasites were enriched by FACS (FACSAria III, BD Biosciences) and sorted into 96-well plates. Correct integration of the repair template was confirmed by PCR.

Labelling and characterization

Maternal and de-novo protein discrimination

Fresh tachyzoites expressing Halo-tagged reporters were mechanically egressed, filtered through a 3 μm filter, and resuspended in cold DMEM containing a membrane-permeable Halo dye, Janelia Fluor® 646 (1:1000, Promega), for 1 h. Parasites were centrifuged at 2,500 rpm for 5 min and washed three times with fresh medium to remove unbound dye. Parasites were then seeded onto HFF-covered Ibidi live-cell dishes for overnight replication. After replication, a second labeling was performed with a new membrane-permeable Halo dye, Janelia Fluor® 549 (1:1000, Promega) for 1 h at 37 °C, followed by three washes before imaging.

Fluorescence intensity across the replication

Parasites were labeled as described above and allowed to replicate for 24 h on HFF-coated Ibidi live-cell dishes. Approximately 15 fields of view were imaged, and around 25 vacuoles at stages 1, 2, 4, and 8 were analyzed. For each experiment, the maximum fluorescence intensity was measured. Regions overlapping with neighboring parasites or showing non-uniform signal distribution were excluded. Within each biological replicate, the highest recorded value was set to 100%, and all other values were expressed as percentages relative to this maximum. The experiment was performed in three independent biological replicates. Data are presented as mean ± SD.

Measurement of the signal area of the fluorescence

To quantify the average and total fluorescent area, images were analyzed in Fiji. About 25 isolated vacuoles per stage (1 to 8) were cropped and analyzed individually. A threshold was applied to the maternal signal, and both average and total signal area were measured. Three independent biological replicates were performed, and values are reported as mean ± SD.

Quantification of percentage of parasite per vacuole with maternal protein

Images were analyzed in Fiji to determine the proportion of parasites per vacuole with maternal protein signal. Approximately 25 vacuoles per stage (1 to 8) were cropped and analyzed individually. Vacuoles where all parasites retained maternal signal were scored as 100%. If some parasites lacked signal, the percentage of maternal-positive parasites was calculated. The experiment was performed in three independent biological replicates. Values are reported as mean ± SD.

Live replication assay

Parasites were labeled with Janelia Fluor® 646 (1:1000, Promega) for 1 h and washed three times before transfer to Ibidi live-cell dishes covered with HFFs. Parasites were allowed to invade for 1–2 h to obtain ∼10 parasites per field of view. Excess parasites were removed by washing three times. Imaging was performed on a Leica DMI8 microscope at 37 °C in 5% CO₂. Images were acquired every 15–30 min for 12 h using minimal laser power and exposure. Experiments were performed in three biological replicates and analyzed in Fiji.

Colocalization assays: ANKER1-Halo with HDEL-GFP

Parasites expressing ANKER1-Halo were transiently transfected with an HDEL-GFP plasmid to label the ER. After overnight replication in HFF-coated Ibidi dishes, parasites were labeled with Janelia Fluor® 646 (1:1000, Promega) for 1 h and washed three times prior to imaging. Z-stacks of HDEL-GFP-positive parasites were acquired, and colocalization was quantified in Fiji. Three biological replicates were performed.

Colocalization assays: MIC2/RON2 with Cb-Emerald

For colocalization with the F-actin network, MIC2 or RON2 was endogenously tagged and labeled in a strain stably expressing Cb-Emerald. Freshly egressed parasites were labeled with Janelia Fluor® 646 (1:1000, Promega) for 1 h, washed, and seeded onto HFF-coated live-cell dishes. After 24 h replication, parasites were fixed in 4% paraformaldehyde and imaged. Z-stacks of vacuoles showing clear F-actin networks were acquired, and colocalization was analyzed in the cytoplasm and residual body. Three biological replicates were performed. Data are reported as mean ± SD. Images were analyzed in Fiji.

Quantification of the maternal microneme distribution

To assess the distribution of maternal micronemes at stage 8, parasites were labeled as described in “Maternal and de novo protein discrimination.” Twenty-five stage 8 vacuoles were selected. Using the de novo MIC2 signal, the outline of each tachyzoite was traced, and the number of maternal MIC2-positive parasites was counted. Tachyzoites were numbered 1 to 8 from left to right. Three biological replicates were performed. Data are reported as mean ± SD. Images were analyzed in Fiji.

Phenotypic assays

Induction of MyoF KD

MyoF-mAID parasites were genetically modified to endogenously tag MIC2, RON2 and SortLR. As previously described for MyoF mAID parasites (Carmeille et al., 2021), MyoF-mAID MIC2/RON2 or SortLR-Halo parasites were induced +/- auxin for 4h prior to labelling and experiments.

Quantification of the accumulation of protein in the residual body

Parasites were treated ± auxin for 4 h and labeled as described above. After 24 h replication on live-cell dishes, ∼100 vacuoles were imaged across ∼15 fields of view. The percentage of vacuoles showing accumulation of material in the residual body was calculated. The experiment was performed with three independent biological triplicates. The values are then expressed as the mean values of the three independent experiments ± SD. Images were analysed via Fiji.

Auxin chased experiment

Parasites were induced +/- auxin for 4h prior labelling as describe above in “Maternal and de-novo protein discrimination”. After transfer on live cell dish and 48 h of replication +/- auxin, the parasites were images. For the chased condition, the parasites were grown for 24h under auxin. After this initial 24h, the dishes were washed three time to completely remove the auxin, a fresh media (without auxin) was added. Parasites were then allowed to replication for an extra 24h. About 15 fields of view were images for a total of about 100 vacuoles. The percentage of vacuole with accumulation in the residual body was calculated. Some parasites were exhibiting a redistribution where the microneme were no longer stuck in the residual body but not as homogeneously distributed as control parasite microneme. This profile was name “redistributed” and quantified along with the “accumulated” and “normal” patterns. The experiment was performed with three independent biological triplicates. The values are then expressed as the mean values of the three independent experiments ± SD. Images were analysed via Fiji.

MyoF-mAID Live replication assay

Parasites were induced +/- auxin for 4h prior labelling. Parasites were labelled with the Halo membrane permeable dye Halo 646 (Janelia Fluor®646 1:1000 Promega) for 1 h and washed three-time prior transfer onto a live cell dish covered with HFF. Parasites were allowed to invade for 1h-2h to obtain enough invasion events to reach an average of 10 parasites per FOV. After the invasion, the excess of parasites was removed by 3 washes. A new membrane permeable dye Halo 549 (Janelia Fluor®549 1:10000 Promega) was added to the media to record the de-novo generation of the POI live. Live cell dishes were then transferred to the Leica-DMI8, heated at 37°C under a chamber containing 5% CO2. Laser power and exposure were adjusted to the lowest values allowing reliable imaging. Imaged were taken every 15-30 minutes to follow the tagged protein behaviour during replication for 12h. For the Chased assay, the imaging started after the removal of the auxin after 24h of replication as described above in “Auxin chased experiment.” _The experiments were performed with three independent biological triplicates. Images were analysed via Fiji.

Imaging

Widefield microscopy

Unless stated otherwise, all images were acquired on a Leica-DMI8, objective 100x with the LasX software (v3.7.4). Fiji (v1.53c) was used to analyse the picture and all counts were made manually. LasX software (v.) from Leica was used to obtain parasite imaging data and

All images and movies were processed using Fiji (ImageJ) software v Image Processing Software (Schindelin et al., 2012).

Software

Fiji ( FIJI ImageJ v1.54f) was used to analyze the picture and all counts were made manually.

Data analysis

All data were plotted using Microsoft Excel

Statistical analysis

For statistic relevance, one-tailed unpaired Student’s T-tests were performed.

eLife Assessment

This study presents valuable new insights into the patterns of organelle inheritance in the protozoan parasite Toxoplasma gondii. An innovative dual-labeling approach used in this study to track maternal-derived and de novo synthesized organelles provides a technical advance with potential to be more broadly applied. Solid evidence is provided that different organelles show distinct inheritance fates during cell replication; however, the data describing the residual body component in this process is incomplete.

Reviewer #1 (Public review):

Summary:

This work asks the question of how different organelles and structures in the apicomplexan parasite Toxoplasma gondii are recycled and/or segregated to the daughter cells during cell replication. In particular, they consider an unusual cell structure called the residual body that links replicating cells during the intracellular infection stage of this parasite. The residual body has historically been considered a 'dumping ground' for unnecessary relics of the mother cell during division, but this notion is increasingly being revised. Indeed, cell replication in Toxoplasma is often misinterpreted as cell division (cytokinesis), but in fact, the cell replicates its organelles and structures to multiple 10s of copies in seemingly distinctly formed daughter cells, but cytokinesis is delayed for many such cycles and typically only occurs simultaneously with parasite egress from its host cell. The residual body is, in fact, the connection between these pre-cytokinetic replicated daughters, and effectively, this is still a single cell at this stage. The authors have previously shown that an actin network extends through the residual body between these daughter cells, and ER and mitochondria common to all cells are also linked through this structure. This study examining the fates of organelles during cell replication is timely for continuing our understanding of how this fascinating component of the cell participates in these processes. The authors use Halo-tags as their principal tool to track discrete populations of proteins, labelling their organelle locations, and this provides beautiful insight into these processes.

Strengths:

Using dyes conjugated to Halo tags, this work elegantly tracks the fates of proteins synthesised by an original 'mother' cell over several replication cycles of pre-cytokinetic 'daughters'. Using this tool, they show that some organelles are made intact just once and that some of these can be subsequently sorted to the daughters (micronemes and rhoptries) while others are dismantled (IMC) and the daughters must make their own. A third set of organelles (largely synthesis, sorting, and metabolic compartments) is divided and inherited, and new daughter-synthesised proteins are added to the preexisting maternal proteins in these structures. A role for actin and myosin is clearly demonstrated for micronemes and rhoptries, and this correlates with their relatively late inheritance into the developing daughters. Overall, this work gives clarity to the behaviours of several cell structures during replication and paves the way to a better understanding of the mechanisms that drive the differences between structures and the universality of these processes in other apicomplexan parasites.

Weaknesses:

In addressing the question of residual body participation in sorting of organelles, it would be useful to clearly define this structure and when and where it is delineated from the posterior of a mother cell during the formation of daughter structures. This might seem like a moot point, but it would give clarity to notions of recycling and 'reservoirs'. Mother cells retain their active invasion apparatus until very late in daughter formation, and the need for micronemes and rhoptries to be released from this service late in the process might explain why they are only then trafficked to the cell posterior and then into the daughters. So, is this a distinct 'residual body' body function/reservoir or just a spatial constraint of this sequence of daughter formation? In subsequent cell replications (4, 8, 16... stages), is there a separation between the residual body that links them all and the posterior of each new 'mother cell', and if so, when is this distinction lost? This is important because without a definition, we might be confusing different processes. Are rhoptries/micronemes that originate in one 'mother' able to be sorted to the 'daughters' from a distinct mother in this syncytium? If so, this would make it a sorting centre, but otherwise we could be just capturing the activities at the posterior of any given cell during replication. The authors' further thoughts on this would be very interesting.

The Group 2 structures are described as those that are divided between daughters and receive newly synthesised proteins that add to the maternal protein of these compartments. While this is a logical conclusion for several that are mentioned, where the maternal protein signal is seen to be depleted with replication (including for the apicoplast, ER, glideosome, and Golgi). Data for the addition of new proteins to these existing structures is actually only presented in direct support of this for the Golgi.

Reviewer #2 (Public review):

Summary:

Toxoplasma gondii is an obligate intracellular parasite and the causative agent of Toxoplasmosis. Parasite invasion into host cells, intracellular replication, and then egress, which results in the destruction of the infected cell, is central to pathogenicity. This manuscript focuses on understanding how maternal resources (in this case, cellular organelles) are shared between daughter parasites during cell division. Many organelles are single copy, meaning that division and inheritance by the daughters is crucial for successful replication. The major strength of this study was the use of a Halo-based pulse chase assay to characterize patterns of organelle inheritance. The results show that both microneme and rhoptries (secretory vesicles) previously thought to be synthesized de novo are inherited by daughter parasites. Thus, this paper adds new insight to our understanding of cell division in this important parasite.

Strengths:

This study demonstrated that pulse labeling of proteins can be used to monitor protein synthesis, turnover, and movement. This approach will be of great interest to the field. Using this method, the authors demonstrate three main modes of organelle inheritance.

(1) Organelles, where there are multiple copies (such as secretory vesicles, micronemes, and rhoptries), are divided between the daughter parasites, with additional contribution of newly formed vesicles. New and old material remain as separate entities in the cell.

(2) Single-copy organelles, which are expanded to include newly synthesized material prior to division, such as the Golgi and apicoplast.

(3) Cytoskeletal structures that are synthesized anew during each round of division. These studies provide more refined insight into patterns or organelle inheritance and demonstrate that secretory organelles are not made de novo during each round of division as previously thought. The paper has a logical flow, and overall, the data is presented in a clear and organized fashion.

Weaknesses:

(1) Descriptions of methodology and statistical analysis were incomplete.

(2) There are inconsistencies between the data in Figures 1 and 5. In Figure 1, a small amount of maternal IMC is visible in stage 2 parasites. Although this is a ~90% reduction, these parasites should be quantified as parasites with material IMC. However, the graph in Figure 5C indicates that no material parasites have GAPM1a, given that graph 5C is a binary measure (present vs. absent), one would expect a non-zero percent of parasites to have maternal material.

(3) The conclusion from Figure 6 was not justified based on the data. I agree with the author's conclusion that the accumulation of micronemes and rhoptries in the residual body was time-dependent. In Figure 6A, the signal observed in the residual body at times 6:30, 13, and 14 hours is not observed in subsequent time points. However, the fate of these micronemes and rhoptries is unclear. It cannot be concluded that these vesicles are recycled back to the mother. They could also have been degraded. In fact, the graphs of microneme inheritance in Figure 2B show a decrease in maternal signal from 100% to 80% between stages 1 and 2, indicating that some microneme degradation is taking place.

(4) To convincingly demonstrate that the redistribution of micronemes and rhoptries was due to recovery of MyoF protein levels after auxin washout, a Western blot should be performed to show MyoF protein levels over time. In addition, the decrease in mMIC2 protein levels in the residual body in Figure 8F should be measured and normalized for photobleaching. Both apical and basal signals appear to be reduced over the time course of imaging.

Reviewer #3 (Public review):

Summary:

Knoerzer-Suckow et al. explore the mechanisms of organelle inheritance during endodyogeny in Toxoplasma gondii using an innovative dual-labeling approach to track the distribution of maternal organelles into daughter parasites. They can clearly distinguish between maternal and daughter-derived organelles using their dual-labeling Halo Tag approach. They reveal that different organelles are trafficked to daughter parasites in three broad patterns, which they have binned into groups. Their findings reveal a role for MyoF in the inheritance of micronemes and rhoptries, and notably, they observe that the inner membrane complex (IMC) is not recycled. Instead, the IMC undergoes a pronounced relocalization to the posterior of the maternal cell, where it is likely targeted for degradation.

Strengths:

The data surrounding their MyoF knockdown experiments, IMC degradation, and trafficking of MIC2 after auxin washout are compelling. These data add to the knowledge of how organelle inheritance occurs in T. gondii, increasing the field's understanding of endodyogeny.

Weaknesses:

(1) The evidence provided to support the claim that microneme and rhoptry inheritance specifically traffics through the residual body does not sufficiently substantiate the claim. The temporal resolution of the imaging is inadequate to precisely trace the path of microneme and rhoptry inheritance. From the data shown in the manuscript, it can be concluded that at least some of the micronemes and rhoptries might be recycled through the residual body, but it is unclear whether many or most of these organelles do so.

(2) The absence of specific markers for the residual body brings into question whether microneme inheritance occurs through a discrete residual body or simply via the basal end of the maternal parasite. The authors need a robust way to visualize and define the residual body to claim that micronemes and rhoptries are specifically transported through this structure.

Author response:

The following is the authors’ response to the original reviews.

We sincerely thank the reviewers and the Reviewing Editor for their careful evaluation of our manuscript and for their constructive and insightful comments. Their suggestions have helped us to improve the clarity, rigor, and presentation of our work. In response to these comments, we have substantially revised the manuscript and performed several additional analyses and experiments, as summarized below.

Major additions and modifications made during revision

In response to the reviewers' comments, we have substantially revised the manuscript and performed several additional analyses and experiments:

New analyses

- Quantification of MyoF recovery following auxin washout using MyoF-mAID-HA immunofluorescence (Figure 8B, Figure S10D).

- Quantification of maternal MIC2 fluorescence intensity following 24 h MyoF depletion and subsequent redistribution after auxin washout (150 micronemes per condition; Figure S10A,B).

- Pearson correlation analysis of ANKER1-Halo and HDEL-GFP localization (Pearson's R = 0.92 ± 0.04; n = 30 parasites).

- Additional probability-based analysis supporting regulated microneme inheritance.

- Expanded analysis of microneme redistribution across larger replication stages (Figure S5D).

New figures

- Figure S6: Dual-labelling analysis of additional Group 2 organelles (ER, apicoplast, glideosome).

- Figure S10A, B: MIC2 fluorescence intensity analysis following RB retention and redistribution.

- Figure S10D: Correlation between MyoF recovery and phenotype rescue.

- Figure S11: Schematic overview of quantification and analysis workflow.

Additional experimental efforts

- Generation of a MIC2-Halo / IMC1-mKATE / Cb-Emerald parasite line to improve visualization of RB-associated trafficking.

- Multiple attempts to perform higher-temporal-resolution live-cell imaging. However, prolonged acquisition resulted in severe phototoxicity, replication arrest, and parasite death, preventing reliable long-term recordings.

Textual and methodological revisions

- Expanded Materials and Methods section with detailed descriptions of fluorescence quantification, colocalization analyses, and statistical procedures.

- Re-evaluation of statistical analyses using two-tailed tests throughout.

- Revision of manuscript text to clarify the evidence supporting RB-associated trafficking and to better acknowledge current limitations.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This work asks the question of how different organelles and structures in the apicomplexan parasite Toxoplasma gondii are recycled and/or segregated to the daughter cells during cell replication. In particular, they consider an unusual cell structure called the residual body that links replicating cells during the intracellular infection stage of this parasite. The residual body has historically been considered a 'dumping ground' for unnecessary relics of the mother cell during division, but this notion is increasingly being revised. Indeed, cell replication in Toxoplasma is often misinterpreted as cell division (cytokinesis), but in fact, the cell replicates its organelles and structures to multiple 10s of copies in seemingly distinctly formed daughter cells, but cytokinesis is delayed for many such cycles and typically only occurs simultaneously with parasite egress from its host cell. The residual body is, in fact, the connection between these pre-cytokinetic replicated daughters, and effectively, this is still a single cell at this stage. The authors have previously shown that an actin network extends through the residual body between these daughter cells, and ER and mitochondria common to all cells are also linked through this structure. This study examining the fates of organelles during cell replication is timely for continuing our understanding of how this fascinating component of the cell participates in these processes. The authors use Halo-tags as their principal tool to track discrete populations of proteins, labelling their organelle locations, and this provides beautiful insight into these processes.

Strengths:

Using dyes conjugated to Halo tags, this work elegantly tracks the fates of proteins synthesised by an original 'mother' cell over several replication cycles of pre-cytokinetic 'daughters'. Using this tool, they show that some organelles are made intact just once and that some of these can be subsequently sorted to the daughters (micronemes and rhoptries) while others are dismantled (IMC) and the daughters must make their own. A third set of organelles (largely synthesis, sorting, and metabolic compartments) is divided and inherited, and new daughter-synthesised proteins are added to the preexisting maternal proteins in these structures. A role for actin and myosin is clearly demonstrated for micronemes and rhoptries, and this correlates with their relatively late inheritance into the developing daughters. Overall, this work gives clarity to the behaviours of several cell structures during replication and paves the way to a better understanding of the mechanisms that drive the differences between structures and the universality of these processes in other apicomplexan parasites.

Weaknesses:

In addressing the question of residual body participation in sorting of organelles, it would be useful to clearly define this structure and when and where it is delineated from the posterior of a mother cell during the formation of daughter structures. This might seem like a moot point, but it would give clarity to notions of recycling and 'reservoirs'. Mother cells retain their active invasion apparatus until very late in daughter formation, and the need for micronemes and rhoptries to be released from this service late in the process might explain why they are only then trafficked to the cell posterior and then into the daughters. So, is this a distinct 'residual body' body function/reservoir or just a spatial constraint of this sequence of daughter formation? In subsequent cell replications (4, 8, 16... stages), is there a separation between the residual body that links them all and the posterior of each new 'mother cell', and if so, when is this distinction lost? This is important because without a definition, we might be confusing different processes.

We thank the reviewer for this excellent and thoughtful question. The residual body (RB) emerges at the end of the first replication cycle, where it is delineated by the basal complex and persists as an IVN-associated compartment connecting all daughter parasites through both plasma membrane and cytoplasm. Previous EM and live-cell studies, including ours, have shown that the RB is not a passive remnant but a dynamic structure dependent on F-actin and unconventional myosins, supporting recycling, inter-parasite connectivity, and synchronous growth (Delbac et al., 2001; Muñiz-Hernández et al., 2011; Frénal et al., 2017; Periz et al., 2017).

In the present study, the MyoF reversibility experiment provides strong support for a model in which RB functions as an active recycling hub. Upon MyoF depletion, maternal microneme and rhoptry proteins accumulate within the RB. Following restoration of MyoF expression, this material is redistributed to daughter organelles. We interpret this reversible phenotype as evidence that the RB represents a distinct and regulated trafficking intermediate rather than simply a by-product of late daughter cell formation.

We agree with the reviewer that mother cells retain a functional invasion apparatus until very late during daughter formation, and that the delayed release of micronemes and rhoptries likely contributes to their late trafficking toward the cell posterior. However, our data indicate that once released, these organelles transit through a defined RB compartment that actively participates in their recycling rather than merely reflecting positional constraints. This has been previously well illustrated for micronemes, which are trafficked along F-actin filaments within the residual body (Periz et al., 2019).

At later rounds of replication (4, 8, 16 parasites), previous studies have demonstrated the presence of multiple residual body centres within the same vacuole. However, the precise temporal and structural distinction between the RB linking parasites within the vacuole and the posterior of newly formed mother cells remains insufficiently resolved and is beyond the scope of the present study. Importantly, available ultrastructural and live-cell imaging supports the persistence of shared RB compartments connecting parasites within a vacuole, arguing against a simple conflation of posterior membranes and residual body material.

While the primary aim of the current work was to investigate the RB's role in organelle recycling, we fully agree that a more precise definition of when and how the RB is formed, remodelled, and ultimately resolved during successive replication cycles will be essential to distinguish recycling from spatial constraints. We have revised the Discussion to better acknowledge this limitation and to avoid overinterpreting the role of the RB in organelle inheritance.

Are rhoptries/micronemes that originate in one 'mother' able to be sorted to the 'daughters' from a distinct mother in this syncytium? If so, this would make it a sorting centre, but otherwise we could be just capturing the activities at the posterior of any given cell during replication. The authors' further thoughts on this would be very interesting.

We agree with the reviewer that our current data do not definitively demonstrate whether rhoptries or micronemes originating from one “mother” parasite can be redistributed to daughters derived from another mother within the same syncytial vacuole. Nevertheless, our MyoF chase experiments are consistent with a model in which the RB/IVN functions as an active recycling and sorting hub rather than simply representing posterior trafficking events associated with individual parasites.

Upon MyoF depletion, maternal micronemes accumulated within the RB. Following restoration of MyoF expression, these accumulated micronemes were subsequently redistributed to daughter parasites. This reversible redistribution is more consistent with an active recycling process than with passive accumulation alone.

To further support this interpretation, we expanded the analysis presented in Figure S5 by including additional vacuoles and larger replication stages (new panel D). These analyses show that maternal micronemes are redistributed broadly and relatively evenly among daughter parasites. We additionally performed a probability-based analysis demonstrating that the recurrent and homogeneous redistribution patterns observed are highly unlikely to arise from stochastic capture events occurring independently at the posterior end of each parasite during replication. Together, these analyses support the interpretation that microneme redistribution is a regulated process.

Direct demonstration of recycling between all parasites within a vacuole would require a system allowing simultaneous differential labeling of (i) daughter parasites derived from a specific mother cell and (ii) the maternal organelles originating from that same mother during a subsequent replication cycle. To our knowledge, such an approach is not currently technically feasible. Nevertheless, our live-cell imaging experiments provide additional support for communal redistribution, as microneme material accumulated within the RB was subsequently observed redistributing, albeit unevenly, across multiple tachyzoites within the same vacuole.

The Group 2 structures are described as those that are divided between daughters and receive newly synthesised proteins that add to the maternal protein of these compartments. While this is a logical conclusion for several that are mentioned, where the maternal protein signal is seen to be depleted with replication (including for the apicoplast, ER, glideosome, and Golgi). Data for the addition of new proteins to these existing structures is actually only presented in direct support of this for the Golgi.

We thank the reviewer for this important clarification. We initially selected the Golgi as a representative example because its morphology and restricted localization provide the clearest visualization of the dual-labeling dynamics. However, the same experimental approach was applied to all Group 2 organelles analyzed in this study. To address the reviewer's concern more directly, we have now included a new supplementary figure (Figure S6) showing that the same pattern is also observed for the apicoplast, ER, and glideosome.

We would also like to clarify that the maternal protein signal is not lost during replication but instead becomes progressively diluted as these organelles expand, are partitioned into daughter parasites, and incorporate newly synthesized proteins. The Golgi was originally highlighted because these dynamics are most readily visualized in this compartment; however, the same principle applies to all Group 2 organelles analyzed in this study, as now illustrated in Figure S6.

Reviewer #2 (Public review):

Summary:

Toxoplasma gondii is an obligate intracellular parasite and the causative agent of Toxoplasmosis. Parasite invasion into host cells, intracellular replication, and then egress, which results in the destruction of the infected cell, is central to pathogenicity. This manuscript focuses on understanding how maternal resources (in this case, cellular organelles) are shared between daughter parasites during cell division. Many organelles are single copy, meaning that division and inheritance by the daughters is crucial for successful replication. The major strength of this study was the use of a Halobased pulse chase assay to characterize patterns of organelle inheritance. The results show that both microneme and rhoptries (secretory vesicles) previously thought to be synthesized de novo are inherited by daughter parasites. Thus, this paper adds new insight to our understanding of cell division in this important parasite.

Strengths:

This study demonstrated that pulse labeling of proteins can be used to monitor protein synthesis, turnover, and movement. This approach will be of great interest to the field. Using this method, the authors demonstrate three main modes of organelle inheritance.

(1) Organelles, where there are multiple copies (such as secretory vesicles, micronemes, and rhoptries), are divided between the daughter parasites, with additional contribution of newly formed vesicles. New and old material remain as separate entities in the cell.

(2) Single-copy organelles, which are expanded to include newly synthesized material prior to division, such as the Golgi and apicoplast.

(3) Cytoskeletal structures that are synthesized anew during each round of division. These studies provide more refined insight into patterns or organelle inheritance and demonstrate that secretory organelles are not made de novo during each round of division as previously thought. The paper has a logical flow, and overall, the data is presented in a clear and organized fashion.

Weaknesses:

(1) Descriptions of methodology and statistical analysis were incomplete.

We agree with the reviewer that the description of the methodology and statistical analyses required further clarification. To address this, we have added a new supplementary figure (Figure S11) illustrating the experimental workflow, quantification strategy, and analysis pipeline. We have also expanded the Materials and Methods section to provide detailed descriptions of the experimental design, fluorescence quantification procedures, statistical analyses, and the number of biological replicates. These revisions provide a clearer and more comprehensive description of the methodology and data analysis.

(2) There are inconsistencies between the data in Figures 1 and 5. In Figure 1, a small amount of maternal IMC is visible in stage 2 parasites. Although this is a ~90% reduction, these parasites should be quantified as parasites with material IMC. However, the graph in Figure 5C indicates that no material parasites have GAPM1a, given that graph 5C is a binary measure (present vs. absent), one would expect a non-zero percent of parasites to have maternal material.

We agree with Reviewer 2 that, based on the raw fluorescence signal, one might expect a non-zero percentage of parasites to retain maternal IMC material after the first replication. The apparent discrepancy between Figures 1 and 5 reflects our thresholding strategy rather than inconsistent data.

Figure 5C presents a binary analysis (presence versus absence) using a threshold calibrated from stage 1 parasites and applied uniformly across all markers. Under this criterion, the residual GAPM1a signal after the first replication falls below the detection threshold, resulting in 0% positive vacuoles. Although normalization to stage 2 parasites would detect this weak residual signal, such a protein-specific threshold would compromise direct comparison across the dataset.

To clarify this point, we have updated the Figure 5C legend to explain the analytical approach and the asterisk associated with GAPM1a. The residual maternal IMC signal visible in Figure 1 represents a rare example selected to illustrate the remaining ~10% signal and is consistent with the absence of detectable maternal IMC1 after replication in Figures 2C and 5E.

(3) The conclusion from Figure 6 was not justified based on the data. I agree with the author's conclusion that the accumulation of micronemes and rhoptries in the residual body was timedependent. In Figure 6A, the signal observed in the residual body at times 6:30, 13, and 14 hours is not observed in subsequent time points. However, the fate of these micronemes and rhoptries is unclear. It cannot be concluded that these vesicles are recycled back to the mother. They could also have been degraded. In fact, the graphs of microneme inheritance in Figure 2B show a decrease in maternal signal from 100% to 80% between stages 1 and 2, indicating that some microneme degradation is taking place.

We agree with the reviewer that Figure 6 alone does not definitively establish the fate of micronemes and rhoptries accumulating within the residual body (RB), and that both recycling and degradation remain possible interpretations. Our conclusion that maternal micronemes are predominantly recycled is therefore based on the integration of Figure 6 with our MyoF depletion and recovery experiments, additional quantitative analyses, and previous work demonstrating F-actin-dependent microneme trafficking through the RB (Periz et al., 2019).

Consistent with this model, MyoF depletion results in the accumulation of maternal micronemes within the RB, whereas restoration of MyoF expression following auxin washout leads to their redistribution across multiple tachyzoites within the same vacuole (Figure 8). Furthermore, maternal microneme signal remains detectable even after prolonged MyoF depletion (up to 48 h) and multiple rounds of replication (Figures 7 and 8), arguing against extensive degradation.

To further address this possibility, we quantified the fluorescence intensity of individual maternal MIC2-positive micronemes retained within the RB after 24 h of MyoF depletion and following redistribution after auxin washout (150 micronemes per condition). No significant difference in fluorescence intensity was observed compared with control maternal micronemes (Figure S10A,B), indicating that maternal microneme signal is preserved during RB retention and redistribution.

We therefore interpret the decrease in maternal microneme signal observed between stages 1 and 2 in Figure 2B primarily as a consequence of redistribution and dilution rather than degradation, consistent with the stable fluorescence intensity of individual micronemes (Figure 3). Regarding rhoptries, we note that the majority (~90%) are incorporated into daughter parasites before budding is complete, limiting their accumulation within the RB and suggesting that RB-associated trafficking primarily reflects redistribution rather than bulk degradation.

(4) To convincingly demonstrate that the redistribution of micronemes and rhoptries was due to recovery of MyoF protein levels after auxin washout, a Western blot should be performed to show MyoF protein levels over time. In addition, the decrease in mMIC2 protein levels in the residual body in Figure 8F should be measured and normalized for photobleaching. Both apical and basal signals appear to be reduced over the time course of imaging.

We agree with the reviewer that demonstrating MyoF recovery following auxin washout is important. Rather than performing a Western blot, we monitored MyoF recovery by immunofluorescence using the HA tag in the MyoF-mAID-HA strain, allowing direct correlation between MyoF reappearance and microneme redistribution at the single-vacuole level. These data are now included in Figure 8B, with the corresponding MyoF presence–phenotype association analysis presented in Figure S10D.

Regarding photobleaching, we agree that fluorescence loss during time-lapse imaging is an important consideration. However, in this experiment, changes in fluorescence intensity reflect not only photobleaching but also biological redistribution of micronemes and movement of parasites in and out of the imaging plane. In the absence of a stable internal reference fluorophore, applying a standard photobleaching correction could therefore introduce additional inaccuracies. For this reason, we did not quantify fluorescence intensity during the redistribution phase.

Instead, to assess whether maternal microneme signal is lost during RB retention and redistribution, we quantified the fluorescence intensity of individual maternal MIC2-positive micronemes following 24 h of MyoF depletion and subsequent auxin washout (Figure S10A,B). No significant difference was observed compared with control maternal micronemes, supporting the conclusion that redistribution occurs without substantial loss of the maternal microneme pool.

Reviewer #3 (Public review):

Summary:

Knoerzer-Suckow et al. explore the mechanisms of organelle inheritance during endodyogeny in Toxoplasma gondii using an innovative dual-labeling approach to track the distribution of maternal organelles into daughter parasites. They can clearly distinguish between maternal and daughterderived organelles using their dual-labeling Halo Tag approach. They reveal that different organelles are trafficked to daughter parasites in three broad patterns, which they have binned into groups. Their findings reveal a role for MyoF in the inheritance of micronemes and rhoptries, and notably, they observe that the inner membrane complex (IMC) is not recycled. Instead, the IMC undergoes a pronounced relocalization to the posterior of the maternal cell, where it is likely targeted for degradation.

Strengths:

The data surrounding their MyoF knockdown experiments, IMC degradation, and trafficking of MIC2 after auxin washout are compelling. These data add to the knowledge of how organelle inheritance occurs in T. gondii, increasing the field's understanding of endodyogeny.

Weaknesses:

(1) The evidence provided to support the claim that microneme and rhoptry inheritance specifically traffics through the residual body does not sufficiently substantiate the claim. The temporal resolution of the imaging is inadequate to precisely trace the path of microneme and rhoptry inheritance. From the data shown in the manuscript, it can be concluded that at least some of the micronemes and rhoptries might be recycled through the residual body, but it is unclear whether many or most of these organelles do so.

We thank the reviewer for this important comment and refer also to our response to Reviewer 1 above.

Previous work has demonstrated F-actin-dependent trafficking of micronemes within the residual body (RB) (Periz et al., 2019). Consistent with these findings, our data support a model in which RB-mediated trafficking contributes to maternal microneme recycling. We acknowledge, however, that the temporal resolution of our imaging does not allow continuous tracking of every individual organelle throughout the entire replication process.

In contrast, our observations indicate that the majority of maternal rhoptry material is incorporated into daughter cells before replication is complete and therefore does not necessarily transit through the RB under normal conditions (Figure 6). Nevertheless, rhoptry inheritance remains dependent on the actin–MyoF trafficking machinery, as MyoF depletion results in the accumulation of rhoptry material within the RB (Figures 6 and 7).

Taken together, our data support a model in which the RB serves as an important recycling hub for maternal micronemes and can, under conditions of impaired trafficking, also transiently accommodate rhoptry material. However, our imaging resolution does not allow us to conclude that all microneme or rhoptry inheritance obligatorily transits through the RB, and we have revised the manuscript to reflect this limitation more explicitly.

(2) The absence of specific markers for the residual body brings into question whether microneme inheritance occurs through a discrete residual body or simply via the basal end of the maternal parasite. The authors need a robust way to visualize and define the residual body to claim that micronemes and rhoptries are specifically transported through this structure.

We agree with the reviewer that the absence of a dedicated residual body (RB) marker remains a limitation and that such a tool would improve the precision of our analyses. To date, no specific RB marker has been identified (see also our response to Reviewer 1). The most reliable proxy currently available is the F-actin chromobody, which labels the dense F-actin network associated with the RB. Using this approach, previous work demonstrated F-actin-dependent trafficking of micronemes within the RB (Periz et al., 2019).

Building on these findings, our data support a model in which RB-associated trafficking contributes to maternal microneme recycling, whereas rhoptries are more frequently incorporated directly into daughter cells without obvious RB transit. In addition, functional perturbation of the actin–MyoF transport machinery, through MyoF depletion and subsequent recovery, supports the interpretation that the RB represents a discrete actin-associated compartment involved in organelle redistribution. Nevertheless, we acknowledge that our current imaging resolution does not allow us to determine the extent to which all microneme or rhoptry inheritance occurs through the RB, and we have revised the manuscript accordingly.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Comments for revision where either the clarity or accuracy could be improved:

(1) The methods could do with some further detail with respect to the fluorescence intensity measurement. For example, where the Z-series was taken, and for measurements, were maximum projects taken or single Z-planes? Were all measurements made on unprocessed images, or was any deconvolution, etc, undertaken?

We updated the Materials and Methods for better clarity and now read as: “Parasites were labeled as described above and allowed to replicate for 24 h on HFF-coated Ibidi live-cell dishes. Approximately 15 fields of view were imaged using Z-stacks spanning 3 μm centred on the vacuoles. For each replication stage (1, 2, 4, and 8 parasites per vacuole), individual tachyzoites were sampled across multiple vacuoles.

Maximum-intensity projections were generated from non-deconvolved images. Fluorescence intensity (FI) was quantified as the maximum grey value measured within regions of interest (ROIs) drawn on individual tachyzoites from each vacuole stage. ROIs excluded overlapping parasites, neighbouring vacuoles, and regions with atypical signal intensity. For each biological replicate, up to 25 tachyzoites per replication stage were analyzed, and the mean FI value was calculated for each stage. The highest mean FI observed among the stages within a replicate was defined as 100%, and FI values for the other stages were expressed relative to this maximum. Relative FI values were then averaged across three independent biological replicates. Data are presented as mean ± SD.”

(2) Line 93: Is more JF549 added at each of stages 2, 4, and 8? I assume so to get the progressive increase, but it would help to clarify this here.

As illustrated in Figure 1a, the second dye is added only once, after 24 h of replication, and is then washed off before imaging. It is not maintained throughout the replication steps. The observed increase in signal is related to the presence of newly synthesized (de novo) proteins generated during replication. The Halo tags of these new proteins are initially free of ligand, as no ligand is present during replication. During the second labelling step, all free Halo tags can bind the dye. The signal intensity at each replication step therefore reflects the amount of de novo material present, which is higher at step 4 than at step 2, because the proteins generated de novo during stage 2 are also present in stage 4. The signal will be determined by both the number of newly produced molecules and their concentration at the same localization.

(3) Line 103: Check that the Carruthers and Sibley, 1997, ref for Tic20 in the apicoplast is correct. I don't think this could be correct given the date.

The reference it will be corrected to G van Dooren et al. 2008.

(4) Figure legends: It would be useful to state what form of microscopy was used in each figure.

Following the reviewer's advice the legends has been updated.

(5) Figure 2, S2: How are single organelles tracked, such as rhoptries? I'd assume that a cell will gain new de novo organelles as well, and that this would reduce the signal per cell. Stage 2 has a rhoptry signal in both daughters, so I'd expect the signal to be roughly half for the whole cell, unless the authors can resolve individual rhoptries (which would surprise me with this microscopy). If individual rhoptries were resolved, how was this done, and what was the confidence in this (were there controls?)

We thank the reviewer for this important point. We do not resolve individual rhoptries with the imaging conditions used in this study. Instead, fluorescence intensity was measured as the maximum grey value within a representative region of interest (ROI) encompassing the apical rhoptry signal while excluding overlapping parasites and regions with atypical fluorescence intensity. The same ROI selection strategy was applied consistently across all replication stages, allowing direct comparison with stage 1 parasites.

The analyses presented in Figures 3 and S2 show an increasing proportion of tachyzoites lacking detectable maternal rhoptry signal as replication progresses, while the fluorescence intensity of the remaining maternal signal remains relatively stable. Together, these observations are consistent with the redistribution of intact maternal rhoptries rather than a progressive loss of rhoptry fluorescence.

(6) Line 150: it is unclear what is meant by 'regulated partitioning'. The more equal inheritance of micronemes versus rhoptries might not indicate a 'regulated partitioning' but just a more uniform distribution, given the larger number of micronemes versus rhoptries.

We agree with the reviewer that this statement required clarification, and we have revised the text accordingly. Our intention was to emphasize that microneme inheritance appears to be a regulated process, rather than to directly compare it with rhoptry inheritance. The observed differences between these organelles are likely influenced, at least in part, by their different abundances.

Across successive rounds of replication, daughter parasites consistently inherit comparable amounts of maternal micronemes, even at later replication stages (Figure S4). Given that a single mother parasite contains approximately 30–40 micronemes, whereas successive rounds of endodyogeny can generate up to 32 daughter parasites, a purely stochastic segregation would be unlikely to produce the relatively uniform distribution observed (~1–2 maternal micronemes per tachyzoite). To support this interpretation, we performed an additional probability-based analysis, which indicates that the observed redistribution patterns are unlikely to arise by chance alone. We therefore interpret these findings as supporting the existence of mechanisms that promote balanced microneme inheritance during parasite replication.

(7) Line 184: How is the maternal signal measured without detecting the internal daughter signal? Is this an average signal for the full parasite, or just for a cross-section of the IMC? And if the latter, how are the different profiles of mother and daughter accounted for? Also, the abbreviation in the brackets doesn't make sense here.

We thank the reviewer for this important point. We have revised the Materials and Methods section to provide a clearer description of the fluorescence intensity (FI) measurements and added a new supplementary figure (Figure S11) illustrating the analysis workflow.

Briefly, FI measurements were performed on maximum-intensity projections generated from Z-stack images without deconvolution. A representative region of interest (ROI) was selected, and the maximum grey value was used for quantification. This approach minimizes variability arising from differences in ROI size and provides a robust metric for comparison across replication stages.

Daughter cell fluorescence was measured using the same approach while excluding overlapping signals from neighboring daughter cells and the maternal IMC. Maternal and daughter signals were distinguished based on their spatial localization and fluorescence labeling. Finally, the abbreviation in brackets has been corrected for clarity.

(8) Line 191: Subheading a bit unclear. Distinct from other organelles, or are miconeme and rhoptry pathways distinct from each other?

We agree with the reviewer and have updated the subheading to “Whole-organelle inheritance of micronemes and rhoptries occurs via distinct recycling pathways”

(9) Line 193: The site of disassembly of the IMC (suggested RB here) might not be the same as the site of degradation. I suggest using 'disassembly' instead here.

We agree that “disassembly” is an appropriate term to describe the breakdown of the IMC at the residual body (RB). However, we also believe that the RB represents the primary site of IMC degradation, for two reasons. First, if IMC material were not degraded at this site, we would expect to detect Halo-positive signal elsewhere following IMC collapse, which we do not observe. Second, transport of IMC material to an alternative degradation site would be required, but no IMC-positive vesicles are observed, arguing against significant redistribution. Together, these observations support the conclusion that the RB is both the site of disassembly and degradation of maternal IMC.

(10) Line 215: The conclusion for a difference in timing of microneme and rhoptry segregation is not clearly supported by the data presented. Also, if there are more micronemes than rhoptries, then the frequency of observing a microneme being trafficked through the RB would need to be higher than for rhoptries if the mechanisms were the same. So, a difference in frequency here cannot be used to argue for a different mechanism.

We agree with the reviewer and the text have been edited to soften our conclusion.

(11) Line 234: 'segregation' might be a better term than 'recycling' here because it is actually the sorting into daughter cells that is the important process.

The text have been edited

(12) Line 235: I don't think this can be what the authors intend to say. If the maternally-inherited rhoptries are not trafficked through the RB (every time), then how do they get into the daughters? Perhaps this is a case where a clear definition of the RB is required.

Our observations indicate that maternally inherited rhoptries are frequently incorporated into daughter cells before collapse of the mother cell and establishment of the residual body (RB). Although F-actin is enriched within the RB, an actin network is also present throughout the parasite cytoplasm, where MyoF is likewise localized. We therefore propose that, unlike micronemes, rhoptries do not necessarily transit through the RB during every replication cycle but can be incorporated directly into developing daughter cells while still relying on the same actin–MyoF-dependent trafficking machinery.

(13) The MyoF Rescue, the experimental plan is not fully described in order to be clear. If the endomembrane architecture was disrupted by MyoF depletion, and this secondary effect caused the segregation phenotype, restoration of MyoF might also simply restore the endomembrane system. So a direct role for MyoF doesn't seem to have been tested in this case.

We appreciate the reviewer's concern that the segregation phenotype could, in principle, arise indirectly from disruption of endomembrane architecture following MyoF depletion. However, although Golgi morphology is altered in MyoF-depleted parasites, its core functions appear largely preserved. This is supported by the normal biogenesis of de novo micronemes, their correct targeting to the apical pole, their efficient secretion, and the previously reported preservation of parasite invasion. In addition, Golgi markers are not detected in the residual body, where maternally inherited micronemes accumulate, arguing against Golgi-mediated trafficking as the primary cause of the segregation phenotype.

Taken together, these observations support the interpretation that the segregation defects are more likely to reflect a direct role of MyoF in organelle trafficking and inheritance than a secondary consequence of generalized disruption of endomembrane organization.

(14) Line 279: Why call it a checkpoint? What is the evidence for its presence here being sensed before a further process is activated, which is what a checkpoint does?

We agree the reviewer that checkpoint is a misleading term and have been updated to trafficking hub.

(15) Line 287 confuses replication of the daughters from cytokinesis, which only happens when each cell loses cytoplasmic connectivity with the other.

We will clarify this point. In Toxoplasma gondii, cytokinesis represents the final step of daughter cell formation, during which the two fully assembled daughter parasites separate from the mother cell following collapse of the maternal cytoplasm. Historically, the residual body was proposed to arise simply as leftover material from this process. However, multiple studies have now shown that residual body formation is an active and regulated process, dependent on specific cytoskeletal and trafficking factors. Importantly, although cytokinesis marks the physical separation of daughter cells from the mother, parasites within a vacuole remain connected via the residual body and continue to share cytoplasmic and plasma membrane components until egress.

(16) Line 298: I don't think there is direct evidence of degradation in the RB. There might be disassembly, but degradation implies proteolysis, which hasn't been tested for.

We agree with the reviewer that our data do not provide direct biochemical evidence of proteolysis within the residual body (RB) and primarily demonstrate disassembly of the maternal IMC at this site. However, several observations are consistent with local degradation. Following IMC collapse, we do not detect Halo-positive signal elsewhere in the parasite, nor do we observe IMC-positive vesicles or other structures that would suggest transport to a distinct degradation compartment.

In addition, previous work identified the E3 ubiquitin ligase CSAR1 as a mediator of protein turnover within the RB, supporting the idea that this compartment is associated with degradation-related processes (O'Shaughnessy et al., 2023). While we cannot formally demonstrate proteolysis, these observations support a model in which IMC disassembly is closely coupled to local degradation within the RB.

(17) The paragraph structure gets a bit confusing at times. See single sentence paragraph, Line 224. Does this sentence justify its own paragraph?

The text has been edited.

(18) Make sure Toxoplasma gondii is in italics throughout.

The text has been edited

(19) Line 279 cites Figure 10. But there is none.

The text has been edited

(20) I advocate introducing a few new acronyms, like DCs. I find that this ultimately reduces the ease with which readers read the work if they don't learn them all quickly.

We agree that excessive use of acronyms can negatively impact readability. In the present manuscript, all abbreviations used in the text are introduced at their first occurrence in the Introduction, including DCs (line 32), IMC (line 41), PV (line 35), ER (lines 38–39), RB (line 50), and IVN (line 49). We have carefully limited the use of abbreviations to commonly used terms in the field and to those that recur frequently throughout the manuscript, with the aim of balancing clarity and readability. Nevertheless, we are happy to reduce or remove specific abbreviations if the reviewer feels this would further improve clarity.

Reviewer #2 (Recommendations for the authors):

(1) Descriptions of methodology and statistical analysis were incomplete as follows:

(1a) It was unclear how the fluorescence intensity measurements (used to evaluate inheritance vs. new synthesis) were carried out. The y-axis on the graph is labeled average fluorescence intensity (% of max intensity). However, it does not state what was averaged (average fluorescence per vacuole?) and what was max intensity (max pixel intensity in each image or time point with the highest average intensity, relative to the other time points?)

We agree with the reviewer that the original description of the fluorescence intensity (FI) measurements lacked clarity. We have therefore revised the Materials and Methods section and added a new supplementary figure (Figure S11) illustrating the analysis workflow.

Briefly, vacuoles were imaged as Z-stacks, and maximum-intensity projections were used for analysis. FI was quantified as the maximum grey value measured within representative regions of interest (ROIs) drawn on individual tachyzoites, rather than as an integrated fluorescence signal across the vacuole. This approach minimizes variability arising from differences in ROI size and allows direct comparison between replication stages.

For each biological replicate, up to 25 tachyzoites per replication stage were analyzed. The mean FI for each stage was normalized to the highest mean value within that replicate, and data from three independent biological replicates were subsequently averaged.

(1b) Given the uncertainties with how these measurements were performed, it is difficult to interpret the data. For example, one would expect that the fluorescence intensity of newly synthesized IMC1 in 8-parasite vacuoles would be 4 times higher than that of a 2-parasite vacuole; however, based on the graph in Figure 1B, the measured increase was only 30%.

We agree that the original description of the fluorescence intensity (FI) measurements required further clarification and have revised the Materials and Methods accordingly. As the reviewer correctly notes, a fourfold increase in FI between 2- and 8-parasite vacuoles would be expected if total IMC fluorescence across the entire vacuole had been measured. However, this was not the parameter quantified.

Instead, FI was measured as the maximum grey value within representative regions of the daughter IMC, providing a per-cell rather than a whole-vacuole measurement. Using this approach, FI increases between the 2- and 4-parasite stages and then reaches a plateau.

This behavior is consistent with the biology of IMC biogenesis. Although the total amount of IMC per vacuole increases with parasite number, the amount of IMC protein incorporated into each daughter parasite remains relatively constant. Consequently, once daughter IMCs are fully assembled from de novo-synthesised material, additional rounds of replication increase the total IMC content per vacuole but not the fluorescence intensity measured for individual parasites.

(1c) T. gondii replicates in an asynchronous manner, so that at the 24-hour time point, a single dish can contain vacuoles containing 2, 4, and 8 parasites. This should be stated explicitly so readers unfamiliar with T. gondii's growth patterns can understand how the experiment was performed.

We agree with the reviewer and have updated the text line 93. “As Toxoplasma gondii replicates in an asynchronous manner, after 24 of replication, vacuoles containing 1, 2, 4, and 8 parasites can be observed in a single dish.”

(1d) Colocalization package in Fiji used for ANKER1-Halo with HDEL-GFP and MIC2/RON2 with CbEmeraldFP should be specified.

We thank the reviewer for this suggestion. Following this recommendation, we performed Pearson correlation analysis for the ANKER1–HDEL-GFP experiment using the Coloc 2 plugins of FiJi. ANKER1Halo and HDEL-GFP showed a strong spatial correlation (Pearson's R = 0.92 ± 0.04, n=30 parasites from three independent biological replicates), supporting localization of ANKER1 to the ER.

We note, however, that this analysis should be interpreted as evidence for co-distribution within the same organelle rather than direct molecular colocalization, as ANKER1 is a transmembrane protein whereas HDEL-GFP labels the ER lumen.

For all the rest of our analysis, no automated colocalization package or plugin in Fiji was used for the analyses involving ANKER1-Halo with HDEL-GFP or MIC2/RON2 with Cb-EmeraldFP. Colocalization was assessed manually across all experiments by inspecting both full Z-stacks and maximum-intensity projections to ensure robust spatial overlap.

For MIC2 and RON2, the presence of signal within the Cb-Emerald–positive filament was scored as either cytoplasmic, on the residual body or absence of colocalisation.

In total, more than 300 and 500 vacuoles were analyzed for MIC2 and RON2–Cb-Emerald colocalization respectively (stable expression of both markers), and more than 150 vacuoles were analyzed for ANKER1-Halo and HDEL-GFP colocalization (transient expression of HDEL-GFP).

This information has now been added to the Methods section.

(1e) Statistical methods should be described on an experiment-by-experiment basis. The authors should justify why a one-tailed t-test was conducted. A two-tailed t-test seems more appropriate.

We agree that statistical methods should be clearly justified on an experiment-by-experiment basis. All the statistical analysis have been performed using two tails and updated in the figures.

(1f) In Figures 5E and 5F, using boxes to indicate the exact areas of the cell that were used in the fluorescence intensity measurements, rather than arrows, would make this data easier to interpret.

The figure has been updated.

Reviewer #3 (Recommendations for the authors):

The current time-lapse images and videos do not clearly demonstrate microneme movement from the maternal parasite apical end to the residual body and back to the apical end of daughter parasites. As such, the route by which micronemes enter daughter parasites remains inconclusive. To strengthen their claims, the authors should employ higher temporal resolution imaging to definitively capture the movement of micronemes from the maternal apical region into the daughters. From the current data, it also seems plausible that the micronemes may be trafficked into the daughters through the conoid as well, as there is no evidence provided showing a movement of micronemes away from the apical end of the maternal parasite before being present in the daughter parasites.

We agree with the reviewer that higher temporal resolution imaging would provide a more definitive view of microneme trafficking. However, long-term live imaging of replicating Toxoplasma gondii requires a compromise between temporal resolution and parasite viability. In our experiments, images were acquired every 15–30 min over periods of up to 16 h, as more frequent acquisition consistently induced phototoxicity and prevented completion of parasite replication.

Despite this limitation, our imaging reliably tracked maternal micronemes over successive rounds of endodyogeny and consistently showed microneme signal associated with the residual body. These observations are in agreement with previous high-temporal-resolution studies, which demonstrated F-actin-dependent microneme trafficking within the residual body over shorter imaging periods (Periz et al., 2019).

We cannot formally exclude the possibility that some micronemes are transferred directly to daughter parasites through the apical end. However, together with previous studies showing enrichment of F-actin at the basal region of developing daughter cells rather than at the apical tip (Periz et al., 2017), our observations support a model in which RB-mediated trafficking contributes to maternal microneme inheritance.

The lack of a clear residual body marker needs to be addressed, as the distinction between the basal end of the maternal cell and a bona fide residual body must be explicitly defined to substantiate the major claim of the study. As it stands, it remains unclear whether micronemes and rhoptries as a whole travel through the residual body to be transported into the daughter parasites.

We agree with the reviewer that a marker specific to the residual body would strengthen this study. Unfortunately, no such marker has been identified to date. The F-actin chromobody is currently the best available proxy, as previous studies have shown that the F-actin network is enriched within the residual body (Periz et al., 2017; Kellermeier et al., 2024). Moreover, high-resolution live-cell imaging has previously demonstrated F-actin-dependent microneme trafficking within this compartment (Periz et al., 2019). We have revised the manuscript to more clearly acknowledge this limitation.

The conclusions drawn from the actin colocalization data in Figure 6C are based entirely on fixed samples, despite all experimental tools being compatible with live-cell imaging. Supplementing the fixed imaging with live cell data would increase its biological relevance. Published studies have shown that fixation of the actin chromobody results in the loss of resolution of an appreciable amount of the cytosolic F-actin network, and while the localizations analyzed here are primarily along the periphery, since the quantification and text make claims about the colocalization within the cytosol, this potential loss of cytosolic F-actin becomes an issue as there may be more actin available for analysis that is lost due to fixation within the cytosol of the parasites.

We thank the reviewer for raising this important point and agree that conventional fixation can compromise preservation of the F-actin network. However, we used the same fixation protocol described by Periz et al. (2019), which allows reliable visualization of the RB-associated F-actin network. We have also corrected the description of the fixation protocol in the Materials and Methods.

Fixation was necessary to image entire vacuoles with sufficient spatial resolution and signal-to-noise ratio for the volumetric analyses presented in Figure 6C. Although some loss of cytosolic F-actin cannot be excluded, this would be expected to reduce, rather than artificially increase, the detection of organelle–actin associations.

Importantly, previous live-cell imaging studies demonstrated F-actin-dependent microneme trafficking (Periz et al., 2019), and our observations are consistent with these findings. Moreover, the defects observed following MyoF depletion provide independent functional evidence that the trafficking events described here rely on the actin–MyoF transport machinery.

The statement of colocalization should be backed up by quantitative coefficients like Pearson's coefficient.

We thank the reviewer for this helpful suggestion. Following this recommendation, we performed a Pearson correlation analysis of ANKER1-Halo and HDEL-GFP using the Coloc 2 plugin in Fiji. ANKER1-Halo showed a strong spatial correlation with HDEL-GFP (Pearson's R = 0.92 ± 0.04, n = 30 parasites from three independent biological replicates), supporting localization of ANKER1 to the ER. As ANKER1 is a transmembrane protein and HDEL-GFP labels the ER lumen, this analysis should be interpreted as evidence of co-distribution within the same organelle rather than direct molecular colocalization.

In contrast, we do not consider Pearson's coefficient appropriate for evaluating the association of micronemes or rhoptries with F-actin. These organelles are predominantly concentrated at the apical pole and, when associated with F-actin, are typically positioned along rather than directly overlapping the filaments. Consequently, Pearson's coefficient would underestimate these biologically relevant associations. We therefore relied on morphological and spatial criteria, which we consider more appropriate for assessing organelle–cytoskeleton interactions.

In addition, from the methods and presented figure images, specifically in Figure 6C for RON2, how the cytosolic and residual body actin is separated is difficult to discern, as there is a clear residual body actin signal overlapping a parasite. The methods for how this was separated and analyzed should be clearer to remove doubts about how this area was measured, as the current description raises concerns about the counting of the residual body actin within the cytosol.

We agree with the reviewer that the distinction between cytosolic and residual body (RB)-associated F-actin required further clarification. All analyses were performed manually, as described in our response to Reviewer 2 (comment 1d). The RB was identified by the presence of thick, bundled F-actin filaments at the basal pole that formed a continuous structure connecting parasites within the vacuole, whereas cytosolic F-actin was defined as the thinner filamentous network within the parasite body.

No automated or threshold-based segmentation was used because the marked differences in filament morphology and fluorescence intensity make reliable thresholding difficult and prone to misclassification. Manual annotation based on spatial localization and filament morphology was therefore considered the most appropriate approach. We have clarified these criteria in the Materials and Methods section.

Line comments:

(1) 113: round to rounds - "did not obtain maternal organelles after successive rounds of replication...".

The text has been updated

(2) 146: grammatical, "As consequence a progressive" -> "As a consequence", or "Consequently".

The text has been updated

(3) 164-166: "Autonomous duplication" implies the separation and duplication of the Golgi occurs on its own, i.e., without any outside intervention, when we know from Carmeille et al. 2021 and Figure 7C here that the Golgi becomes fragmented over rounds of division in the absence of MyoF. I think this is primarily a word choice error with "autonomous".

We agree with the reviewer and the word autonomous has been removed

(4) 184: The wording suggests that DC's refers to daughter IMC's, when DC has already been given as an abbreviation for daughter cells previously.

The text has been updated to correct this error

(5) 189: de novo is not italicized.

The text has been updated

(6) 192: The data shown so far do not show that the RB plays a selective role in organelle recycling.

The text has been edited to fit better our results “Our data suggest that the organelles trafficking through the residual body (RB) have different fate”

(7) 208: State that it depends on F-actin, but never show that it is dependent on F-actin through actin disruption, such as cytochalasin D treatment or a specific conditional disruption of F-actin.

We agree with the reviewer that we did not repeat F-actin disruption experiments (e.g., cytochalasin D or jasplakinolide treatments) in this study. These experiments were performed in our previous work, where pharmacological disruption of F-actin was shown to impair microneme trafficking (Periz et al., 2019). We therefore chose not to repeat these assays.

Instead, the present study provides complementary evidence by demonstrating that depletion of Myosin F (MyoF), a motor that uses F-actin as a transport track (Kellermeier et al., 2024), disrupts the trafficking of both maternal micronemes and rhoptries. Together, our previous F-actin perturbation experiments and the MyoF depletion data presented here support the interpretation that these trafficking events depend on the actin–MyoF transport machinery.

(8) 209: This suggests that the chromobody was transiently expressed in the RON2-Halo line, but the methods suggest MIC2-Halo and RON2-Halo were integrated into a parasite line stably expressing Cb-Emerald.

The text has been edited.

(9) 229: The section is confusing with the mention of (now maternal). If I understand correctly, the point being made is that the de novo synthesized MIC2 at stage 2 is now the maternal MIC2 for stage 4, but coloring-wise within the figure, the now maternal MIC2 at stage 4 from stage 2 would still be green. The methods suggest these images were all taken simultaneously, and not at specific timepoints of the same vacuole, so the now maternal line remains confusing.

The text has been revised for clarity and now reads: “Because Toxoplasma gondii replicates asynchronously, vacuoles at different replication stages coexist within the same culture after 24 h. The second labeling step marks all proteins synthesized since the beginning of the experiment, allowing discrimination between proteins present in the original mother parasite and those synthesized during subsequent replication cycles. As daughter parasites form, they inherit material from their mother, such that proteins synthesized during one replication cycle become maternal proteins in the next. Under MyoF depletion, these newly synthesized protein pools accumulate within the residual body instead of being redistributed to daughter parasites during subsequent rounds of replication (Figure 7, stage 4).”

(10) 238: The sentences here indicate that Golgi inheritance occurs without issue: "golgi inheritance remained unaffected by MyoF depletion". But, it is evident from the images shown that the Golgi is extremely fragmented, with many more Golgi fragments by stage 8 than there are parasites. This could be solved by rewording and including a line along the lines of "in accordance with the results found in Carmeille et al. 2021".

The reference to this article is already stated later in the text now line 299-303 “This active role is further supported by the dependence of RB-mediated recycling on F-actin and the class XXII myosin MyoF, which we show to be essential for retrieval of maternal MIC2 and RON2 but dispensable for Golgi inheritance although we noticed a fragmentation of the Golgi, which has been described to depend on MyoF (Carmeille et al., 2021).”

(11) 319: missing a comma, "Many of these, particularly...".

The text has been edited.

(12) 437: Images -> imaged.

The text has been edited.

(13) 439: a fresh media -> and fresh media.

The text has been edited.

(14) 439: Replication -> replicate.

The text has been edited.

(15) 440: images -> imaged.

The text has been edited.

Other grammatical issues within the methods:

(1) Figure 5C: Within the y-axis label of Figure 5C, there is an asterisk with no asterisk explanation within the legend.

We thank the reviewer for pointing out this oversight. The figure legend has been updated to include an explanation of the asterisk, providing a clearer understanding of the results.

(2) Figure 5E: The addition of an IMC1 label to match the magenta color would be helpful to readers.

The figure has been updated

(3) Figure 6D: No statistics showing significance of colocalization.

We thank the reviewer for this comment. In Figure 6D, we report the frequency of observed colocalization between organelles (MIC2 and RON2) and F-actin filaments, based on manual analysis across >300 vacuoles for MIC2 and >500 vacuoles for RON2. Specifically, we observed cytoplasmic F-actin colocalization in 94% of vacuoles for MIC2 and 79% for RON2, and residual body F-actin colocalization in 85% of vacuoles for MIC2 and 11% for RON2. This analysis is descriptive and is not intended to compare MIC2 versus RON2 quantitatively; rather, it illustrates the general association of these organelles with F-actin filaments. Standard statistical measures such as Pearson correlation are not meaningful in this context because the organelles are punctate and primarily localized along filaments rather than overlapping continuously. Importantly, the percentages reported represent the fraction of vacuoles in which colocalization can be observed, not the percentage of colocalization between Cb-Emerald and MIC2/RON2 within individual vacuoles.

(4) Figure 7: The legend title of Figure 7 is at the end of the legend of Figure 6.

The text has been edited.

Version 2

Abstract

Toxoplasma gondii replicates through endodyogeny, an unconventional form of internal budding in which two daughter cells are assembled within a single mother cell. During this process, daughter cells must acquire a full complement of organelles, which may be inherited from the mother, formed de novo, or assembled through a combination of both mechanisms. To date the fate of maternal components during replication remains poorly understood. We previously showed that F-actin–driven dynamics generate the intravacuolar network, which defines the residual body (RB) and facilitates recycling of microneme proteins. However, the inheritance and recycling of other organelles have not been systematically analysed. To address this, we employed a dual HaloTag-based pulse-chase fluorescence labelling strategy to distinguish between de novo–synthesized and recycled proteins in replicating tachyzoites. This approach reveals three distinct organelle inheritance patterns: (1) direct transmission of intact maternal organelles (e.g., rhoptries, micronemes), (2) expansion and division of pre-existing maternal organelles with incorporation of newly synthesized components (e.g., Golgi apparatus, apicoplast), and (3) degradation of maternal structures without recycling (e.g., inner membrane complex). Furthermore, we identify Myosin F (MyoF) as the key motor protein that mediates the selective recycling of maternal organelles via the RB. These findings support a model in which the RB functions as an active trafficking hub and reveal a selective, regulated system of organelle inheritance and recycling that is critical for intracellular organization and parasite development.

Introduction

Toxoplasma gondii tachyzoites replicate through a specialized form of internal budding known as endodyogeny, wherein two daughter cells (DCs) are assembled within the cytoplasm of the mother cell (Francia and Striepen, 2014; Hu et al., 2002). This unique replication mechanism demands precise orchestration of organelle inheritance and spatial organization within the parasitophorous vacuole (PV). Apicomplexan parasites, including T. gondii, exhibit a highly polarized cellular architecture defined by specialized secretory organelles—micronemes, rhoptries, and dense granules—essential for host cell invasion and intracellular survival. These are accompanied by canonical organelles such as the Golgi apparatus, endoplasmic reticulum (ER), a single mitochondrion, and the apicoplast, a non-photosynthetic plastid. Notably, unlike most eukaryotes, T. gondii is enveloped by a pellicle consisting of the plasma membrane and an underlying inner membrane complex (IMC) (Ouologuem and Roos, 2014), a structure composed of flattened vesicles tightly associated with the subpellicular cytoskeleton (Harding and Meissner, 2014).

As DCs emerge during endodyogeny, they progressively encapsulate nearly all maternal cytoplasmic contents. While the sequence of organelle acquisition has been extensively characterized (Nishi et al., 2008), the extent to which maternal organelles and proteins are recycled remains unclear. We previously showed that maternal micronemal proteins, such as MIC2, are almost entirely recycled into daughter cells via an F-actin–dependent mechanism and are trafficked through the intravacuolar network (IVN), which structurally organizes the residual body (RB) (Periz et al., 2019). Historically viewed as a passive remnant of maternal cytoplasm, the RB has emerged as a key structure in coordinating replication, facilitating organelle recycling, and maintaining cytoplasmic continuity across individual parasites (Periz et al., 2017b). The RB forms at the posterior end of dividing parasites, physically linking them through the actin-rich IVN and enabling cytoplasmic exchange and synchronized development within the PV (Periz et al., 2017b; Tosetti et al., 2019). The RB originates from the collapse of the maternal parasite during daughter cell budding and occupies the space previously occupied by the mother cell. As a highly dynamic structure, its morphology and organization change throughout parasite replication. At the end of the first replication cycle, the newly formed RB is initially visible as a thin F-actin-rich connection between newly formed daughter parasites and subsequently becomes more prominent as recycled organelles and other cellular material traffic through this compartment. Although no RB-specific molecular marker has yet been identified, the RB is consistently characterized by a dense F-actin network connecting parasites within the parasitophorous vacuole (Periz et al., 2017b). Recent studies have shed some light on the molecular architecture and regulation of the IVN. We demonstrated that F-actin filaments organize the IVN, with disruption of actin (ACT1) resulting in parasite disorganization and impaired microneme recycling (Periz et al., 2019; Periz et al., 2017b). The RB and IVN also require the function of unconventional myosins MyoI and MyoJ, which are implicated in cell-cell communication (Frenal et al., 2017). Additionally, the actin-nucleating factors Formin-2 and Formin-3 are required for IVN formation (Stortz et al., 2019; Tosetti et al., 2019). However, the mechanisms driving the physical transport of maternal organelles remain poorly defined. F-actin could mediate this process either by serving as a track for myosin-dependent vesicular transport, as described for myosin VI in other eukaryotes (Frank et al., 2004), or through the intrinsic mobility and dynamic association of actin bundles themselves, which could transiently interact to promote vesicle exchange (Das et al., 2021; Khaitlina, 2014; Moore et al., 2021).

Although the biogenesis of daughter cell structures has been well described, the fate of maternal organelles during replication remains incompletely understood. Evidence suggests that maternal components such as MIC2 and the IMC protein GAP40 can be recycled via actin-dependent mechanisms (Ouologuem and Roos, 2014; Periz et al., 2019). However, a systematic investigation of organelle recycling during endodyogeny has yet to be performed.

In this study, we employ a dual-labeling strategy using the HaloTag system (Urh and Rosenberg, 2012) to investigate the fate of maternal organelles during T. gondii replication. Through fluorescence intensity-based analysis, we identify three distinct fates for maternal proteins: full inheritance of intact organelles, incorporation of newly synthesized components into expanded maternal structures, and degradation without recycling. Furthermore, we demonstrate that the residual body functions as a key recycling hub, with this process dependent on the activity of the class XXII myosin F (MyoF). These findings provide new mechanistic insights into organelle inheritance and reveal how intracellular organization is maintained during parasite replication.

Results

HaloTag-Based pulse chase labelling allows discrimination of recycled and de novo material. To evaluate HaloTag as a tool for distinguishing maternal from newly synthesized proteins during endodyogeny, two marker proteins with known fates were used: MIC2, which undergoes extensive recycling (Periz et al., 2019), and IMC1, primarily synthesized de novo (Ouologuem and Roos, 2014). Using CRISPR/Cas9, endogenous loci were tagged to generate MIC2-Halo and IMC1-Halo strains (Singer et al., 2023).

Both strains were sequentially labeled by incubation with membrane permeable dyes for 1h: maternal proteins were marked with Halo JF646 before invasion (M-MIC2, M-IMC1), and 24 hours later after replication, by Halo JF549 for 1h to label newly synthesized proteins (n-MIC2, n-IMC1) (Figure 1A). As Toxoplasma gondii replicates in an asynchronous manner, after 24 hours of replication, vacuoles containing 1, 2, 4, and 8 parasites can be observed in a single dish.

Discrimination between maternal and de novo material.

A) Dual staining scheme. Maternal material is labelled initially with Janila Fluor®-646, and the excess washed out. After replication of parasites de novo synthesised material is labelled with Janila Fluor®-549, allowing efficient discrimination of the two populations. B) Fluorescence intensity quantification of maternal (M) and de novo sythenetised (n) MIC2 and IMC1 during replication from stage 1 to 8. Magenta: M-MIC2, Rosa: M-IMC1, Green: n-MIC2, Dark Green: n-IMC1. C) Representative picture of MIC2-Halo parasites during replication (stage 1 to 8) stained for both maternal (M) and de-novo (n) MIC2. Magenta: M-MIC2, Green: n-MIC2. Maternal MIC2 is efficiently recycled into the daughters, while de novo MIC2 is formed after each replication cycle D) Representative picture of IMC1-Halo parasites during replication (stage 1 to 8) stained for both maternal (M) and de-novo (n) IMC1. Magenta: M-IMC1, Green: n-IMC1. In contrast to micronemes, maternal IMC is degraded, and only residual amount is detected after the first replication cycle. Three biological replicates were used for all analysis with a total of 300 IMC-1 Halo vacuoles and 260 MIC2-Halo vacuoles analysed; error bars are standard deviations, and the centre measurement of the graph bars is the mean. Panels C and D show maximum-intensity projections of Z-stack images. All scale bars = 1 µm.

Fluorescence intensity (FI) analysis across those replication stages (Figure 1B–D, S1) revealed a progressive decline of M-MIC2 and an increase in n-MIC2, confirming protein recycling. In contrast, M-IMC1 FI dropped from 100% to 14% between stages 1 and 2, with a corresponding rise in n-IMC1 to ∼80%, indicating predominant de novo synthesis. These distinct FI dynamics demonstrate HaloTag’s utility for resolving protein origin and support the de novo assembly of the IMC.

HaloTag profiling reveals three distinct patterns of protein inheritance

To further investigate protein inheritance during endodyogeny, endogenous HaloTag fusions were generated for a set of organelle-associated proteins: RON2 and ROP1 (rhoptries) (Besteiro et al., 2009; Striepen et al., 2001), GRA1 (dense granules) (Carruthers and Sibley, 1997), TIC20 (apicoplast) (Carruthers and Sibley, 1997), SortLR (Golgi) (Sloves et al., 2012), ANKER1 (Barylyuk et al., 2020), GAPM1a (IMC) (Harding et al., 2019), and MyoA (glideosome) (Meissner et al., 2002) (Figure 2A, Figure S2).

Three recycling fates for the maternal organelles.

A) Overview of T. gondii organelles and molecular markers used for visualisation with endogenous Halo-Tags. B) Fluorescence intensity (FI) quantification of maternal (M) molecular marker listed in A. The fluorescence intensity variation during replication, allows to group them into three groups: 1) Efficient, almost quantitative recycling. 2) even distribution of maternal material and 3) almost exclusive de novo synthesis. Dark blue: RON2, Blue: ROP1, Pale blue: MIC2, Dark green: SortLR, Green: MyoA, Pale green: Tic20, Greenish white: ANKER1, Gold: GAPM1a, Yellow: IMC1. C) Representative picture of parasites expressing indicated Halo tagged proteins (stage 1 to 8) stained for maternal (M) proteins. Three biological replicates were used for all analysis with an average of 265 vacuoles analysed per protein; the graph indicate the mean value of FI calculated from the triplicate. Panels C shows maximum-intensity projections of Z-stack images. All scale bars = 1 µm.

Following the same sequential labeling strategy used for MIC2 and IMC1, fluorescence intensity (FI) of maternal proteins was quantified across the different replication stages (Figure 2B, C, S2C-G). Analysis revealed three distinct inheritance profiles:

Group 1: RON2 and ROP1 exhibited stable maternal FI, with minimal decline of FI within a single rhoptry, indicating recycling of whole organelles. In good agreement with this hypothesis, detailed analysis of M-RON2 inheritance demonstrated that the maternal rhoptries are separated during the distribution and that some daughter cells did not obtain maternal organelles after successive rounds of replication (Figure 3D).

Group 2: Proteins from diverse organelles (SortLR, ANKER1, TIC20, MyoA) showed a stepwise ∼50% reduction in maternal FI per division, suggesting that recycled, maternal material and de novo material end up in the same organelle

Group 3: IMC proteins (IMC1 and GAPM1a) displayed a sharp FI loss between the first and second replication cycles, consistent with predominant de novo synthesis of the IMC and degradation of maternal material.

Micronemes and rhoptries are recycled as whole organelles

Representative picture of MIC2-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) MIC2. Magenta: M-MIC2, Green: n-MIC2. Zoom windows allow to visualise isolated apical M-MIC2 signal for which the fluorescence intensity is quantified in C. B) Quantification of the inheritance of M-MIC2 by the daughter cells during replication (stage 1 to 8). A total of 260 vacuoles were analysed. All daughter cells inherited M-MIC2. C) Fluorescence intensity quantification of isolated M-MIC2 signal as illustrated in A. A total of 203 isolated M-MIC2 signals were analysed. D) Representative picture of RON2-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) MIC2. Magenta: M-RON2, Green: n-RON2. Zoom windows allow to visualise isolated apical M-RON2 signal for which the fluorescence intensity is quantified in F. E) Quantification of the inheritance of M-RON2 by the daughter cells during replication (stage 1 to 8). A total of 289 vacuoles were analysed. Not all daughter cells inherit M-RON2 (asterisks), best seen in later replication stages. F) Fluorescence intensity quantification of isolated M-RON2 signal as illustrated in D. A total of 215 isolated M-RON2 signals were analysed. Three biological replicates were used for all analysis; error bars are standard deviations, and the centre measurement of the graph bars is the mean. All p-values ≤ 0.001 (***), using two-tailed unpaired Student’s t-test. Panels A and D show maximum-intensity projections of Z-stack images. All scale bars = 1 µm.

Interestingly, MIC2 exhibited intermediate behavior, prompting further analysis to refine its classification. These patterns highlight distinct organelle-specific protein inheritance mechanisms in T. gondii.

Micronemes and rhoptries are recycled as intact organelles during Toxoplasma endodyogeny

Micronemes are known to comprise distinct subpopulations (Kremer et al., 2013), and prior work has shown that recycled and de novo MIC2 localize to separate microneme subsets (Periz et al., 2019). Based on these observations, we hypothesized that both micronemes and rhoptries are recycled as intact organelles, with new organelles formed independently via de novo synthesis.

To test this, we conducted detailed analyses of MIC2 and rhoptry inheritance. As expected, recycled (M-MIC2) and newly synthesized MIC2 (n-MIC2) were largely segregated into distinct micronemes with minimal colocalization (Figure 3A, S3)(Periz et al., 2019). Focusing on sparsely distributed micronemes outside the apical region (Figure 3A, boxed area), we measured fluorescence intensity (FI) of individual M-MIC2-positive micronemes at replication stages 2, 4, and 8 (Figure 3A, zoomed region; 3C). The relatively stable FI of these isolated structures supports whole organelle recycling, consistent with the FI profiles of rhoptries (Figure 2B, C).

Rhoptry analysis further corroborated this model. As with MIC2, M-RON2 and n-RON2 localized to distinct rhoptry populations (Figure S3), with divergence becoming more apparent at stage 8 (Figure 3D, Figure S4A). M-RON2 FI remained relatively constant through replication (Figure 3 F), while n-RON2 gradually increased from stage 2 onward (Figure S4B). Due to their low copy number (8–12 per tachyzoite), maternal rhoptries were separated and often absent in some daughter cells after successive divisions, consistent with stochastic whole-organelle inheritance illustrated by decrease in the average area of the M-RON2 fluorescence signal and the constancy of the total surface area (Figure S4C, D). As a consequence, a progressive decline in M-RON2-positive parasites per vacuole can be observed (Figure 3E, D; Figure S4A).

For the micronemes, even in larger parasitophorous vacuoles, all daughter parasites retained a mix of recycled and newly synthesized micronemes (Figure S5). Interestingly, the maternal micronemes were observed to be relatively well distributed among the daughter cells (Figure S5A, B). Considering that a single mother parasite contains approximately 64 micronemes and that successive rounds of endodyogeny generate up to 32 daughter cells, a purely random segregation of existing micronemes would be unlikely to produce the relatively uniform distribution observed (∼0–3 maternal micronemes per tachyzoite). Simple probabilistic modelling indicates that the observed distribution would occur in less than 1% of vacuoles under a random inheritance scenario. Yet similar distributions were consistently observed across multiple vacuoles and independent biological replicates (Figure S5C, D). These observations strongly argue against stochastic segregation and instead support the existence of a regulated mechanism that promotes balanced microneme inheritance between daughter parasites. Together, these results demonstrate that both micronemes and rhoptries are recycled as intact organelles. De novo organellogenesis acts in parallel to maintain a full complement in each daughter cell, ensuring faithful inheritance during endodyogeny.

Golgi inheritance involves coordinated expansion and partitioning of maternal and de novo proteins

Next, we analyzed SortLR, a Golgi marker, using dual HaloTag labeling. Maternal SortLR (M-SortLR) fluorescence declined by ∼50% per replication round, while newly synthesized SortLR (n-SortLR) increased proportionally, without abrupt loss (Figure 4A–B). This gradual dilution contrasts sharply with the rapid decay seen in Group 3 proteins.

Expansion of the mother organelle for equivalent sharing to the daughter.

A) Representative picture of SortLR-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) SortLR. Magenta: M-SortLR, Green: n-SortLR. B) Fluorescence intensity quantification of M-SortLR and n-SortLR illustrated in A. Magenta: M-SortLR, Green: n-SortLR. A total of 251 vacuoles were analysed. C) Quantification of the inheritance of M-SortLR by the daughter cells during replication (stage 1 to 8). All daughter cells inherited M-SortLR. A total of 100 vacuoles were analysed. D) Quantification of the signal area of M-SortLR during replication. A total of 100 vacuoles were analysed. The total surface of M-SortLR increase at each replication by about a factor 2. Three biological replicates were used for all analysis; error bars are standard deviations, and the centre measurement of the graph bars is the mean. All p-values ≤ 0.001 (***), using two-tailed unpaired Student’s t-test. Panels A shows maximum-intensity projections of Z-stack images. All scale bars = 1 µm.

M-SortLR and n-SortLR exhibited strong colocalization across replication stages, and all daughter cells retained M-SortLR (Figure 4C). The total Golgi area expanded from ∼2 µm² to ∼10 µm² by stage 8 (Figure 4D), consistent with coordinated Golgi elongation and medial fission, as described previously (Pelletier et al., 2002). These results demonstrate that the Golgi undergoes duplication, incorporating both maternal and de novo proteins during its growth and partitioning. Similar signal evolution was observed for the ER, MyoA and apicoplast (Figure S2, S6). Thus, Group 2 organelles follow a distinct inheritance mode, involving regulated expansion, integration of new material, and equal distribution to progeny—distinct from the whole-organelle recycling of rhoptries/micronemes and the complete turnover seen in IMC proteins.

Maternal IMC proteins are degraded in the RB during replication

Group 3 proteins, typified by IMC components like GAPM1a and IMC1, exhibit a sharply distinct inheritance profile, characterized by rapid loss of maternal fluorescence during replication. Using HaloTag-based dual labeling, we observed a ∼90% reduction in maternal GAPM1a signal by stage 2, with minimal retention through later stages (Figure 5A–C), indicative of active degradation rather than recycling. M-GAPM1a and n-GAPM1a did not colocalize, and residual maternal signal localized transiently at the posterior pole before disappearing entirely, suggesting disposal via the residual body (RB) (Figure 5D).

Degradation of the inner membrane complex.

Representative picture of GAPM1a-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) GAPM1a. B) Fluorescence intensity quantification of M- GAPM1a and n-GAPM1a illustrated in A. A total of 265 vacuoles were analysed. C) Quantification of the inheritance of M- GAPM1a by the daughter cells during replication (stage 1 to 8). After replication, M-GAPM1a is not visible in daughter cells. A total of 265 vacuoles were analysed. *The calculation was performed using single stage parasite intensity to set up the exposure as performed for all the other organelles. Under those conditions the remaining signal of the maternal GAPM1a is below the detection level and cannot be observed. D) GAPM1a is degraded in the RB. Representative images of GAPM1a-Halo parasites at different stages of daughter cell development. M-GAPM1a collapses toward the forming residual body, where the signal disappears after completion of replication, indicating it’s degradation. E) Time series of IMC1-Halo parasites during the first replication. Parasite were stained for the maternal (M) IMC1 prior invasion. Three regions of the parasites were analysed: the apical (green ROI), the cytoplasmic (blue ROI) and the basal (magenta ROI). F) Fluorescence intensity analysis of the three regions defined in E. The curve and analysed ROI share the same colour code. Green: Apical, Blue: Cytoplasm, Magenta: Basal. After the accumulation of the mother IMC at the basal pole, the FI of M-IMC1 decrease without redistributing to any other region. Three biological replicates were used for all analysis; error bars are standard deviations, and the centre measurement of the graph bars is the mean. All p-values ≤ 0.001 (***), using two-tailed unpaired Student’s t-test. Panels A and D show maximum-intensity projections of Z-stack images. Panel E displays single optical plane images acquired during live imaging. All scale bars = 1 µm.

Live-cell imaging of IMC1 further supported this model: maternal IMC1 (M-IMC1) condensed at the posterior pole during daughter budding, followed by a marked fluorescence drop, without redistribution to other regions (Figure 5E, F and Figure S7, Video S1). This signal loss was replication-specific, as non-dividing parasites retained stable M-IMC1 levels (Figure S7D). Quantification across regions revealed a transient increase at the posterior, followed by complete disappearance, consistent with degradation (Figure 5F, Figure S7C).

Interestingly, DCs IMCs exhibited brighter fluorescence than the maternal IMC (Figure S7E, F), especially during early stages of elongation (Figure S7G, H). FI peaked when DCs reached ∼3 μm in length, suggesting the IMC is fully assembled before emergence and subsequently unfolds without requiring additional material. This preformation explains the absence of maternal IMC recycling. These findings align with earlier observations that IMC components are synthesized de novo during elongation (Ouologuem and Roos, 2014).

Whole-organelle inheritance of micronemes and rhoptries occurs via similar recycling pathways

Our data suggest that the organelles trafficking through the residual body (RB) have different fate: while microneme and rhoptry proteins are routed through the RB, IMC proteins are primarily degraded within this compartment. In contrast, recycling of maternal Golgi, ER, and apicoplast proteins appears to occur directly, bypassing the RB.

To investigate this further, we performed live-cell imaging of rhoptry and microneme inheritance (Figure 6A, B, Video S2, S3). MIC2-Halo parasites co-expressing IMC1-YFP, showed that maternal MIC2 (M-MIC2) transits through the RB (Figure 6A, white arrows). Notably, after endodyogeny was completed, M-MIC2 was redistributed from the RB to the apical tip of the daughter cells (Figure 6A, 11:30, 16:00 h), confirming that the RB serves as a temporary reservoir during microneme recycling (Periz et al., 2019).

Micronemes and rhoptries are recycled similarly, but differ in timing and location.

A) Time-lapse of MIC2-Halo over two replication cycles. Maternal MIC2 (M-MIC2) is stained, and daughter cell formation is visualized via IMC1-YFP. M-MIC2 is transported via the residual body (RB) (white arrows; see Video S2). B) Time-lapse of RON2-Halo parasites with IMC-YFP. Maternal RON2 (M-RON2) is transported before mother cell collapse and RB formation (white arrows). C) Representative images showing maternal organelles associating with F-actin and the RB. Insets highlight colocalization of maternal proteins and F-actin. D) Quantification of M-MIC2 and M-RON2 colocalization with F-actin. Data from three biological replicates, a total of 534 RON2-Halo and 342 MIC2-Halo vacuoles were analysed; bars represent means ± SD. Panel A and B display single optical plane images acquired during live imaging. Panel C show single Z plane of Z-stack images. Scale bars = 1 µm.

For rhoptries, we examined the inheritance of maternal RON2 (M-RON2) in RON2-Halo parasites co-expressing IMC1-YFP. In approximately 90% of parasites undergoing replication, M-RON2 was integrated into daughter rhoptries prior to mother cell collapse and formation of the RB (Figure 6B, 4:15–4:30 h and 10:30–10:45 h). Like M-MIC2, M-RON2 was occasionally detected in the RB, though less prominently, suggesting more rapid, tightly regulated or easier recycling due to their lower number.

To test whether recycled rhoptries associate with F-actin, as previously reported for micronemes (Periz et al., 2019), we stably integrated the F-actin marker Cb-Emerald (Periz et al., 2017b) in the RON2-Halo line (Figure 6C, D). While m-MIC2 displayed clear F-actin–associated movement along filaments in the residual body (Figure 6C), colocalization of M-RON2 with Cb-Emerald in the residual body was infrequent. On the other hand, both secretory organelles were found to be associated with cytoplasmic F-actin (Figure 6D).

In summary, both rhoptries and micronemes are inherited as intact organelles during T. gondii replication. These findings highlight the organelle-specific complexity of recycling pathways and support whole-organelle segregation of micronemes and rhoptries into the daughter cells. Given the F-actin dependency of this recycling pathway demonstrated previously for the micronemes (Periz et al., 2019), we hypothesized that a conserved apicomplexan myosin mediates this process. MyoF, a class XXII myosin is conserved across Apicomplexa (Jacot et al., 2013) and has been implicated in diverse functions—from apicoplast inheritance to F-actin–dependent vesicular transport (Carmeille et al., 2021; Heaslip et al., 2016; Kellermeier and Heaslip, 2024).

We endogenously Halo-tagged MIC2, RON2, and SortLR in the auxin-inducible degron line MyoF-mAID (Carmeille et al., 2021) and performed dual HaloTag labelling to differentiate maternal (M) from de novo (n) protein pools. Live imaging revealed that MyoF depletion resulted in the accumulation of maternal but not de novo MIC2 and RON2 in the RB during early replication stages (Figure 7A, B). As Toxoplasma replicates asynchronously, different replication stages coexist within the same dish after 24 h. The second labeling step marks all proteins synthesized since the beginning of the experiment, allowing discrimination between proteins present in the initial mother parasite and those synthesized during subsequent replication cycles. Yet, during each round of replication, daughter cells form within and inherit material from their mother cell. As replication proceeds, proteins synthesized during earlier cycles are progressively inherited and recycled in subsequent stages. Under MyoF depletion, this results in the accumulation of green-labeled protein pools, initially synthesized de novo during the previous cycle, within the residual body, rather than their redistribution to daughter cells starting from the second round of replication (Figure 7 Stage 4).Immunofluorescence with α-AMA1, α-MIC4, and α-MIC8 confirmed that this phenotype broadly affects the entire repertoire of micronemal proteins (Figure S8, S9). Quantification showed that approximately 90% of vacuoles exhibited RB accumulation of maternal material upon auxin treatment (Figure 7D), indicating that MyoF specifically mediates the recycling of maternal micronemes and rhoptries. Interestingly, although our time lapse analysis indicated that rhoptries only rarely traffic via the RB, upon depletion of MyoF, they accumulate in the RB, supporting the hypothesis that micronemes and rhoptries are transported in the same manner during recycling. In contrast, Golgi inheritance (marked by SortLR) remained unaffected by MyoF depletion (Figure 7C). Neither accumulation in the RB nor separation of maternal and de novo SortLR signals was observed, consistent with previous reports that Golgi duplicates and partitions independently (Pelletier et al., 2002). Yet, similar fragmentation of the Golgi as previously reported was observed.

Myosin-F drives F-actin mediated recycling of maternal organelles via the residual body.

A) Effect of MyoF knockdown (KD) on maternal microneme inheritance. MyoF-mAID MIC2-Halo parasites were labeled and grown ± auxin for 24h. Magenta: maternal MIC2 (M-MIC2), Green: newly synthesized MIC2 (n-MIC2). Without MyoF, M-MIC2 accumulates in the residual body (RB) instead of being passed to daughter cells (Stage 1–2), with increasing accumulation over replication cycles (Stage 2–8). B) Effect of MyoF KD on maternal rhoptry inheritance. MyoF-mAID RON2-Halo parasites show M-RON2 retention in the RB, mirroring the pattern seen with MIC2. C) Effect of MyoF KD on maternal Golgi inheritance. MyoF-mAID SortLR-Halo parasites show no significant RB accumulation of M-SortLR, unlike MIC2 and RON2. D) Quantification of vacuoles showing RB accumulation. White: control; Gray: MyoF-KD. Three biological replicates were used; a total of 675 Ctrl and 743 KD vacuoles were analyzed for MIC2-Halo, 645 Ctrl and 594 KD for RON2-Halo and 795 Ctrl and 619 KD for SortLR-Halo, bars show means ± SD. All p-values ≤ 0.001 (***), using two-tailed unpaired Student’s t-test. Panels A, B and C show maximum-intensity projections of Z-stack images. Scale bars = 1 µm

To confirm that this phenotype results directly from MyoF depletion and not broader disruption of endomembrane architecture, we performed a rescue experiment. MIC2-Halo MyoF-mAID parasites were cultured in auxin for 24 hours to induce M-MIC2 retention in the RB, followed by auxin chase and continued replication (Figure 8A). Following auxin washout, M-MIC2 was redistributedto daughter parasites during subsequent replication cycles, although occasionally in an uneven manner (Figures 8A, 8C, and S10A). To determine whether this redistribution depended on restoration of MyoF expression, we monitored MyoF recovery by IFA using the HA epitope fused to the MyoF-mAID protein (Figure 8B). Following auxin washout, MyoF expression was restored. Quantification revealed a strong association between the presence of MyoF and the redistribution of maternal micronemes, supporting the conclusion that microneme recycling from the RB is dependent on MyoF function (Figure S10B).

The residual body (RB) functions as a recycling center, not a dead end.

To assess the fate of material accumulated in the RB, auxin chase experiments were performed: 24h with auxin followed by 24h without. A) Representative images of MyoF-mAID MIC2-Halo parasites after 48h in control (no auxin), continuous auxin (Aux), or auxin washout (Aux washed). Magenta: M-MIC2, Green: n-MIC2. Continuous auxin led to M-MIC2 accumulation in the RB, while auxin washout enabled redistribution. B) Representative images of MyoF-mAID MIC2-Halo parasites after 48h in control (no auxin), continuous auxin (Aux), or auxin washout (Aux washed). Magenta: M-MIC2, Green: α-HA (MyoF). In absence of auxin MyoF is visible, induction with auxin deplete MyoF led to M-MIC2 accumulation in the RB, while auxin washout enabled MyoF production and M-MIC2 redistribution. C) Quantification of vacuoles showing normal, accumulated, or redistributed M-MIC2. White: control; Gray: MyoF-KD; Blue: MyoF-KD auxin chase. D) Time-lapse without auxin: M-MIC2 briefly passes through the RB (white arrows). A total of 170 Ctrl, 186 KD and 179 auxin chased vacuoles were analyzed. E) Time-lapse with auxin: sustained M-MIC2 accumulation in the RB throughout replication. F) Time-lapse after auxin washout: initial M-MIC2 accumulation in the RB resolves by the second replication cycle, with redistribution observed. Three biological replicates were used; bars show means ± SD. All p-values ≤ 0.001 (***), using two-tailed unpaired Student’s t-test. Panels A and B show maximum-intensity projections of Z-stack images. Panel D, E and F display single optical plane images acquired during live imaging. Scale bars = 1 µm.

Live-cell imaging confirmed these dynamics (Video S4, S5). In untreated parasites, M-MIC2 entered the RB ∼7 h post-replication and was later redistributed to daughter cells (Figure 8D, Video S4). Under auxin treatment, M-MIC2 accumulated post-budding and failed to redistribute (Figure 8E; Video S5). Washout of auxin rescued microneme redistribution during the subsequent replication cycle, although live visualization required intensity adjustments due to differences in signal intensity between the accumulated and redistributed microneme pools (Figure 8F; Video S6). Redistribution was not observed immediately upon MyoF re-expression, but rather during the following replication cycle. This delay suggests that restoration of MyoF-dependent trafficking may require sufficient recovery of MyoF function before efficient redistribution of maternal micronemes can occur.

To assess whether microneme degradation occurred during retention within the residual body, the fluorescence intensity of individual maternal micronemes was compared between control vacuoles and vacuoles subjected to MyoF depletion followed by auxin washout (Figure S10C, D). No significant difference in microneme fluorescence intensity was detected between control micronemes and micronemes that had remained within the RB for 24 h prior to redistribution. These results indicate that retention within the RB does not result in detectable loss of maternal microneme signal. Together, these findings establish that MyoF is essential for recycling maternal micronemes and rhoptries via the RB, redefining the RB as a dynamic trafficking hub, not merely a degradative compartment. MyoF acts as a mechanistic bridge between F-actin dynamics and organelle inheritance, orchestrating precise redistribution of maternal cargo.

Discussion

Use of dual labelling to differentiate maternal and de novo material

Understanding the complexity of organelle inheritance in Toxoplasma gondii requires tools capable of resolving both the spatial and temporal dynamics of protein trafficking. Conventional single-labelling approaches have been limited to static snapshots, obscuring key processes such as organelle biogenesis, recycling, and selective degradation. To overcome these challenges, we employed a dual HaloTag pulse-chase labelling strategy to distinguish maternally inherited from de novo–synthesized proteins. This approach, previously underutilized in apicomplexan parasites (Koreny et al., 2023; Periz et al., 2019), allowed us to resolve protein fates across replication cycles and map distinct inheritance routes for multiple organelles (Figure 9A).

Schematic summaries.

A: Summary of the three fates of the maternal organelles. Group 1: Efficient, intact organelle recycling. De novo organelles are generated independently of the maternal organelle but in a similar location. Both de novo and maternal are distributed to the daughter cells. The Fluorescence intensity of the maternal organelle is relatively stable. Group 2: Even distribution of maternal material through organelle expansion with the insertion of the de novo material in the mother organelle. The Fluorescence intensity of the maternal organelle is divided by two at each replication step but their signal surface is increased by two. Group 3: Almost exclusive de novo synthesis. The de novo organelle is generated independently of the maternal organelle and in a different location. The maternal organelle is not clearly observable after a cycle of replication. The fluorescence intensity of the maternal organelle drastically drops after the first replication. Magenta: maternal organelle, Green: de novo organelle, Yellow: Colocalization between maternal and de novo. B: Summary of the MyoF regulated inheritance. The maternal and the de novo micronemes are generated independently. In presence of MyoF (F1), the daughter cells form a chimera possessing both maternal and de novo micronemes, which are distributed in a relatively equal manner even after multiple replication cycles. In absence of MyoF, (F2) the inherited micronemes accumulate in the residual body and the daugthers cells are mostly composed of de novo micronemes. If the depletion of MyoF is maintained (F2.1) more and more inherited micronemes accumulate inside the residual body without important degradation. In opposition, if the expression of the MyoF is reestablished by the auxin removal (F3), the inherited micronemes stuck inside the residual body are redistributed but in a less equal manner than usually observed for control.

The residual body as a regulatory hub

Our findings reaffirm the emerging view of the RB as an active, multifunctional structure rather than a passive cytoplasmic remnant (Frenal et al., 2017; Periz et al., 2017a). First described over 50 years ago (Sheffield and Melton, 1968), the RB was long believed to serve as a repository for discarded material, often called a “waste bin” (Nishi et al., 2008). However, we show that the RB temporarily stores maternal secretory organelles, such as micronemes and rhoptries, which are later redistributed to daughter cells in a MyoF-dependent manner (Figure 9B). Live-cell imaging revealed that organelles in the RB are not static but can re-enter functional pathways (Figures 8), establishing the RB as a trafficking hub (Periz et al., 2019). This active role is further supported by the dependence of RB-mediated recycling on F-actin and the class XXII myosin MyoF, which we show to be essential for retrieval of maternal MIC2 and RON2 but dispensable for Golgi inheritance although we noticed a fragmentation of the Golgi, which has been described to depend on MyoF (Carmeille et al., 2021). The RB is also critical for cytoplasmic continuity across parasites, synchronizing replication and facilitating inter-parasite protein exchange (Frenal et al., 2017; Muniz-Hernandez et al., 2011; Periz et al., 2017a). Its formation is dependent on actin-nucleating factors and myosins, indicating a complex and regulated origin rather than a passive leftover of cytokinesis (Stortz et al., 2019; Tosetti et al., 2019). Given these parallels, the RB may be functionally analogous to the mammalian midbody remnant, which regulates post-mitotic signaling and intracellular trafficking (Kuriyama et al., 2025; Peterman and Prekeris, 2019).

Fate determination: inheritance or degradation?

An important question raised by our study is how the parasite determines whether maternal proteins are recycled or degraded. Our results indicate that this decision is selective rather than stochastic (Figure 9). For instance, maternal micronemes persist for at least six replication cycles in the absence of MyoF, without signs of degradation (Figure S10C, D). In contrast, IMC proteins like GAPM1a and IMC1 are lost within a single cycle and show no evidence of recycling. Our data are consistent with degradation occurring within the RB., raising questions about the underlying mechanism. Although T. gondii encodes homologs of lysosomal and autophagic degradation machinery (Besteiro et al., 2011; Smith et al., 2021; Thaprawat et al., 2025), none have been localized to the RB or directly linked to IMC degradation. However, ubiquitin ligases and proteasomal components have been identified in the RB proteome (O’Shaughnessy et al., 2023; Que et al., 2002). IMC proteins are also part of the ubiquitinated proteome (Silmon de Monerri et al., 2015), and recent work showed that deletion of the kinase ERK7 prevents sequestration of the E3 ligase CSAR1, resulting in aberrant degradation of conoids (O’Shaughnessy et al., 2023). These findings support a model in which protein fate within the RB is determined through selective ubiquitination and regulated proteolysis.

A call to reassess recycling factors in Toxoplasma gondii

The recognition of the residual body (RB) as a key site for organelle recycling has fundamentally revised our understanding of intracellular trafficking in T. gondii (Periz et al., 2019; Tosetti et al., 2019). Earlier models emphasized de novo synthesis of organelles and proposed a repurposing of classical endocytic machinery for secretory functions (Tomavo, 2014). However, our HaloTag-based pulse-chase studies show that secretory organelles such as micronemes and rhoptries are extensively recycled via the RB, in a process dependent on F-actin and the unconventional myosin MyoF.

This shift in our understanding calls for a systematic re-evaluation of trafficking factors traditionally implicated in protein targeting, vesicle transport, and organelle biogenesis. Many of these, particularly Rab-GTPases, SNAREs, and dynamins were previously studied without consideration of recycling pathways. Notably, accumulation of microneme material in the RB has been observed upon perturbation of several trafficking regulators, including Rab5A (Kremer et al., 2013), ArlX3 (Klinger et al., 2024), SORTLR (Sloves et al., 2012), VPS8 (Morlon-Guyot et al., 2018), and VPS11 (Morlon-Guyot et al., 2015). Yet, whether these phenotypes reflect a block in secretion, synthesis, or recycling remains unclear. To resolve this, a targeted reanalysis of trafficking factors using our dual-labelling HaloTag assay combined with the splitCas9 system (Li et al., 2022) would allow high-throughput functional screening to specifically distinguish defects in recycling from those in de novo biogenesis, offering a path to redefine the roles of classical and lineage-specific trafficking regulators in T. gondii.

Material and methods

Parasite culture and genetic manipulation

Growth and generation of transgenic T. gondii

T. gondii tachyzoites from the RH strain and derived lines, including RH Δku80/TATi and RH Δku80/Tir1, were maintained at 37 °C with 5% CO₂ in human foreskin fibroblasts (HFFs; ATCC, SCRC 1041) cultured in Dulbecco’s Modified Eagle Medium (DMEM; Sigma, D6546) supplemented with 10% fetal bovine serum (FBS; BioSell, FBS.US.0500), 4 mM L-glutamate (Sigma, G7513), and 20 μg/mL gentamycin (Sigma, G1397) as previously described (Gras et al., 2019).

Generation of transgenic parasites

New strains were generated using CRISPR/Cas9 as previously described (Li et al., 2022). Guide RNAs (gRNAs) targeting the regions of interest were designed using EuPaGDT (Alvarez-Jarreta et al., 2024). All gRNA and primer sequences are listed in Supplementary Table 1. Briefly, gRNA oligos were annealed, ligated into the Cas9-YFP vector, and verified by sequencing (Eurofins Genomics). Repair templates were generated by PCR amplification of the Halo or YFP tag flanked by 50 bp homology arms to the target gene, using Q5 High-Fidelity DNA Polymerase (New England BioLabs). PCR products were purified with a PCR purification kit (Blirt, EM26.1). Tachyzoites were mixed with the repair template and 10 µg of the corresponding Cas9 vector, and transfected using the Amaxa 4D-Nucleofector system (Lonza, AAF-1003X). Transfected parasites were allowed to invade fresh HFFs and replicate for 48 h. Following manual egress and filtration through a 3 μm filter, Cas9-YFP-expressing parasites were enriched by FACS (FACSAria III, BD Biosciences) and sorted into 96-well plates. Correct integration of the repair template was confirmed by PCR.

Labelling and characterization

Maternal and de-novo protein discrimination

Fresh tachyzoites expressing Halo-tagged reporters were mechanically egressed, filtered through a 3 μm filter, and resuspended in cold DMEM containing a membrane-permeable Halo dye, Janelia Fluor® 646 (1:1000, Promega), for 1 h. Parasites were centrifuged at 2,500 rpm for 5 min and washed three times with fresh medium to remove unbound dye. Parasites were then seeded onto HFF-covered Ibidi live-cell dishes for overnight replication. After replication, a second labeling was performed with a new membrane-permeable Halo dye, Janelia Fluor® 549 (1:1000, Promega) for 1 h at 37 °C, followed by three washes before imaging.

Fluorescence intensity across the replication

Parasites were labeled as described above and allowed to replicate for 24 h on HFF-coated Ibidi live-cell dishes. Approximately 15 fields of view were imaged using Z-stacks spanning 3 μm centered on the vacuoles (Figure S11A). For each replication stage (1, 2, 4, and 8 parasites per vacuole), individual tachyzoites were sampled across multiple vacuoles.

Maximum-intensity projections were generated from non-deconvolved images. Fluorescence intensity (FI) was quantified as the maximum gray value measured within regions of interest (ROIs) drawn on individual tachyzoites from each vacuole stages (Figure S11B). ROIs excluded overlapping parasites, neighboring vacuoles, and regions with atypical signal intensity (Figure S11C). For each biological replicate, up to 25 tachyzoites per replication stage were analyzed, and the mean FI value was calculated for each stage. The highest mean FI observed among the stages within a replicate was defined as 100%, and FI values for the other stages were expressed relative to this maximum. Relative FI values were then averaged across three independent biological replicates. Data are presented as mean ± SD. A total of 272, 286, 251, 276, 223, 284, 265, 300 and 260 vacuoles were analysed for RON1, ROP1, SortLR, MyoA, TIC20, ANKER1, GAPM1a, IMC1 and MIC2 respectively.

Fluorescence intensity of the daughter cells vs new mother cells

Images were processed as described above. Fluorescence intensity was quantified as the maximum gray value measured within regions of interest (ROIs) drawn on individual daughter cells and on newly formed mother cells. To reliably discriminate newly formed mother IMC from daughter IMC (both in the same channel colour, Jan. 549), the channel corresponding to the initial maternal IMC (Jan. 646) was intentionally overexposed, allowing unambiguous identification of the new mother cell IMC signal. ROIs were drawn on non-overlapping regions and excluded areas containing overlapping parasites, neighboring vacuoles, or atypical signal intensity, following the same exclusion criteria used for replication-stage measurements (Figure S11D). Quantification was performed on maximum-intensity projections generated from non-deconvolved Z-stacks. Data are presented as mean ± SD from vacuoles pooled across three independent biological replicates.

Measurement of the signal area of the fluorescence

To quantify the average and total fluorescent area, images were analyzed in Fiji. Up to 25 isolated vacuoles per stage (1 to 8) were cropped and analyzed individually. A threshold was applied to the maternal signal, and both average and total signal area were measured. Three independent biological replicates were performed, and values are reported as mean ± SD.

Quantification of percentage of parasite per vacuole with maternal protein

Images were analyzed in Fiji to determine the proportion of parasites per vacuole with maternal protein signal. Up to 25 vacuoles per stage (1 to 8) were cropped and analyzed individually. Vacuoles where all parasites retained maternal signal were scored as 100%. If some parasites lacked signal, the percentage of maternal-positive parasites was calculated. The experiment was performed in three independent biological replicates. Values are reported as mean ± SD.

Live replication assay

Parasites were labeled with Janelia Fluor® 646 (1:1000, Promega) for 1 h and washed three times before transfer to Ibidi live-cell dishes covered with HFFs. Parasites were allowed to invade for 1–2 h to obtain ∼10 parasites per field of view. Excess parasites were removed by washing three times. Imaging was performed on a Leica DMI8 microscope at 37 °C in 5% CO₂. Images were acquired every 15–30 min for 12 h using minimal laser power and exposure. Experiments were performed in three biological replicates and analyzed in Fiji.

Colocalization assays: ANKER1-Halo with HDEL-GFP

Parasites expressing ANKER1-Halo were transiently transfected with an HDEL-GFP plasmid to label the ER. After overnight replication in HFF-coated Ibidi dishes, parasites were labeled with Janelia Fluor® 646 (1:1000, Promega) for 1 h and washed three times prior to imaging. Z-stacks of HDEL-GFP-positive parasites were acquired. The proportion of vacuoles displaying ER localization of ANKER1-Halo was assessed manually in Fiji/ImageJ by inspection of full Z-stacks and maximum-intensity projections. In addition, the degree of colocalization between ANKER1-Halo and HDEL-GFP was quantified using Pearson’s correlation coefficient calculated with the Coloc2 plugin in Fiji/ImageJ. Pearson analysis was performed on 30 individual parasites obtained from three independent biological replicates (10 parasites per replicate). The mean Pearson’s correlation coefficient was 0.92 ± 0.04 (mean ± SD).

Colocalization assays: MIC2/RON2 with Cb-Emerald

For analysis of association with the F-actin network, MIC2 or RON2 was endogenously tagged and labeled in a strain stably expressing Cb-Emerald. Freshly egressed parasites were labeled with Janelia Fluor® 646 (1:1000, Promega) for 1 h, washed, and seeded onto HFF-coated live-cell dishes. After 24 h of replication, parasites were fixed in 4% paraformaldehyde and imaged. Z-stacks of vacuoles displaying a clearly defined F-actin network were acquired, and the association of MIC2- or RON2-positive structures with F-actin was assessed in both the parasite cytoplasm and the RB. Analysis was performed manually in Fiji/ImageJ by inspection of full Z-stacks and maximum-intensity projections. The RB was defined by the presence of thick bundles of F-actin located at the basal pole of the parasites and forming a continuous structure connecting parasites within the vacuole, consistent with the characteristic morphology of the RB. Cytoplasmic F-actin was defined as all actin signal located within the parasite body excluding this RB-associated region. Three independent biological replicates were analyzed, and data are reported as mean ± SD.

Quantification of the maternal microneme distribution

To assess the distribution of maternal micronemes at stage 8, parasites were labeled as described in “Maternal and de novo protein discrimination.” Twenty-five stage 8 vacuoles were selected. Using the de novo MIC2 signal, the outline of each tachyzoite was traced, and the number of maternal MIC2-positive parasites was counted. Tachyzoites were numbered 1 to 8 from left to right. Three biological replicates were performed. Data are reported as mean ± SD. Images were analyzed in Fiji.

Probability of maternal microneme inheritance

To evaluate whether maternal micronemes could be distributed among daughter parasites by stochastic segregation alone, we modeled the allocation of 64 micronemes from a mother parasite to 32 daughter tachyzoites. We specifically assessed the probability that each daughter would receive between 0 and 3 micronemes, corresponding to the range observed experimentally. As an approximation, microneme allocation was modeled as an independent random process following a Poisson distribution: P(0≤X≤3)=e−λ(1+λ+λ22!+λ33!) where X is the number of micronemes inherited by a daughter and λ=6432=2.

This gives: P(0≤X≤3)≈0.8571 Assuming independent inheritance events, the probability that all 32 daughters within a vacuole receive between 0 and 3 micronemes is: 0.857132≈0.0072 Finally, assuming independence between vacuoles, the probability of observing this outcome in 20 consecutive vacuoles is: 0.007220≈1.4×10−4 These calculations indicate that the experimentally observed microneme distribution would be highly unlikely under a simple stochastic segregation model and are therefore consistent with the existence of a regulated mechanism promoting balanced microneme inheritance.

Phenotypic assays

Induction of MyoF KD

MyoF-mAID parasites were genetically modified to endogenously tag MIC2, RON2 and SortLR. As previously described for MyoF mAID parasites (Carmeille et al., 2021), MyoF-mAID MIC2/RON2 or SortLR-Halo parasites were induced +/- auxin for 4h prior to labelling and experiments.

Quantification of the accumulation of protein in the residual body

Parasites were treated ± auxin for 4 h and labeled as described above. After 24 h replication on live-cell dishes, ∼100 vacuoles were imaged across ∼15 fields of view. The percentage of vacuoles showing accumulation of material in the residual body was calculated. The experiment was performed with three independent biological triplicates. The values were then expressed as the mean values of the three independent experiments ± SD. Images were analysed via Fiji.

Auxin chased experiment

Parasites were induced ± auxin for 4 h prior to labeling as described above in “Maternal and de novo protein discrimination”. Following transfer to live-cell dishes, parasites were allowed to replicate for 48 h in the continued presence or absence of auxin before imaging. For the auxin chase condition, parasites were first maintained in auxin-containing medium for 24 h. The medium was then removed, and dishes were washed three times with fresh medium to ensure complete auxin removal before addition of fresh medium lacking auxin. Parasites were subsequently allowed to replicate for an additional 24 h prior to imaging.

To assess recovery of MyoF expression following auxin washout, parallel samples were fixed in 4% paraformaldehyde and processed for immunofluorescence using an anti-HA antibody to detect the MyoF-mAID-HA fusion protein. Vacuoles were categorized according to the presence or absence of detectable MyoF signal and correlated with the corresponding microneme distribution phenotype.

Approximately 15 fields of view were imaged per condition, corresponding to ∼100 vacuoles. The percentage of vacuoles displaying microneme accumulation within the residual body was determined. In addition to the “accumulated” and “normal” phenotypes, a third category termed “uneven distribution” was defined for vacuoles in which micronemes were no longer retained within the residual body but had not yet reached the homogeneous distribution observed in control parasites. The frequency of each phenotype was quantified. Experiments were performed in three independent biological replicates, and values are presented as mean ± SD. Image analysis was performed using Fiji/ImageJ.

MyoF-mAID Live replication assay

Parasites were induced +/- auxin for 4h prior labelling. Parasites were labelled with the Halo membrane permeable dye Halo 646 (Janelia Fluor®646 1:1000 Promega) for 1 h and washed three-time prior transfer onto a live cell dish covered with HFF. Parasites were allowed to invade for 1h-2h to obtain enough invasion events to reach an average of 10 parasites per FOV. After the invasion, the excess of parasites was removed by 3 washes. A new membrane permeable dye Halo 549 (Janelia Fluor®549 1:10000 Promega) was added to the media to record the de-novo generation of the POI live. Live cell dishes were then transferred to the Leica-DMI8, heated at 37°C under a chamber containing 5% CO2. Laser power and exposure were adjusted to the lowest values allowing reliable imaging. Imaged were taken every 15-30 minutes to follow the tagged protein behaviour during replication for 12h. For the Chased assay, the imaging started after the removal of the auxin after 24h of replication as described above in “Auxin chased experiment.” _The experiments were performed with three independent biological triplicates. Images were analysed via Fiji.

Imaging

Widefield microscopy

Unless stated otherwise, all images were acquired on a Leica-DMI8, objective 100x with the LasX software (v3.7.4). Fiji (v1.53c) was used to analyse the picture and all counts were made manually. LasX software (v.) from Leica was used to obtain parasite imaging data and All images and movies were processed using Fiji (ImageJ) software v Image Processing Software (Schindelin et al., 2012).

Software

Fiji ( FIJI ImageJ v1.54f) was used to analyze the picture and all counts were made manually.

Data analysis

All data were plotted using Microsoft Excel

Statistical analysis

Two-tailed unpaired Student’s t-tests were performed to evaluate statistical significance.

-For Figures 2C, 2E, 2F, 3B, 3D, 4B, 4C, 5C (stage 1–2), and Supplementary Figures S1A, S1B, S2C–S2G, S4B–S4D, differences between stages 1–2, 2–4, and 4–8 were assessed.

-For Figures 2B, 4C, and 5C (stage 2–8), variance could not be calculated because values were constant (0 or 100%), and these were therefore labeled as non-significant.

-For Figures 7D, S8D, and S9B, differences between uninduced controls and MyoF knockdown (KD) induced samples were analyzed.

-For Figure 8C, the distribution of phenotypes (normal, accumulated, and redistributed micronemes) was compared between control, auxin-induced (48 h), and auxin-washed (chased) conditions.

-For Figure S5B, microneme distribution across positions 1–2, 2–3, 3–4, 4–5, 5–6, 6–7, and 7–8 was analyzed; no statistically significant differences were observed, and all were labeled as non-significant.

-Finally, for Figure S7F, the difference in fluorescence intensity between daughter and mother IMC signals was assessed.

Data are presented as mean ± SD. Statistical significance ns: not significant (p ≥ 0.05); * : p < 0.05; ** : p < 0.01; *** : p < 0.001.

eLife Assessment

This important study provides new insights into the patterns of organelle inheritance in the protozoan parasite Toxoplasma gondii. The authors introduce an innovative dual-labeling approach to distinguish maternally inherited from de novo synthesized organelles, representing convincing evidence that different organelles follow distinct inheritance fates during parasite replication. Future studies will be needed to determine whether the residual body functions as a central recycling hub, as the current data are also consistent with alternative models.

Reviewer #1 (Public review):

Summary:

This work asks the question of how different organelles and structures in the apicomplexan parasite Toxoplasma gondii are recycled and/or segregated to the daughter cells during cell replication. In particular, they consider an unusual cell structure called the residual body that links replicating cells during the intracellular infection stage of this parasite. The residual body has historically been considered a 'dumping ground' for unnecessary relics of the mother cell during division, but this notion is increasingly being revised. Indeed, cell replication in Toxoplasma is often misinterpreted as cell division (cytokinesis), but in fact, the cell replicates its organelles and structures to multiple 10s of copies in seemingly distinctly formed daughter cells, but cytokinesis is delayed for many such cycles and typically only occurs simultaneously with parasite egress from its host cell. The residual body is, in fact, the connection between these pre-cytokinetic replicated daughters, and effectively, this is still a single cell at this stage. The authors have previously shown that an actin network extends through the residual body between these daughter cells, and ER and mitochondria common to all cells are also linked through this structure. This study examining the fates of organelles during cell replication is timely for continuing our understanding of how this fascinating component of the cell participates in these processes. The authors use Halo-tags as their principal tool to track discrete populations of proteins, labelling their organelle locations, and this provides beautiful insight into these processes.

Strengths:

Using dyes conjugated to Halo tags this work elegantly tracks the fates of proteins synthesised by an original 'mother' cell over several replication cycles of pre-cytokinetic 'daughters'. Using this tool, they show that some organelles are made intact just once and that some of these can be subsequently sorted to the daughters (micronemes and rhoptries) while others are dismantled (IMC) and the daughters must make their own. A third set of organelles (largely synthesis, sorting and metabolic compartments) are divided and inherited, and new daughter-synthesised proteins are added to the preexisting maternal proteins in these structures. A role for actin and myosin is clearly demonstrated for micronemes and rhoptries, and this correlates with their relatively late inheritance into the developing daughters. Overall, this work gives clarity to the behaviours of several cell structures during replication and paves the way to better understanding the mechanisms that drive the differences between structures and the universality of these processes in other apicomplexan parasites. In particular, this study shows that the residue body is a region of the cell syncytium that organelles can be actively transported from. Therefore, it is a space that can actively contribute to the segregation of the late segregating micronemes and rhoptries.

Weaknesses:

In addressing the question of residual body participation in sorting of organelles, a clear definition of this structure is required including when and where it is delineated from the posterior of a mother cell during the formation of daughter structures. The authors' definition is as follows: 'The RB originates from the collapse of the maternal parasite during daughter cell budding and occupies the space previously occupied by the mother cell.' As such, a clear marker of the mother cell 'collapse' is required, but such a marker is not identified or used in the study to separate what might be considered an active part of the mother cell during early daughter formation, and the residual body. This might seem like moot a point, but it would help to give clarity to notions of recycling and 'reservoirs'. Mother cells retain their active invasion apparatus until very late in daughter formation and the need for micronemes and rhoptries to be released from this service late in the process might explain why they are only then trafficked to the cell posterior and then into the daughters. So, is this a distinct 'residual body' body function/reservoir or just a spatial constraint of this sequence of daughter formation? The authors elegantly show that MyoF is necessary for segregation of micronemes and rhoptries into daughters, and that MyoF depletion leads to accumulation of these organelles within the residual body. Moreover, restored expression of MyoF can then recover these organelles. This clearly demonstrates the activity of the residual body as part of the syncytium space that participates in the maintenance of the vacuole. But does it imply that this space necessarily handles all inherited micronemes and rhoptries as a 'trafficking hub'? My concern with the lack of a clear definition could provide some misinterpretation or overinterpretation of the contribution residual body.

A further, remarkable conclusion is that maternal micronemes are evenly segregated into daughters through an active process for 'balanced microneme inheritance'. The proportion of maternal micronemes is quantified up to the 8-cell stage and shown to be not significantly different between cells. But would this result be expected with random assortment at this stage? The authors model the probability of a 32-cell stage vacuole occurring with each daughter having within 0-3 maternal micronemes and this is considered unlikely. However, the authors neither present the modelling for the 8-cell stage or show quantification of 32-cell vacuoles. They do show some images of large vacuoles, but it is not possible to determine the distribution of maternal micronemes in these images. A regulated process of segregation would require a complex mechanism where some form of microneme counting would be required to create the proposed balance. It is, therefore, important to have strong data supporting such a hypothesis, but this is not currently presented.

Reviewer #2 (Public review):

Summary:

Toxoplasma gondii is an obligate intracellular parasite and the causative agent of toxoplasmosis. Parasite invasion of host cells, intracellular replication, and subsequent egress, which results in destruction of the infected cell, are central to pathogenicity. This manuscript focuses on understanding how maternal resources, specifically cellular organelles, are shared between daughter parasites during cell division. Many organelles are present as a single copy, making their division and inheritance essential for successful replication. In T. gondii, our understanding of how organelles are divided during cell division remains limited, and this study helps address this important knowledge gap.

Strengths:

The major strength of this study is the use of a Halo-based pulse-chase assay to characterize patterns of organelle inheritance and to monitor protein synthesis, turnover, and movement. This approach will be of considerable interest to the field. Using this method, the authors identify three major modes of organelle inheritance:

(1) Organelles present in multiple copies (such as micronemes and rhoptries) are partitioned between daughter parasites, with additional contributions from newly formed vesicles. Newly synthesized and pre-existing material remain as distinct populations within the cell.

(2) Single-copy organelles, such as the Golgi and apicoplast, are expanded through the incorporation of newly synthesized material before division.

(3) Cytoskeletal structures are synthesized de novo during each round of cell division.

These findings provide a more refined understanding of organelle inheritance and demonstrate that secretory organelles are not generated entirely de novo during each round of division, as was previously thought.

The paper places particular emphasis on the fate of maternal micronemes and rhoptries during division. The data show that (1) during division in wild-type cells, maternal micronemes and rhoptries are detectable in the residual body (RB); however, the majority of these organelles are localized within the parasite body, either at the apical or basal ends of the daughter parasites (Fig. 6). (2) In the absence of the myosin motor MyoF, micronemes and rhoptries accumulate in the residual body and are not properly trafficked to the daughter cells. Upon restoration of MyoF protein levels, these organelles redistribute to the daughter cells, although in an uneven manner.

Weaknesses:

The second half of the paper focuses on a more detailed characterization of microneme and rhoptry recycling. The authors strongly argue that the RB is a central hub for recycling micronemes and rhoptries; however, this conclusion is not fully supported by the data. For example, the authors state:

Line 227:

"Notably, after endodyogeny was completed, M-MIC2 was redistributed from the RB to the apical tip of the daughter cells (Figure 6A, 11:30, 16:00 h), confirming that the RB serves as a temporary reservoir during microneme recycling (Periz et al., 2019)."

Line 231:

"In approximately 90% of parasites undergoing replication, M-RON2 was integrated into daughter rhoptries prior to mother cell collapse and formation of the RB (Figure 6B, 4:15-4:30 h and 10:30-10:45 h). Like M-MIC2, M-RON2 was occasionally detected in the RB, though less prominently, suggesting more rapid, tightly regulated, or more efficient recycling due to their lower number."

Line 326:

"However, we show that the RB temporarily stores maternal secretory organelles, such as micronemes and rhoptries, which are later redistributed to daughter cells in a MyoF-dependent manner (Figure 9B)."

Thus, the model that all microneme and rhoptry trafficking is RB-dependent is based primarily on the MyoF depletion phenotype (which results in RB accumulation) together with the observation that a relatively small amount of maternal microneme and rhoptry material is detectable in the RB of wild-type parasites. Although the authors' interpretation-that recycling is RB-dependent-is one possible explanation, alternative models are not discussed.

For example, an alternative possibility is that the majority of micronemes and rhoptries are trafficked directly from the apical end of the mother parasite to the daughter cells without passing through the RB. In this scenario, only a subset of the organelles would enter the residual body, perhaps reflecting imperfect trafficking efficiency rather than an obligatory recycling step. Loss of MyoF would impair this trafficking pathway, resulting in the accumulation of secretory organelles within the RB. In other words, RB accumulation could be a consequence of MyoF depletion rather than evidence that all trafficking in wild-type parasites normally proceeds through the RB.

This alternative interpretation seems particularly relevant for the rhoptries, given that the authors themselves state that "M-RON2 was integrated into daughter rhoptries prior to mother cell collapse and formation of the RB."

Other comments:

Figure S10C

To determine whether microneme degradation occurs in the RB, the authors quantified the fluorescence intensity of individual micronemes in control parasites and following auxin washout, showing that after redistribution the fluorescence intensity of individual vesicles is unchanged. However, this is not the appropriate analysis to address the question being asked. To conclude that micronemes are not degraded, the authors would need to quantify the total fluorescence intensity within the entire vacuole. For example, if half of the micronemes were degraded, the remaining micronemes would be expected to retain the same fluorescence intensity as those in the control parasites. Thus, unchanged fluorescence intensity of individual vesicles does not exclude the possibility that degradation has occurred.

Reviewer #3 (Public review):

Summary:

Knoerzer-Suckow et al. explore the mechanisms of organelle inheritance during endodyogeny in Toxoplasma gondii using an innovative dual-labeling approach to track the distribution of maternal organelles into daughter parasites. They can clearly distinguish between maternal and daughter-derived organelles using their dual-labeling Halo Tag approach. They reveal that different organelles are trafficked to daughter parasites in three broad patterns they have binned into groups. Their findings reveal a role for MyoF in the inheritance of micronemes and rhoptries, and notably, they observe that the inner membrane complex (IMC) is not recycled. Instead, the IMC undergoes a pronounced relocalization to the posterior of the maternal cell, where it is likely targeted for degradation.

Strength:

The data surrounding their MyoF knockdown experiments, IMC degradation, and trafficking of MIC2 after auxin washout are convincing. These data add to the knowledge of how organelle inheritance occurs in T. gondii, increasing the field's understanding of endodyogeny.

Weakness:

The inability to achieve higher temporal resolution due to phototoxicity precluded tracking of single micronemes, thus it remains possible that some micronemes follow a path similar to rhoptries and enter daughter cells before development of the residual body while others are recycled via the residual body.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Weaknesses:

(1) In addressing the question of residual body participation in sorting of organelles, a clear definition of this structure is required including when and where it is delineated from the posterior of a mother cell during the formation of daughter structures. The authors' definition is as follows: 'The RB originates from the collapse of the maternal parasite during daughter cell budding and occupies the space previously occupied by the mother cell.' As such, a clear marker of the mother cell 'collapse' is required, but such a marker is not identified or used in the study to separate what might be considered an active part of the mother cell during early daughter formation, and the residual body. This might seem like moot a point, but it would help to give clarity to notions of recycling and 'reservoirs'. Mother cells retain their active invasion apparatus until very late in daughter formation and the need for micronemes and rhoptries to be released from this service late in the process might explain why they are only then trafficked to the cell posterior and then into the daughters. So, is this a distinct 'residual body' body function/reservoir or just a spatial constraint of this sequence of daughter formation? The authors elegantly show that MyoF is necessary for segregation of micronemes and rhoptries into daughters, and that MyoF depletion leads to accumulation of these organelles within the residual body. Moreover, restored expression of MyoF can then recover these organelles. This clearly demonstrates the activity of the residual body as part of the syncytium space that participates in the maintenance of the vacuole. But does it imply that this space necessarily handles all inherited micronemes and rhoptries as a 'trafficking hub'? My concern with the lack of a clear definition could provide some misinterpretation or overinterpretation of the contribution residual body.

We thank the reviewer for raising this conceptual point. We agree that, in the absence of a molecular marker that uniquely defines the nascent RB, the precise transition between posterior maternal cytoplasm and a morphologically distinct RB cannot be determined during early daughter formation. We have therefore clarified our terminology in the revised manuscript and define the RB operationally as the posterior compartment/connection between daughter parasites. We also avoid assigning early posterior trafficking events unambiguously to a fully formed RB. We further agree that the current data do not establish that all inherited micronemes and rhoptries must transit through the RB. We have therefore revised the Results and Discussion and softened terminology such as “central trafficking hub” and now conclude that the RB represents an important dynamic compartment in organelle recycling and redistribution, without implying that it is an obligatory intermediate for every inherited secretory organelle.

(2) A further, remarkable conclusion is that maternal micronemes are evenly segregated into daughters through an active process for 'balanced microneme inheritance'. The proportion of maternal micronemes is quantified up to the 8-cell stage and shown to be not significantly different between cells. But would this result be expected with random assortment at this stage? The authors model the probability of a 32-cell stage vacuole occurring with each daughter having within 0-3 maternal micronemes and this is considered unlikely. However, the authors neither present the modelling for the 8cell stage or show quantification of 32-cell vacuoles. They do show some images of large vacuoles, but it is not possible to determine the distribution of maternal micronemes in these images. A regulated process of segregation would require a complex mechanism where some form of microneme counting would be required to create the proposed balance. It is, therefore, important to have strong data supporting such a hypothesis, but this is not currently presented.

We agree that the previous wording implied a mechanistic conclusion beyond what can be established from the present dataset. We have therefore revised the manuscript so that the relatively even distribution of maternal micronemes at the 8-cell stage is presented as an observation that is consistent with a non-random or regulated partitioning process, rather than evidence for an established microneme-counting mechanism. The 32-cell model is now presented as supportive rather than definitive evidence, and we explicitly acknowledge that the quantitative experimental dataset was obtained at the 8-cell stage.

Reviewer #2 (Public review):

(1) The second half of the paper focuses on a more detailed characterization of microneme and rhoptry recycling. The authors strongly argue that the RB is a central hub for recycling micronemes and rhoptries; however, this conclusion is not fully supported by the data. For example, the authors state…Thus, the model that all microneme and rhoptry trafficking is RB-dependent is based primarily on the MyoF depletion phenotype (which results in RB accumulation) together with the observation that a relatively small amount of maternal microneme and rhoptry material is detectable in the RB of wild-type parasites. Although the authors' interpretation-that recycling is RB-dependent-is one possible explanation, alternative models are not discussed. For example, an alternative possibility is that the majority of micronemes and rhoptries are trafficked directly from the apical end of the mother parasite to the daughter cells without passing through the RB. In this scenario, only a subset of the organelles would enter the residual body, perhaps reflecting imperfect trafficking efficiency rather than an obligatory recycling step. Loss of MyoF would impair this trafficking pathway, resulting in the accumulation of secretory organelles within the RB. In other words, RB accumulation could be a consequence of MyoF depletion rather than evidence that all trafficking in wild-type parasites normally proceeds through the RB. This alternative interpretation seems particularly relevant for the rhoptries, given that the authors themselves state that "M-RON2 was integrated into daughter rhoptries prior to mother cell collapse and formation of the RB."

We agree with the reviewer that the current data do not establish obligatory transit of all maternal micronemes and rhoptries through the RB. We have revised the manuscript throughout to make this distinction explicit. In particular, we now emphasize that maternal MIC2 can be directly observed entering the RB, whereas most maternal RON2 is incorporated into daughter rhoptries before mother-cell collapse and formation of a morphologically distinct RB. This observation leaves open the possibility that a substantial fraction of maternal rhoptries is transferred directly from the mother to developing daughters. We have also revised our interpretation of the MyoF-depletion phenotype. The accumulation of maternal MIC2 and RON2 in the RB following MyoF depletion demonstrates that MyoF is required for efficient redistribution of both organelle populations, but does not by itself demonstrate that both normally follow an identical spatial route through the RB. We now explicitly state that their precise trafficking routes and timing may differ.

We nevertheless retain the conclusion that the RB is a dynamic compartment involved in organelle recycling because maternal MIC2 can be directly observed entering and leaving this compartment, and material accumulated there following MyoF depletion can subsequently be redistributed after restoration of MyoF.

(2) Figure S10C. To determine whether microneme degradation occurs in the RB, the authors quantified the fluorescence intensity of individual micronemes in control parasites and following auxin washout, showing that after redistribution the fluorescence intensity of individual vesicles is unchanged. However, this is not the appropriate analysis to address the question being asked. To conclude that micronemes are not degraded, the authors would need to quantify the total fluorescence intensity within the entire vacuole. For example, if half of the micronemes were degraded, the remaining micronemes would be expected to retain the same fluorescence intensity as those in the control parasites. Thus, unchanged fluorescence intensity of individual vesicles does not exclude the possibility that degradation has occurred.

We agree with this criticism and have revised the interpretation of the experiment accordingly. We no longer conclude that the analysis excludes microneme degradation. We now explicitly acknowledge that analysis of individual recovered micronemes cannot exclude degradation of a fraction of the total microneme population during RB retention. Thus, the experiment supports preservation of MIC2 signal in the recovered organelles but is no longer presented as evidence that no microneme degradation occurs.

Reviewer #3 (Public review):

Weakness:

The inability to achieve higher temporal resolution due to phototoxicity precluded tracking of single micronemes, thus it remains possible that some micronemes follow a path similar to rhoptries and enter daughter cells before development of the residual body while others are recycled via the residual body.

We agree and have incorporated this limitation into the revised interpretation. Our live imaging demonstrates that maternal MIC2 can enter the RB and subsequently redistribute to daughter parasites, but it does not establish that every individual maternal microneme follows this route. We thank the reviewer for highlighting this distinction, which has helped us clarify the model presented in the Results and Discussion.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Carruthers and Sibley, 1997, is still the cited ref for Tic20 in the apicoplast despite the authors saying that would correct this to a van Dooren publication.

Corrected

(2) In Figure 2B three biological replicates were used however there are no error bars shown. The only reason for performing replicates is the observe the variance in the data, so if this is not shown the replicates are effectively meaningless. I strongly advice that error bars are given to indicate this seeing that the data in this figure form the basis of the major conclusions of the study. It might be necessary to show this in supplemental forms with fewer proteins if the error bars are too difficult to see in the combined figure.

We agree and have corrected Figure 2B to display the variability between the three independent biological replicates. Error bars now represent the standard deviation.

(3) Line 136: Can you conclude that these inheritance patterns are 'organelle-specific' when each organelle is only sampled with one or two proteins. Isn't it better to conclude that these are protein-specific, with the hypothesis that they might represent the orgnalle as a whole. I imagine that some proteins in organelles such as the apicoplast have shorter half-lives than others, and therefore some apicoplast proteins might behave like 'Group 3' proteins.

We thank the reviewer for raising this point. We agree that individual proteins within the same organelle may differ in their turnover kinetics and that analysis of one or two markers cannot establish that every molecular component of an organelle behaves identically. However, we do not think that describing the observations exclusively as protein-specific inheritance would fully reflect the biological process investigated here. The proteins analysed are established markers of defined organelles, and our conclusions are based not only on changes in fluorescence intensity, but also on the localization, morphology, partitioning, and spatial relationship between maternally inherited and newly synthesized organelle populations.

This is particularly evident for micronemes and rhoptries, where maternal and de novo material remain spatially separated and individual organelles can be followed during inheritance.

In addition, the microneme phenotype observed with MIC2 was confirmed using AMA1, MIC4, and MIC8.

We therefore retain the terminology of organelle inheritance, while acknowledging that individual proteins within a given organelle may exhibit different turnover kinetics and that the markers analysed may not represent the behaviour of every molecular component of the organelle.

(4) Line 222: It is an odd phrase to suggest that the Golgi, ER etc 'bypass' the residual body, which suggests an active avoidance mechanism. Would the authors also conclude that the nucleus 'bypasses' the RB? Moreover, the ER and mitochondria are actually known to be present in the RB forming continuous organelles between daughters in a vacuole. So again, this might be an overstatement that mispresents how the RB participates in vacuole functions.

We agree and have removed the term “bypass.” The revised text now states only that we did not observe comparable accumulation of the analysed Golgi, ER, or apicoplast markers in the RB during inheritance. This avoids implying an active avoidance mechanism and is compatible with the known continuity of ER and mitochondria through the RB.

Reviewer #2 (Recommendations for the authors):

Minor comments:

Figure 8F and Video S6: The authors should specify the time point after IAA washout at which live imaging was initiated. Does time 0 in the video correspond to the point at which IAA was removed?

We have clarified this in the Results and Methods. Auxin was removed after 24 h of replication, and live imaging was subsequently initiated. Time 0 in Figure 8F and Video S6 corresponds to the first acquired frame after auxin washout.

Figure S10 should read auxin, not auxine.

Corrected

Reviewer #3 (Recommendations for the authors):

I have no further suggestions. Congratulations to the authors for a lovely study.

Version 3

Abstract

Toxoplasma gondii replicates through endodyogeny, an unconventional form of internal budding in which two daughter cells are assembled within a single mother cell. During this process, daughter cells must acquire a full complement of organelles, which may be inherited from the mother, formed de novo, or assembled through a combination of both mechanisms. To date, the fate of maternal components during replication remains poorly understood. We previously showed that F-actin–driven dynamics generate an intravacuolar network, associated with the residual body (RB) and facilitates the recycling of microneme proteins. However, the inheritance and recycling of other organelles have not been systematically analysed. To address this, we employed a dual HaloTag-based pulse-chase fluorescence labelling strategy to distinguish between de novo-synthesized and maternally inherited proteins in replicating tachyzoites. This approach reveals three distinct modes of organelle inheritance: (1) inheritance of intact maternal organelles (e.g., rhoptries and micronemes), (2) expansion and division of pre-existing maternal organelles with incorporation of newly synthesized components (e.g., Golgi apparatus and apicoplast), and (3) degradation of maternal structures followed by de novo assembly (e.g., inner membrane complex). Furthermore, we identify Myosin F (MyoF) as an important factor in the redistribution of maternal micronemes and rhoptries. In the absence of MyoF, these organelles accumulate in the RB, while restoration of MyoF permits accumulated maternal micronemes to redistribute to daughter parasites. Together, these findings support a role for the RB as a dynamic compartment in the recycling and redistribution of maternal organelles.

Introduction

Toxoplasma gondii tachyzoites replicate through a specialized form of internal budding known as endodyogeny, wherein two daughter cells (DCs) are assembled within the cytoplasm of the mother cell (Francia and Striepen, 2014; Hu et al., 2002). This unique replication mechanism demands precise orchestration of organelle inheritance and spatial organization within the parasitophorous vacuole (PV). Apicomplexan parasites, including T. gondii, exhibit a highly polarized cellular architecture defined by specialized secretory organelles—micronemes, rhoptries, and dense granules—essential for host cell invasion and intracellular survival. These are accompanied by canonical organelles such as the Golgi apparatus, endoplasmic reticulum (ER), a single mitochondrion, and the apicoplast, a non-photosynthetic plastid. Notably, unlike most eukaryotes, T. gondii is enveloped by a pellicle consisting of the plasma membrane and an underlying inner membrane complex (IMC) (Ouologuem and Roos, 2014), a structure composed of flattened vesicles tightly associated with the subpellicular cytoskeleton (Harding and Meissner, 2014).

As DCs emerge during endodyogeny, they progressively encapsulate nearly all maternal cytoplasmic contents. While the sequence of organelle acquisition has been extensively characterized (Nishi et al., 2008), the extent to which maternal organelles and proteins are recycled remains unclear. Historically viewed as a passive remnant of maternal cytoplasm, the residual body (RB) forms at the posterior end of dividing parasites and contains material remaining from the mother cell after daughter formation. Because no RB-specific molecular marker has yet been identified, the boundary between posterior maternal cytoplasm and a morphologically distinct nascent RB cannot be defined unambiguously at early stages of daughter formation. In this study, we therefore use the term “RB” to refer to the posterior compartment and connection between daughter parasites late during and after budding.

Increasing evidence indicates that the RB is not simply a terminal remnant of cell division but forms part of a dynamic system that maintains physical and functional continuity between parasites within the PV. An extensive F-actin network connects intravacuolar parasites through the RB and is required for parasite organization and material exchange (Periz et al., 2017a). Frénal et al. further demonstrated that intravacuolar tachyzoites are connected through their basal poles and the RB, allowing rapid exchange of soluble proteins between parasites and contributing to synchronized replication (Frenal et al., 2017). This functional connectivity depends on the unconventional myosins MyoI and MyoJ: deletion of either motor disrupts inter-parasite communication and replication synchrony (Frenal et al., 2017). Importantly, RBs are nevertheless still formed in the absence of MyoI or MyoJ, while the connection between the parasites and the RB is disrupted, demonstrating that RB formation can be uncoupled from functional connectivity (Frenal et al., 2017). This distinction is particularly important because it separates formation of the RB itself from its function as a conduit for inter-parasite exchange. The ability of the RB-associated network to mediate exchange between parasites raises the possibility that it may participate not only in the diffusion of soluble material but also in the redistribution of larger cellular components. Indeed, we previously demonstrated that maternal micronemal proteins, such as MIC2, are extensively recycled into daughter cells in an F-actin-dependent manner and can traffic through the intravacuolar network (IVN) (Periz et al., 2019). Together, these observations suggest that the cytoplasmic continuity provided by the RB and its associated F-actin network could contribute to the redistribution of maternal organelles during replication.

MyoF provides an additional link between myosin-dependent transport and organelle inheritance. MyoF is required for centrosome positioning and apicoplast inheritance, and its depletion causes enlargement of the RB accompanied by accumulation of micronemes and rhoptries (Jacot et al., 2013). Although this established a link between MyoF, RB morphology, and secretory-organelle distribution, whether the accumulated organelles represent maternally inherited material and whether they can subsequently be redistributed to daughter parasites remained unresolved. In addition, the actin-nucleating factors Formin-2 and Formin-3 contribute to organization of the intravacuolar F-actin network (Stortz et al., 2019; Tosetti et al., 2019). Together, these studies establish an actomyosin-dependent system connecting daughter parasites through the RB, but how this system contributes to the inheritance and redistribution of maternal organelles remains unclear.

F-actin could mediate this process either by serving as a track for myosin-dependent vesicular transport, as described for myosin VI in other eukaryotes (Frank et al., 2004), or through the intrinsic mobility and dynamic association of actin bundles themselves, which could transiently interact to promote vesicle exchange (Das et al., 2021; Khaitlina, 2014; Moore et al., 2021). In this study, we employ a dual-labeling strategy using the HaloTag system (Urh and Rosenberg, 2012) to investigate the fate of maternal organelles during T. gondii replication. Through fluorescence intensity-based analysis, we identify three distinct modes of organelle inheritance: inheritance of intact maternal organelles, expansion and division of pre-existing maternal organelles with incorporation of newly synthesized components, and degradation of maternal structures followed by de novo assembly. We further show that MyoF is required for efficient redistribution of maternal micronemes and rhoptries and that maternal secretory organelles accumulated in the RB after MyoF depletion can re-enter daughter parasites following restoration of MyoF. Together, these findings support a role for the RB as a dynamic compartment in the recycling and redistribution of maternal organelles.

Results

HaloTag-Based pulse chase labelling allows discrimination of recycled and de novo material

To evaluate HaloTag as a tool for distinguishing maternal from newly synthesized proteins during endodyogeny, two marker proteins with known fates were used: MIC2, which undergoes extensive recycling (Periz et al., 2019), and IMC1, primarily synthesized de novo (Ouologuem and Roos, 2014). Using CRISPR/Cas9, endogenous loci were tagged to generate MIC2-Halo and IMC1-Halo strains (Singer et al., 2023).

Both strains were sequentially labeled by incubation with membrane permeable dyes for 1h: maternal proteins were marked with Halo JF646 before invasion (M-MIC2, M-IMC1), and 24 hours later after replication, by Halo JF549 for 1h to label newly synthesized proteins (n-MIC2, n-IMC1) (Figure 1A). As Toxoplasma gondii replicates in an asynchronous manner, after 24 hours of replication, vacuoles containing 1, 2, 4, and 8 parasites can be observed in a single dish.

Discrimination between maternal and de novo material.

A) Dual staining scheme. Maternal material is labelled initially with Janila Fluor®-646, and the excess washed out. After replication of parasites de novo synthesised material is labelled with Janila Fluor®-549, allowing efficient discrimination of the two populations. B) Fluorescence intensity quantification of maternal (M) and de novo sythenetised (n) MIC2 and IMC1 during replication from stage 1 to 8. Magenta: M-MIC2, Rosa: M-IMC1, Green: n-MIC2, Dark Green: n-IMC1. C) Representative picture of MIC2-Halo parasites during replication (stage 1 to 8) stained for both maternal (M) and de-novo (n) MIC2. Magenta: M-MIC2, Green: n-MIC2. Maternal MIC2 is efficiently recycled into the daughters, while de novo MIC2 is formed after each replication cycle D) Representative picture of IMC1-Halo parasites during replication (stage 1 to 8) stained for both maternal (M) and de-novo (n) IMC1. Magenta: M-IMC1, Green: n-IMC1. In contrast to micronemes, maternal IMC is degraded, and only residual amount is detected after the first replication cycle. Three biological replicates were used for all analysis with a total of 300 IMC-1 Halo vacuoles and 260 MIC2-Halo vacuoles analysed; error bars are standard deviations, and the centre measurement of the graph bars is the mean. Panels C and D show maximum-intensity projections of Z-stack images. All scale bars = 1 µm.

Fluorescence intensity (FI) analysis across those replication stages (Figure 1B–D, S1) revealed a progressive decline of M-MIC2 and an increase in n-MIC2, confirming protein recycling. In contrast, M-IMC1 FI dropped from 100% to 14% between stages 1 and 2, with a corresponding rise in n-IMC1 to ∼80%, indicating predominant de novo synthesis. These distinct FI dynamics demonstrate HaloTag’s utility for resolving protein origin and support the de novo assembly of the IMC.

HaloTag profiling reveals three distinct patterns of protein inheritance

To further investigate protein inheritance during endodyogeny, endogenous HaloTag fusions were generated for a set of organelle-associated proteins: RON2 and ROP1 (rhoptries) (Besteiro et al., 2009; Striepen et al., 2001), GRA1 (dense granules) (Carruthers and Sibley, 1997), TIC20 (apicoplast) (van Dooren et al., 2008), SortLR (Golgi) (Sloves et al., 2012), ANKER1 (Barylyuk et al., 2020), GAPM1a (IMC) (Harding et al., 2019), and MyoA (glideosome) (Meissner et al., 2002) (Figure 2A, Figure S2).

Three recycling fates for the maternal organelles.

A) Overview of T. gondii organelles and molecular markers used for visualisation with endogenous Halo-Tags. B) Fluorescence intensity (FI) quantification of maternal (M) molecular marker listed in A. The fluorescence intensity variation during replication, allows to group them into three groups: 1) Efficient, almost quantitative recycling. 2) even distribution of maternal material and 3) almost exclusive de novo synthesis. Dark blue: RON2, Blue: ROP1, Pale blue: MIC2, Dark green: SortLR, Green: MyoA, Pale green: Tic20, Greenish white: ANKER1, Gold: GAPM1a, Yellow: IMC1. C) Representative picture of parasites expressing indicated Halo tagged proteins (stage 1 to 8) stained for maternal (M) proteins. Three biological replicates were used for all analysis with an average of 265 vacuoles analysed per protein; the graph indicate the mean value of FI calculated from the triplicate. Panels C shows maximum-intensity projections of Z-stack images. All scale bars = 1 µm.

Following the same sequential labeling strategy used for MIC2 and IMC1, fluorescence intensity (FI) of maternal proteins was quantified across the different replication stages (Figure 2B, C, S2C-G). Analysis revealed three distinct inheritance profiles:

Group 1: RON2 and ROP1 exhibited stable maternal FI, with minimal decline of FI within a single rhoptry, indicating recycling of whole organelles. In good agreement with this hypothesis, detailed analysis of M-RON2 inheritance demonstrated that the maternal rhoptries are separated during the distribution and that some daughter cells did not obtain maternal organelles after successive rounds of replication (Figure 3D).

Group 2: Proteins from diverse organelles (SortLR, ANKER1, TIC20, MyoA) showed a stepwise ∼50% reduction in maternal FI per division, suggesting that recycled, maternal material and de novo material end up in the same organelle

Group 3: IMC proteins (IMC1 and GAPM1a) displayed a sharp FI loss between the first and second replication cycles, consistent with predominant de novo synthesis of the IMC and degradation of maternal material.

Micronemes and rhoptries are recycled as whole organelles

Representative picture of MIC2-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) MIC2. Magenta: M-MIC2, Green: n-MIC2. Zoom windows allow to visualise isolated apical M-MIC2 signal for which the fluorescence intensity is quantified in C. B) Quantification of the inheritance of M-MIC2 by the daughter cells during replication (stage 1 to 8). A total of 260 vacuoles were analysed. All daughter cells inherited M-MIC2. C) Fluorescence intensity quantification of isolated M-MIC2 signal as illustrated in A. A total of 203 isolated M-MIC2 signals were analysed. D) Representative picture of RON2-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) MIC2. Magenta: M-RON2, Green: n-RON2. Zoom windows allow to visualise isolated apical M-RON2 signal for which the fluorescence intensity is quantified in F. E) Quantification of the inheritance of M-RON2 by the daughter cells during replication (stage 1 to 8). A total of 289 vacuoles were analysed. Not all daughter cells inherit M-RON2 (asterisks), best seen in later replication stages. F) Fluorescence intensity quantification of isolated M-RON2 signal as illustrated in D. A total of 215 isolated M-RON2 signals were analysed. Three biological replicates were used for all analysis; error bars are standard deviations, and the centre measurement of the graph bars is the mean. All p-values ≤ 0.001 (***), using two-tailed unpaired Student’s t-test. Panels A and D show maximum-intensity projections of Z-stack images. All scale bars = 1 µm.

Interestingly, MIC2 exhibited intermediate behavior, prompting further analysis to refine its classification. These patterns highlight distinct organelle-specific protein inheritance mechanisms in T. gondii.

Micronemes and rhoptries are recycled as intact organelles during Toxoplasma endodyogeny

Micronemes are known to comprise distinct subpopulations (Kremer et al., 2013), and prior work has shown that recycled and de novo MIC2 localize to separate microneme subsets (Periz et al., 2019). Based on these observations, we hypothesized that both micronemes and rhoptries are recycled as intact organelles, with new organelles formed independently via de novo synthesis.

To test this, we conducted detailed analyses of MIC2 and rhoptry inheritance. As expected, recycled (M-MIC2) and newly synthesized MIC2 (n-MIC2) were largely segregated into distinct micronemes with minimal colocalization (Figure 3A, S3)(Periz et al., 2019). Focusing on sparsely distributed micronemes outside the apical region (Figure 3A, boxed area), we measured fluorescence intensity (FI) of individual M-MIC2-positive micronemes at replication stages 2, 4, and 8 (Figure 3A, zoomed region; 3C). The relatively stable FI of these isolated structures supports whole organelle recycling, consistent with the FI profiles of rhoptries (Figure 2B, C).

Rhoptry analysis further corroborated this model. As with MIC2, M-RON2 and n-RON2 localized to distinct rhoptry populations (Figure S3), with divergence becoming more apparent at stage 8 (Figure 3D, Figure S4A). M-RON2 FI remained relatively constant through replication (Figure 3 F), while n-RON2 gradually increased from stage 2 onward (Figure S4B). Due to their low copy number (8–12 per tachyzoite), maternal rhoptries were separated and often absent in some daughter cells after successive divisions, consistent with stochastic whole-organelle inheritance illustrated by decrease in the average area of the M-RON2 fluorescence signal and the constancy of the total surface area (Figure S4C, D). As a consequence, a progressive decline in M-RON2-positive parasites per vacuole can be observed (Figure 3E, D; Figure S4A).

For the micronemes, even in larger parasitophorous vacuoles, all daughter parasites retained a mix of recycled and newly synthesized micronemes (Figure S5). At the 8-cell stage, maternal micronemes were distributed relatively evenly among daughter parasites (Figure S5A, B), and similar patterns were observed across independent vacuoles. This distribution is consistent with a non-random or regulated partitioning process, although the present data do not establish the underlying mechanism or imply the existence of a dedicated microneme-counting system. The simple probabilistic model for later-stage vacuoles further illustrates that highly even inheritance would be unlikely under unrestricted random segregation, but because the quantitative experimental dataset was obtained at the 8-cell stage, we interpret this analysis as supportive rather than definitive evidence for regulated partitioning. Together, these results support inheritance of micronemes and rhoptries as intact organelles, with de novo organellogenesis acting in parallel to replenish the secretory-organelle complement of daughter cells.

Golgi inheritance involves coordinated expansion and partitioning of maternal and de novo proteins

Next, we analyzed SortLR, a Golgi marker, using dual HaloTag labeling. Maternal SortLR (M-SortLR) fluorescence declined by ∼50% per replication round, while newly synthesized SortLR (n-SortLR) increased proportionally, without abrupt loss (Figure 4A–B). This gradual dilution contrasts sharply with the rapid decay seen in Group 3 proteins.

Expansion of the mother organelle for equivalent sharing to the daughter.

A) Representative picture of SortLR-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) SortLR. Magenta: M-SortLR, Green: n-SortLR. B) Fluorescence intensity quantification of M-SortLR and n-SortLR illustrated in A. Magenta: M-SortLR, Green: n-SortLR. A total of 251 vacuoles were analysed. C) Quantification of the inheritance of M-SortLR by the daughter cells during replication (stage 1 to 8). All daughter cells inherited M-SortLR. A total of 100 vacuoles were analysed. D) Quantification of the signal area of M-SortLR during replication. A total of 100 vacuoles were analysed. The total surface of M-SortLR increase at each replication by about a factor 2. Three biological replicates were used for all analysis; error bars are standard deviations, and the centre measurement of the graph bars is the mean. All p-values ≤ 0.001 (***), using two-tailed unpaired Student’s t-test. Panels A shows maximum-intensity projections of Z-stack images. All scale bars = 1 µm.

M-SortLR and n-SortLR exhibited strong colocalization across replication stages, and all daughter cells retained M-SortLR (Figure 4C). The total Golgi area expanded from ∼2 µm² to ∼10 µm² by stage 8 (Figure 4D), consistent with coordinated Golgi elongation and medial fission, as described previously (Pelletier et al., 2002). These results demonstrate that the Golgi undergoes duplication, incorporating both maternal and de novo proteins during its growth and partitioning. Similar signal evolution was observed for the ER, MyoA and apicoplast (Figure S2, S6). Thus, Group 2 organelles follow a distinct inheritance mode, involving regulated expansion, integration of new material, and equal distribution to progeny—distinct from the whole-organelle recycling of rhoptries/micronemes and the complete turnover seen in IMC proteins.

Maternal IMC proteins are degraded during replication

Group 3 proteins, typified by IMC components such as GAPM1a and IMC1, exhibit a sharply distinct inheritance profile, characterized by rapid loss of maternal fluorescence during replication. Using HaloTag-based dual labeling, we observed a ∼90% reduction in maternal GAPM1a signal by stage 2, with minimal retention through later stages (Figure 5A-C), indicative of degradation rather than recycling. M-GAPM1a and n-GAPM1a did not colocalize, and residual maternal signal localized transiently at the posterior pole before disappearing entirely (Figure 5D). This posterior accumulation occurs in the region that gives rise to the RB; however, because no marker uniquely distinguishes the nascent RB from the posterior maternal compartment, we cannot determine precisely when degradation occurs relative to RB formation.

Degradation of the inner membrane complex.

Representative picture of GAPM1a-Halo during replication (stage 1 to 8) stained for both the maternal (M) and de novo (n) GAPM1a. B) Fluorescence intensity quantification of M-GAPM1a and n-GAPM1a illustrated in A. A total of 265 vacuoles were analysed. C) Quantification of the inheritance of M-GAPM1a by the daughter cells during replication (stage 1 to 8). After replication, M-GAPM1a is not visible in daughter cells. A total of 265 vacuoles were analysed. *The calculation was performed using single stage parasite intensity to set up the exposure as performed for all the other organelles. Under those conditions the remaining signal of the maternal GAPM1a is below the detection level and cannot be observed. D) GAPM1a is degraded in the RB. Representative images of GAPM1a-Halo parasites at different stages of daughter cell development. M-GAPM1a collapses toward the forming residual body, where the signal disappears after completion of replication, indicating it’s degradation. E) Time series of IMC1-Halo parasites during the first replication. Parasite were stained for the maternal (M) IMC1 prior invasion. Three regions of the parasites were analysed: the apical (green ROI), the cytoplasmic (blue ROI) and the basal (magenta ROI). F) Fluorescence intensity analysis of the three regions defined in E. The curve and analysed ROI share the same colour code. Green: Apical, Blue: Cytoplasm, Magenta: Basal. After the accumulation of the mother IMC at the basal pole, the FI of M-IMC1 decrease without redistributing to any other region. Three biological replicates were used for all analysis; error bars are standard deviations, and the centre measurement of the graph bars is the mean. All p-values ≤ 0.001 (***), using two-tailed unpaired Student’s t-test. Panels A and D show maximum-intensity projections of Z-stack images. Panel E displays single optical plane images acquired during live imaging. All scale bars = 1 µm.

Live-cell imaging of IMC1 further supported this model: maternal IMC1 (M-IMC1) condensed at the posterior pole during daughter budding, followed by a marked fluorescence drop, without redistribution to other regions (Figure 5E, F and Figure S7, Video S1). This signal loss was replication-specific, as non-dividing parasites retained stable M-IMC1 levels (Figure S7D). Quantification across regions revealed a transient increase at the posterior, followed by complete disappearance, consistent with degradation (Figure 5F, Figure S7C).

Interestingly, DCs IMCs exhibited brighter fluorescence than the maternal IMC (Figure S7E, F), especially during early stages of elongation (Figure S7G, H). FI peaked when DCs reached ∼3 μm in length, suggesting the IMC is fully assembled before emergence and subsequently unfolds without requiring additional material. This preformation explains the absence of maternal IMC recycling. These findings align with earlier observations that IMC components are synthesized de novo during elongation (Ouologuem and Roos, 2014).

Whole-organelle inheritance of micronemes and rhoptries occurs via similar recycling pathways

Our data reveal distinct spatial behaviors among the analysed organelle markers. Maternal microneme material can enter the posterior RB-associated compartment and subsequently redistribute to daughters, whereas most maternal RON2 is incorporated into daughter rhoptries before mother-cell collapse and formation of a morphologically distinct RB. In contrast, we did not observe comparable accumulation of the analysed Golgi, ER, or apicoplast markers in the RB during inheritance. Thus, the present experiments support participation of the RB in secretory-organelle redistribution but do not establish obligatory RB transit for all inherited micronemes or rhoptries.

To investigate this further, we performed live-cell imaging of rhoptry and microneme inheritance (Figure 6A, B, Video S2, S3). MIC2-Halo parasites co-expressing IMC1-YFP showed maternal MIC2 (M-MIC2) entering the posterior RB-associated compartment (Figure 6A, white arrows). After endodyogeny was completed, M-MIC2 present in this compartment was subsequently redistributed to the apical region of daughter cells (Figure 6A, 11:30, 16:00 h). These observations demonstrate that maternal micronemes can transit through the RB, which can act as a temporary reservoir from which they are subsequently redistributed to daughter parasites (Periz et al., 2019).

Micronemes and rhoptries are recycled similarly, but differ in timing and location.

A) Time-lapse of MIC2-Halo over two replication cycles. Maternal MIC2 (M-MIC2) is stained, and daughter cell formation is visualized via IMC1-YFP. M-MIC2 is transported via the residual body (RB) (white arrows; see Video S2). B) Time-lapse of RON2-Halo parasites with IMC-YFP. Maternal RON2 (M-RON2) is transported before mother cell collapse and RB formation (white arrows). C) Representative images showing maternal organelles associating with F-actin and the RB. Insets highlight colocalization of maternal proteins and F-actin. D) Quantification of M-MIC2 and M-RON2 colocalization with F-actin. Data from three biological replicates, a total of 534 RON2-Halo and 342 MIC2-Halo vacuoles were analysed; bars represent means ± SD. Panel A and B display single optical plane images acquired during live imaging. Panel C show single Z plane of Z-stack images. Scale bars = 1 µm.

For rhoptries, we examined the inheritance of maternal RON2 (M-RON2) in RON2-Halo parasites co-expressing IMC1-YFP. In approximately 90% of parasites undergoing replication, M-RON2 was integrated into daughter rhoptries prior to mother-cell collapse and formation of a morphologically distinct RB (Figure 6B, 4:15-4:30 h and 10:30-10:45 h). M-RON2 was only occasionally detected in the RB-associated compartment. Thus, while both maternal micronemes and rhoptries can associate with the RB, their trafficking differs in timing, with most maternal RON2 being incorporated into daughter rhoptries before formation of a morphologically distinct RB. This observation also leaves open the possibility that a substantial fraction of maternal rhoptries is transferred directly from the mother into developing daughters. To test whether recycled rhoptries associate with F-actin, as previously reported for micronemes (Periz et al., 2019), we stably integrated the F-actin marker Cb-Emerald (Periz et al., 2017b) in the RON2-Halo line (Figure 6C, D). While m-MIC2 displayed clear F-actin– associated movement along filaments in the residual body (Figure 6C), colocalization of M-RON2 with Cb-Emerald in the residual body was infrequent. On the other hand, both secretory organelles were found to be associated with cytoplasmic F-actin (Figure 6D).

In summary, both rhoptries and micronemes can be inherited as intact organelles during T. gondii replication. Their distinct timing and localization indicate that related cytoskeletal trafficking mechanisms may operate through more than one spatial route during daughter-cell formation.

Given the F-actin dependency of microneme recycling demonstrated previously (Periz et al., 2019), we hypothesized that a conserved apicomplexan myosin contributes to this process. MyoF, a class XXII myosin conserved across Apicomplexa (Jacot et al., 2013), has been implicated in centrosome positioning, apicoplast inheritance, actin organization, and vesicular/endomembrane dynamics (Carmeille et al., 2021; Heaslip et al., 2016; Kellermeier and Heaslip, 2024). Interestingly, Jacot et al. (2013) previously reported enlargement of the residual body and accumulation of micronemes and rhoptries following MyoF depletion. The dual-labeling approach used here allows us to distinguish whether the accumulated material derives from the initial maternal protein pool or from proteins synthesized during subsequent replication cycles.

We endogenously Halo-tagged MIC2, RON2, and SortLR in the auxin-inducible degron line MyoF-mAID (Carmeille et al., 2021) and performed dual HaloTag labelling to differentiate maternal (M) from de novo (n) protein pools. Live imaging revealed that MyoF depletion resulted in accumulation of maternal MIC2 and RON2 in the RB-associated compartment during replication (Figure 7A, B). As Toxoplasma replicates asynchronously, different replication stages coexist within the same dish after 24 h. The second labeling step marks proteins synthesized since the beginning of the experiment, allowing discrimination between proteins present in the initial mother parasite and proteins synthesized during subsequent replication cycles. During successive rounds, proteins synthesized in earlier cycles can themselves become inherited material. Under MyoF depletion, these pools accumulate in the RB-associated compartment rather than being efficiently redistributed to daughters (Figure 7). Immunofluorescence with α-AMA1, α-MIC4, and α-MIC8 confirmed that this phenotype affects multiple micronemal proteins (Figures S8, S9). Approximately 90% of vacuoles exhibited RB-associated accumulation of maternal material upon auxin treatment (Figure 7D), demonstrating that MyoF is required for efficient redistribution of maternal micronemes and rhoptries. Interestingly, although our time-lapse analysis indicated that rhoptries only occasionally traffic through the RB, both maternal RON2 and MIC2 accumulate in the RB following MyoF depletion. This shared phenotype indicates that MyoF contributes to the redistribution of both organelle populations, even though their precise trafficking routes and timing may differ.

Myosin-F drives F-actin mediated recycling of maternal organelles via the residual body.

A) Effect of MyoF knockdown (KD) on maternal microneme inheritance. MyoF-mAID MIC2-Halo parasites were labeled and grown ± auxin for 24h. Magenta: maternal MIC2 (M-MIC2), Green: newly synthesized MIC2 (n-MIC2). Without MyoF, M-MIC2 accumulates in the residual body (RB) instead of being passed to daughter cells (Stage 1–2), with increasing accumulation over replication cycles (Stage 2–8). B) Effect of MyoF KD on maternal rhoptry inheritance. MyoF-mAID RON2-Halo parasites show M-RON2 retention in the RB, mirroring the pattern seen with MIC2. C) Effect of MyoF KD on maternal Golgi inheritance. MyoF-mAID SortLR-Halo parasites show no significant RB accumulation of M-SortLR, unlike MIC2 and RON2. D) Quantification of vacuoles showing RB accumulation. White: control; Gray: MyoF-KD. Three biological replicates were used; a total of 675 Ctrl and 743 KD vacuoles were analyzed for MIC2-Halo, 645 Ctrl and 594 KD for RON2-Halo and 795 Ctrl and 619 KD for SortLR-Halo, bars show means ± SD. All p-values ≤ 0.001 (***), using two-tailed unpaired Student’s t-test. Panels A, B and C show maximum-intensity projections of Z-stack images. Scale bars = 1 µm

In contrast, Golgi inheritance (marked by SortLR) showed no comparable RB accumulation after MyoF depletion (Figure 7C). Neither accumulation in the RB nor separation of maternal and de novo SortLR signals was observed, consistent with previous reports that Golgi duplicates and partitions independently (Pelletier et al., 2002)), although Golgi fragmentation was observed as described previously.

To confirm that this phenotype results directly from MyoF depletion and not broader disruption of endomembrane architecture, we performed a rescue experiment. MIC2-Halo MyoF-mAID parasites were cultured in auxin for 24 hours to induce M-MIC2 retention in the RB-associated compartment, followed by auxin washout and continued replication (Figure 8A). Following auxin washout, M-MIC2 was redistributed to daughter parasites during subsequent replication cycles, although occasionally in an uneven manner (Figures 8A, 8C, and S10A). To determine whether this redistribution correlated with restoration of MyoF expression, we monitored MyoF recovery by IFA using the HA epitope fused to MyoF-mAID (Figure 8B). Following auxin washout, MyoF expression was restored. Quantification revealed a strong association between detectable MyoF and redistribution of maternal micronemes (Figure S10B), supporting a requirement for MyoF in efficient retrieval/redistribution of accumulated maternal MIC2.

The residual body (RB) functions as a recycling center, not a dead end.

To assess the fate of material accumulated in the RB, auxin chase experiments were performed: 24h with auxin followed by 24h without. A) Representative images of MyoF-mAID MIC2-Halo parasites after 48h in control (no auxin), continuous auxin (Aux), or auxin washout (Aux washed). Magenta: M-MIC2, Green: n-MIC2. Continuous auxin led to M-MIC2 accumulation in the RB, while auxin washout enabled redistribution. B) Representative images of MyoF-mAID MIC2-Halo parasites after 48h in control (no auxin), continuous auxin (Aux), or auxin washout (Aux washed). Magenta: M-MIC2, Green: α-HA (MyoF). In absence of auxin MyoF is visible, induction with auxin deplete MyoF led to M-MIC2 accumulation in the RB, while auxin washout enabled MyoF production and M-MIC2 redistribution. C) Quantification of vacuoles showing normal, accumulated, or redistributed M-MIC2. White: control; Gray: MyoF-KD; Blue: MyoF-KD auxin chase. D) Time-lapse without auxin: M-MIC2 briefly passes through the RB (white arrows). A total of 170 Ctrl, 186 KD and 179 auxin chased vacuoles were analyzed. E) Time-lapse with auxin: sustained M-MIC2 accumulation in the RB throughout replication. F) Time-lapse after auxin washout: initial M-MIC2 accumulation in the RB resolves by the second replication cycle, with redistribution observed. Three biological replicates were used; bars show means ± SD. All p-values ≤ 0.001 (***), using two-tailed unpaired Student’s t-test. Panels A and B show maximum-intensity projections of Z-stack images. Panel D, E and F display single optical plane images acquired during live imaging. Scale bars = 1 µm.

Live-cell imaging confirmed these dynamics (Videos S4-S6). In untreated parasites, M-MIC2 entered the RB-associated compartment and was later redistributed to daughter cells (Figure 8D, Video S4). Under auxin treatment, M-MIC2 accumulated post-budding and failed to redistribute efficiently (Figure 8E; Video S5). Following auxin washout, imaging was initiated after removal of auxin at 24 h; time 0 in Figure 8F/Video S6 therefore corresponds to the start of post-washout imaging. Redistribution was observed during the subsequent replication cycle rather than immediately after MyoF re-expression (Figure 8F; Video S6). This delay suggests that efficient redistribution requires sufficient recovery of MyoF function and/or progression into the next replication cycle.

To assess whether microneme degradation occurred during retention within the residual body, the fluorescence intensity of individual maternal micronemes was compared between control vacuoles and vacuoles subjected to MyoF depletion followed by auxin washout (Figure S10C, D). No significant difference in microneme fluorescence intensity was detected between control micronemes and micronemes that had remained within the RB for 24 h prior to redistribution. These results indicate that micronemes recovered after retention within the RB do not show a detectable loss of maternal MIC2 signal. Together, these findings establish that MyoF is required for the efficient recycling and redistribution of maternal micronemes and rhoptries. They further demonstrate that the RB can function as a dynamic compartment from which retained maternal organelles can be redistributed to daughter parasites.

Discussion

Use of dual labelling to differentiate maternal and de novo material

Understanding the complexity of organelle inheritance in Toxoplasma gondii requires tools capable of resolving both the spatial and temporal dynamics of protein trafficking. Conventional single-labelling approaches have been limited to static snapshots, obscuring key processes such as organelle biogenesis, recycling, and selective degradation. To overcome these challenges, we employed a dual HaloTag pulse-chase labelling strategy to distinguish maternally inherited from de novo–synthesized proteins. This approach, previously underutilized in apicomplexan parasites (Koreny et al., 2023; Periz et al., 2019), allowed us to resolve protein fates across replication cycles and map distinct inheritance routes for multiple organelles (Figure 9A).

Schematic summaries.

A: Summary of the three fates of the maternal organelles. Group 1: Efficient, intact organelle recycling. De novo organelles are generated independently of the maternal organelle but in a similar location. Both de novo and maternal are distributed to the daughter cells. The Fluorescence intensity of the maternal organelle is relatively stable. Group 2: Even distribution of maternal material through organelle expansion with the insertion of the de novo material in the mother organelle. The Fluorescence intensity of the maternal organelle is divided by two at each replication step but their signal surface is increased by two. Group 3: Almost exclusive de novo synthesis. The de novo organelle is generated independently of the maternal organelle and in a different location. The maternal organelle is not clearly observable after a cycle of replication. The fluorescence intensity of the maternal organelle drastically drops after the first replication. Magenta: maternal organelle, Green: de novo organelle, Yellow: Colocalization between maternal and de novo. B: Summary of the MyoF regulated inheritance. The maternal and the de novo micronemes are generated independently. In presence of MyoF (F1), the daughter cells form a chimera possessing both maternal and de novo micronemes, which are distributed in a relatively equal manner even after multiple replication cycles. In absence of MyoF, (F2) the inherited micronemes accumulate in the residual body and the daugthers cells are mostly composed of de novo micronemes. If the depletion of MyoF is maintained (F2.1) more and more inherited micronemes accumulate inside the residual body without important degradation. In opposition, if the expression of the MyoF is reestablished by the auxin removal (F3), the inherited micronemes stuck inside the residual body are redistributed but in a less equal manner than usually observed for control.

The residual body as a dynamic compartment for organelle recycling

Our findings reaffirm the emerging view of the RB as an active, multifunctional structure rather than a passive cytoplasmic remnant (Frenal et al., 2017; Periz et al., 2017a). First described over 50 years ago (Sheffield and Melton, 1968), the RB was long believed to serve as a repository for discarded material, often called a “waste bin” (Nishi et al., 2008). Our results extend this view by showing that maternal secretory organelles can enter the RB and subsequently be redistributed to daughter parasites in a MyoF dependent manner (Figure 9B). This is particularly evident for micronemes: live-cell imaging demonstrated that maternal MIC2 can transit through the RB, while MyoF depletion causes maternal micronemes to accumulate within this compartment. Restoration of MyoF allows this accumulated material to re-enter daughter parasites (Figure 8), demonstrating that the RB is not simply a terminal or degradative compartment but can serve as a temporary reservoir during organelle recycling. Rhoptry inheritance reveals a more complex picture. Although maternal RON2 was occasionally detected in the RB and accumulated prominently in this compartment following MyoF depletion, most maternal RON2 was incorporated into daughter rhoptries before mother-cell collapse and formation of a morphologically distinct RB. Thus, our data support a role for the RB in secretory-organelle redistribution but do not imply that transit through the RB is obligatory for all inherited micronemes or rhoptries. Direct transfer from the maternal parasite into developing daughters may operate in parallel, particularly for rhoptries. Importantly, the shared accumulation of maternal MIC2 and RON2 in the RB following MyoF depletion indicates that MyoF contributes to the redistribution of both organelle populations, even though their precise spatial routes and timing may differ.

This active role is further supported by the dependence of RB-mediated recycling on F-actin and the class XXII myosin MyoF, which we show to be essential for retrieval of maternal MIC2 and RON2 but dispensable for Golgi inheritance although we noticed a fragmentation of the Golgi, which has been described to depend on MyoF (Carmeille et al., 2021). The RB is also critical for cytoplasmic continuity across parasites, synchronizing replication and facilitating inter-parasite protein exchange (Frenal et al., 2017; Muniz-Hernandez et al., 2011; Periz et al., 2017a). Importantly, RB formation and functional connectivity can be uncoupled: residual bodies are still formed in the absence of MyoI or MyoJ, whereas parasites lose their connection to the RB and inter-parasite communication is disrupted (Frenal et al., 2017). This distinction suggests that formation of the RB alone is insufficient for its exchange functions and that actomyosin-dependent connectivity is required to establish a functional route between parasites.

Together with the F-actin network extending through the RB (Periz et al., 2017a), these observations support a model in which the RB forms part of a regulated system for exchange and redistribution of cellular material within the vacuole rather than simply representing a passive leftover of daughter-cell formation. Given these parallels, the RB may be functionally analogous to the mammalian midbody remnant, which regulates post-mitotic signaling and intracellular trafficking(Kuriyama et al., 2025; Peterman and Prekeris, 2019).

Fate determination: inheritance or degradation?

An important question raised by our study is how the parasite determines whether maternal proteins are recycled or degraded. Our results indicate that this decision is selective rather than stochastic (Figure 9). For instance, micronemes recovered after retention in the RB following MyoF depletion retain maternal MIC2 fluorescence intensities comparable to control micronemes (Figure S10C, D). This indicates that the recovered organelles do not show detectable loss of maternal MIC2 signal. However, because this analysis measures individual recovered micronemes, it cannot exclude degradation of a fraction of the total microneme population during RB retention.

In contrast, IMC proteins like GAPM1a and IMC1 are lost within a single cycle and show no evidence of recycling. Both proteins transiently accumulate at the posterior pole before their maternal signal disappears raising questions about the underlying mechanism.

Although T. gondii encodes homologs of lysosomal and autophagic degradation machinery (Besteiro et al., 2011; Smith et al., 2021; Thaprawat et al., 2025), none have been localized to the RB or directly linked to IMC degradation. However, ubiquitin ligases and proteasomal components have been identified in the RB proteome (O’Shaughnessy et al., 2023; Que et al., 2002). IMC proteins are present in the ubiquitinated proteome (Silmon de Monerri et al., 2015), and recent work showed that deletion of the kinase ERK7 prevents sequestration of the E3 ligase CSAR1, resulting in aberrant degradation of conoids (O’Shaughnessy et al., 2023). These findings raise the possibility that selective ubiquitination and regulated proteolysis contribute to determining the fate of maternal proteins during daughter-cell formation. Whether these processes occur directly within the RB or within the posterior maternal compartment that subsequently forms the RB remains to be determined.

A call to reassess recycling factors in Toxoplasma gondii

The recognition of the residual body (RB) as a key site for organelle recycling has fundamentally revised our understanding of intracellular trafficking in T. gondii (Periz et al., 2019; Tosetti et al., 2019). Earlier models emphasized de novo synthesis of organelles and proposed a repurposing of classical endocytic machinery for secretory functions (Tomavo, 2014). However, our HaloTag-based pulse-chase studies show that secretory organelles such as micronemes and rhoptries are extensively recycled during daughter-cell formation, with the RB serving as an important site for their redistribution in a MyoF-dependent manner.

This shift in our understanding calls for a systematic re-evaluation of trafficking factors traditionally implicated in protein targeting, vesicle transport, and organelle biogenesis. Many of these, particularly Rab-GTPases, SNAREs, and dynamins were previously studied without consideration of recycling pathways. Notably, accumulation of microneme material in the RB has been observed upon perturbation of several trafficking regulators, including Rab5A (Kremer et al., 2013), ArlX3 (Klinger et al., 2024), SORTLR (Sloves et al., 2012), VPS8 (Morlon-Guyot et al., 2018), and VPS11 (Morlon-Guyot et al., 2015). Yet, whether these phenotypes reflect a block in secretion, synthesis, or recycling remains unclear. To resolve this, a targeted reanalysis of trafficking factors using our dual-labelling HaloTag assay combined with the splitCas9 system (Li et al., 2022) would allow high-throughput functional screening to specifically distinguish defects in recycling from those in de novo biogenesis, offering a path to redefine the roles of classical and lineage-specific trafficking regulators in T. gondii.

Material and Methods

Parasite culture and genetic manipulation

Growth and generation of transgenic T. gondii

T. gondii tachyzoites from the RH strain and derived lines, including RH Δku80/TATi and RH Δku80/Tir1, were maintained at 37 °C with 5% CO₂ in human foreskin fibroblasts (HFFs; ATCC, SCRC 1041) cultured in Dulbecco’s Modified Eagle Medium (DMEM; Sigma, D6546) supplemented with 10% fetal bovine serum (FBS; BioSell, FBS.US.0500), 4 mM L-glutamate (Sigma, G7513), and 20 μg/mL gentamycin (Sigma, G1397) as previously described (Gras et al., 2019).

Generation of transgenic parasites

New strains were generated using CRISPR/Cas9 as previously described (Li et al., 2022). Guide RNAs (gRNAs) targeting the regions of interest were designed using EuPaGDT (Alvarez-Jarreta et al., 2024). All gRNA and primer sequences are listed in Supplementary Table 1. Briefly, gRNA oligos were annealed, ligated into the Cas9-YFP vector, and verified by sequencing (Eurofins Genomics). Repair templates were generated by PCR amplification of the Halo or YFP tag flanked by 50 bp homology arms to the target gene, using Q5 High-Fidelity DNA Polymerase (New England BioLabs). PCR products were purified with a PCR purification kit (Blirt, EM26.1). Tachyzoites were mixed with the repair template and 10 µg of the corresponding Cas9 vector, and transfected using the Amaxa 4D-Nucleofector system (Lonza, AAF-1003X). Transfected parasites were allowed to invade fresh HFFs and replicate for 48 h. Following manual egress and filtration through a 3 μm filter, Cas9-YFP-expressing parasites were enriched by FACS (FACSAria III, BD Biosciences) and sorted into 96-well plates. Correct integration of the repair template was confirmed by PCR.

Labelling and characterization

Maternal and de-novo protein discrimination

Fresh tachyzoites expressing Halo-tagged reporters were mechanically egressed, filtered through a 3 μm filter, and resuspended in cold DMEM containing a membrane-permeable Halo dye, Janelia Fluor® 646 (1:1000, Promega), for 1 h. Parasites were centrifuged at 2,500 rpm for 5 min and washed three times with fresh medium to remove unbound dye. Parasites were then seeded onto HFF-covered Ibidi live-cell dishes for overnight replication. After replication, a second labeling was performed with a new membrane-permeable Halo dye, Janelia Fluor® 549 (1:1000, Promega) for 1 h at 37 °C, followed by three washes before imaging.

Fluorescence intensity across the replication

Parasites were labeled as described above and allowed to replicate for 24 h on HFF-coated Ibidi live-cell dishes. Approximately 15 fields of view were imaged using Z-stacks spanning 3 μm centered on the vacuoles (Figure S11A). For each replication stage (1, 2, 4, and 8 parasites per vacuole), individual tachyzoites were sampled across multiple vacuoles.

Maximum-intensity projections were generated from non-deconvolved images. Fluorescence intensity (FI) was quantified as the maximum gray value measured within regions of interest (ROIs) drawn on individual tachyzoites from each vacuole stages (Figure S11B). ROIs excluded overlapping parasites, neighboring vacuoles, and regions with atypical signal intensity (Figure S11C). For each biological replicate, up to 25 tachyzoites per replication stage were analyzed, and the mean FI value was calculated for each stage. The highest mean FI observed among the stages within a replicate was defined as 100%, and FI values for the other stages were expressed relative to this maximum. Relative FI values were then averaged across three independent biological replicates. Data are presented as mean ± SD. A total of 272, 286, 251, 276, 223, 284, 265, 300 and 260 vacuoles were analysed for RON1, ROP1, SortLR, MyoA, TIC20, ANKER1, GAPM1a, IMC1 and MIC2 respectively.

Fluorescence intensity of the daughter cells vs new mother cells

Images were processed as described above. Fluorescence intensity was quantified as the maximum gray value measured within regions of interest (ROIs) drawn on individual daughter cells and on newly formed mother cells. To reliably discriminate newly formed mother IMC from daughter IMC (both in the same channel colour, Jan. 549), the channel corresponding to the initial maternal IMC (Jan. 646) was intentionally overexposed, allowing unambiguous identification of the new mother cell IMC signal. ROIs were drawn on non-overlapping regions and excluded areas containing overlapping parasites, neighboring vacuoles, or atypical signal intensity, following the same exclusion criteria used for replication-stage measurements (Figure S11D). Quantification was performed on maximum-intensity projections generated from non-deconvolved Z-stacks. Data are presented as mean ± SD from vacuoles pooled across three independent biological replicates.

Measurement of the signal area of the fluorescence

To quantify the average and total fluorescent area, images were analyzed in Fiji. Up to 25 isolated vacuoles per stage (1 to 8) were cropped and analyzed individually. A threshold was applied to the maternal signal, and both average and total signal area were measured. Three independent biological replicates were performed, and values are reported as mean ± SD.

Quantification of percentage of parasite per vacuole with maternal protein

Images were analyzed in Fiji to determine the proportion of parasites per vacuole with maternal protein signal. Up to 25 vacuoles per stage (1 to 8) were cropped and analyzed individually. Vacuoles where all parasites retained maternal signal were scored as 100%. If some parasites lacked signal, the percentage of maternal-positive parasites was calculated. The experiment was performed in three independent biological replicates. Values are reported as mean ± SD.

Live replication assay

Parasites were labeled with Janelia Fluor® 646 (1:1000, Promega) for 1 h and washed three times before transfer to Ibidi live-cell dishes covered with HFFs. Parasites were allowed to invade for 1–2 h to obtain ∼10 parasites per field of view. Excess parasites were removed by washing three times. Imaging was performed on a Leica DMI8 microscope at 37 °C in 5% CO₂. Images were acquired every 15–30 min for 12 h using minimal laser power and exposure. Experiments were performed in three biological replicates and analyzed in Fiji.

Colocalization assays: ANKER1-Halo with HDEL-GFP

Parasites expressing ANKER1-Halo were transiently transfected with an HDEL-GFP plasmid to label the ER. After overnight replication in HFF-coated Ibidi dishes, parasites were labeled with Janelia Fluor® 646 (1:1000, Promega) for 1 h and washed three times prior to imaging. Z-stacks of HDEL-GFP-positive parasites were acquired. The proportion of vacuoles displaying ER localization of ANKER1-Halo was assessed manually in Fiji/ImageJ by inspection of full Z-stacks and maximum-intensity projections. In addition, the degree of colocalization between ANKER1-Halo and HDEL-GFP was quantified using Pearson’s correlation coefficient calculated with the Coloc2 plugin in Fiji/ImageJ. Pearson analysis was performed on 30 individual parasites obtained from three independent biological replicates (10 parasites per replicate). The mean Pearson’s correlation coefficient was 0.92 ± 0.04 (mean ± SD).

Colocalization assays: MIC2/RON2 with Cb-Emerald

For analysis of association with the F-actin network, MIC2 or RON2 was endogenously tagged and labeled in a strain stably expressing Cb-Emerald. Freshly egressed parasites were labeled with Janelia Fluor® 646 (1:1000, Promega) for 1 h, washed, and seeded onto HFF-coated live-cell dishes. After 24 h of replication, parasites were fixed in 4% paraformaldehyde and imaged. Z-stacks of vacuoles displaying a clearly defined F-actin network were acquired, and the association of MIC2- or RON2-positive structures with F-actin was assessed in both the parasite cytoplasm and the RB. Analysis was performed manually in Fiji/ImageJ by inspection of full Z-stacks and maximum-intensity projections. The RB was defined by the presence of thick bundles of F-actin located at the basal pole of the parasites and forming a continuous structure connecting parasites within the vacuole, consistent with the characteristic morphology of the RB. Cytoplasmic F-actin was defined as all actin signal located within the parasite body excluding this RB-associated region. Three independent biological replicates were analyzed, and data are reported as mean ± SD.

Quantification of the maternal microneme distribution

To assess the distribution of maternal micronemes at stage 8, parasites were labeled as described in “Maternal and de novo protein discrimination.” Twenty-five stage 8 vacuoles were selected. Using the de novo MIC2 signal, the outline of each tachyzoite was traced, and the number of maternal MIC2-positive parasites was counted. Tachyzoites were numbered 1 to 8 from left to right. Three biological replicates were performed. Data are reported as mean ± SD. Images were analyzed in Fiji.

Probability of maternal microneme inheritance

To explore whether the relatively even distribution of maternal micronemes could arise through stochastic segregation alone, we previously considered a simplified model allocating 64 micronemes among 32 daughter tachyzoites. Because the experimental quantification presented in Figure S5 was performed at the 8-cell stage, this 32-cell calculation is not used here as direct evidence for regulated segregation. A rigorous test of regulated partitioning will require a null model matched to the experimentally quantified stage and, ideally, direct quantification of later-stage vacuoles. P(0≤X≤3)=e−λ(1+λ+λ22!+λ33!)Xλ=6432=2P(0≤X≤3)≈0.85710.857132≈0.00720.007220≈1.4×10−43 The relatively even maternal MIC2 distribution observed at the 8-cell stage is therefore described as an empirical observation rather than evidence for an established microneme-counting or balancing mechanism.

Phenotypic assays

Induction of MyoF KD

MyoF-mAID parasites were genetically modified to endogenously tag MIC2, RON2 and SortLR. As previously described for MyoF mAID parasites (Carmeille et al., 2021), MyoF-mAID MIC2/RON2 or SortLR-Halo parasites were induced +/- auxin for 4h prior to labelling and experiments.

Quantification of the accumulation of protein in the residual body

Parasites were treated ± auxin for 4 h and labeled as described above. After 24 h replication on live-cell dishes, ∼100 vacuoles were imaged across ∼15 fields of view. The percentage of vacuoles showing accumulation of material in the residual body was calculated. The experiment was performed with three independent biological triplicates. The values were then expressed as the mean values of the three independent experiments ± SD. Images were analysed via Fiji.

Auxin chased experiment

Parasites were induced ± auxin for 4 h prior to labeling as described above in “Maternal and de novo protein discrimination”. Following transfer to live-cell dishes, parasites were allowed to replicate for 48 h in the continued presence or absence of auxin before imaging. For the auxin chase condition, parasites were first maintained in auxin-containing medium for 24 h. The medium was then removed, and dishes were washed three times with fresh medium to ensure complete auxin removal before addition of fresh medium lacking auxin. Parasites were subsequently allowed to replicate for an additional 24 h prior to imaging.

To assess recovery of MyoF expression following auxin washout, parallel samples were fixed in 4% paraformaldehyde and processed for immunofluorescence using an anti-HA antibody to detect the MyoF-mAID-HA fusion protein. Vacuoles were categorized according to the presence or absence of detectable MyoF signal and correlated with the corresponding microneme distribution phenotype.

Approximately 15 fields of view were imaged per condition, corresponding to ∼100 vacuoles. The percentage of vacuoles displaying microneme accumulation within the residual body was determined. In addition to the “accumulated” and “normal” phenotypes, a third category termed “uneven distribution” was defined for vacuoles in which micronemes were no longer retained within the residual body but had not yet reached the homogeneous distribution observed in control parasites. The frequency of each phenotype was quantified. Experiments were performed in three independent biological replicates, and values are presented as mean ± SD. Image analysis was performed using Fiji/ImageJ.

MyoF-mAID Live replication assay

Parasites were induced +/- auxin for 4 h prior to labelling. Parasites were labelled with the membrane-permeable Halo dye Janelia Fluor® 646 (1:1000, Promega) for 1 h and washed three times prior to transfer onto HFF-covered live-cell dishes. Parasites were allowed to invade for 1–2 h, after which excess parasites were removed by three washes. Janelia Fluor® 549 (1:10000, Promega) was added to the media to record de novo generation of the protein of interest during live imaging. Live-cell dishes were then transferred to the Leica DMI8 at 37 °C and 5% CO₂. Laser power and exposure were adjusted to the lowest values allowing reliable imaging. Images were taken every 15–30 min for 12 h to follow the tagged protein during replication. For the auxin-washout assay, auxin was removed after 24 h of replication as described above (“Auxin chased experiment”), and live imaging was subsequently initiated. Time 0 in Figure 8F and Video S6 corresponds to the first acquired frame after auxin washout. Experiments were performed in three independent biological replicates and images were analysed in Fiji.

Imaging

Widefield microscopy

Unless stated otherwise, all images were acquired on a Leica-DMI8, objective 100x with the LasX software (v3.7.4). Fiji (v1.53c) was used to analyse the picture and all counts were made manually. LasX software (v.) from Leica was used to obtain parasite imaging data and All images and movies were processed using Fiji (ImageJ) software v Image Processing Software (Schindelin et al., 2012).

Software

Fiji (FIJI ImageJ v1.54f) was used to analyze the picture and all counts were made manually.

Data analysis

All data were plotted using Microsoft Excel.

Statistical analysis

Two-tailed unpaired Student’s t-tests were performed to evaluate statistical significance.

For Figures 2C, 2E, 2F, 3B, 3D, 4B, 4C, 5C (stage 1–2), and Supplementary Figures S1A, S1B, S2C–S2G, S4B–S4D, differences between stages 1–2, 2–4, and 4–8 were assessed.

For Figures 2B, 4C, and 5C (stage 2–8), variance could not be calculated because values were constant (0 or 100%), and these were therefore labeled as non-significant.

For Figures 7D, S8D, and S9B, differences between uninduced controls and MyoF knockdown (KD) induced samples were analyzed.

For Figure 8C, the distribution of phenotypes (normal, accumulated, and redistributed micronemes) was compared between control, auxin-induced (48 h), and auxin-washed (chased) conditions.

For Figure S5B, microneme distribution across positions 1–2, 2–3, 3–4, 4–5, 5–6, 6–7, and 7– 8 was analyzed; no statistically significant differences were observed, and all were labeled as non-significant.

Finally, for Figure S7F, the difference in fluorescence intensity between daughter and mother IMC signals was assessed.

Data are presented as mean ± SD. Statistical significance ns: not significant (p ≥ 0.05); * : p < 0.05; ** : p < 0.01; *** : p < 0.001.